---
title: "Hands-on Exercise 9b - Visualising Geospatial Point Data"
author: "Yang Yang"
format:
  html:
    theme: sketchy
    toc: true
    toc-depth: 3
    number-sections: false
    code-copy: true
    code-overflow: wrap
editor: visual
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

user_lib <- Sys.getenv("R_LIBS_USER")
if (nzchar(user_lib)) {
  dir.create(user_lib, recursive = TRUE, showWarnings = FALSE)
  .libPaths(c(user_lib, .libPaths()))
}

if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

pacman::p_load(
  sf,
  tmap,
  tidyverse,
  DT,
  knitr,
  scales
)

my_teal <- "#2A6F6F"
my_soft_teal <- "#8AA6A3"
my_dark <- "#1F1F1F"
my_light_teal <- "#D8E8E6"
my_grey <- "#F3F4F4"
my_warning <- "#B07C3E"
my_red <- "#9A4B4B"
my_blue <- "#4C78A8"

find_data_file <- function(file_name) {
  candidates <- c(
    file.path("data", "aspatial", file_name),
    file.path("Hands-on_Ex", "Hands-on_Ex09b", "data", "aspatial", file_name)
  )
  found <- candidates[file.exists(candidates)]

  if (length(found) == 0) {
    stop(
      "Missing required data file: ", file_name,
      ". Put it inside Hands-on_Ex/Hands-on_Ex09b/data/aspatial/ before rendering."
    )
  }

  found[1]
}

theme_yy <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", colour = my_dark),
    plot.subtitle = element_text(colour = my_teal),
    axis.title = element_text(colour = my_dark),
    axis.text = element_text(colour = my_dark),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## 23 Overview

This hands-on exercise follows [Chapter 22 of R4VA](https://r4va.netlify.app/chap22), *Visualising Geospatial Point Data*. The focus is on proportional symbol mapping, where point size represents a numerical value.

The example uses Singapore Pools outlet data. Each outlet has projected SVY21 coordinates and a count of Group 1 and Group 2 winnings.

### 23.1 Learning Outcome

By the end of this exercise, the following skills are practised:

- importing an aspatial CSV file into R;
- converting an aspatial data frame into a simple feature point data frame;
- assigning the correct coordinate reference system, EPSG:3414;
- drawing interactive point symbol maps with `tmap`;
- mapping point size by a numeric variable; and
- mapping point colour by a categorical variable.

## 24 Getting Started

Before starting, the required packages must be installed and loaded. The setup chunk above runs the executable package code below.

```{r}
#| label: package-code
#| eval: false

if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

pacman::p_load(sf, tmap, tidyverse, DT, knitr, scales)
```

```{r}
#| label: package-table

tribble(
  ~Package, ~Purpose,
  "sf", "Convert Singapore Pools outlet records into simple feature point data.",
  "tmap", "Create interactive point symbol and proportional symbol maps.",
  "tidyverse", "Import, wrangle, and summarise the data.",
  "DT", "Display interactive data tables.",
  "knitr", "Create compact summary tables.",
  "scales", "Format numeric labels."
) |>
  datatable(
    rownames = FALSE,
    options = list(pageLength = 10, dom = "tip")
  )
```

## 25 Geospatial Data Wrangling

### 25.1 The Data

The dataset used in this exercise is `SGPools_svy21.csv`. It contains 306 Singapore Pools outlets and branches.

The key fields are:

- `XCOORD` and `YCOORD`: projected SVY21 coordinates;
- `OUTLET TYPE`: whether the record is a branch or outlet; and
- `Gp1Gp2 Winnings`: number of Group 1 and Group 2 winnings.

### 25.2 Data Import and Preparation

The CSV file is imported as a tibble with `read_csv()`.

```{r}
#| label: import-sgpools

sgpools <- read_csv(
  find_data_file("SGPools_svy21.csv"),
  show_col_types = FALSE
)

sgpools |>
  slice_head(n = 15) |>
  datatable(
    rownames = FALSE,
    options = list(pageLength = 10, scrollX = TRUE, dom = "tip")
  )
```

The next checks confirm the imported structure and the main mapping variables.

```{r}
#| label: sgpools-summary

summary_table <- tibble(
  measure = c(
    "Number of records",
    "Number of columns",
    "Minimum winnings",
    "Median winnings",
    "Maximum winnings",
    "Outlet types"
  ),
  value = c(
    nrow(sgpools),
    ncol(sgpools),
    min(sgpools$`Gp1Gp2 Winnings`, na.rm = TRUE),
    median(sgpools$`Gp1Gp2 Winnings`, na.rm = TRUE),
    max(sgpools$`Gp1Gp2 Winnings`, na.rm = TRUE),
    paste(names(table(sgpools$`OUTLET TYPE`)), collapse = ", ")
  )
)

kable(summary_table)
```

```{r}
#| label: outlet-type-counts

sgpools |>
  count(`OUTLET TYPE`, name = "outlets") |>
  mutate(share = percent(outlets / sum(outlets))) |>
  datatable(
    rownames = FALSE,
    options = list(pageLength = 5, dom = "tip")
  )
```

### 25.3 Creating an sf Data Frame from an Aspatial Data Frame

The code below converts the tibble into a simple feature point data frame. The `coords` argument uses the `XCOORD` and `YCOORD` fields. The coordinate reference system is EPSG:3414, Singapore SVY21 / Singapore TM.

```{r}
#| label: create-sf

sgpools_sf <- st_as_sf(
  sgpools,
  coords = c("XCOORD", "YCOORD"),
  crs = 3414
)

sgpools_sf
```

```{r}
#| label: sf-metadata

tibble(
  measure = c(
    "Geometry type",
    "Coordinate reference system",
    "Bounding box xmin",
    "Bounding box ymin",
    "Bounding box xmax",
    "Bounding box ymax"
  ),
  value = c(
    paste(unique(st_geometry_type(sgpools_sf)), collapse = ", "),
    st_crs(sgpools_sf)$input,
    round(st_bbox(sgpools_sf)[["xmin"]], 3),
    round(st_bbox(sgpools_sf)[["ymin"]], 3),
    round(st_bbox(sgpools_sf)[["xmax"]], 3),
    round(st_bbox(sgpools_sf)[["ymax"]], 3)
  )
) |>
  kable()
```

## 26 Drawing Proportional Symbol Map

Interactive maps are drawn by switching `tmap` into view mode.

```{r}
#| label: tmap-view-mode

tmap_mode("view")
```

### 26.1 It All Started with an Interactive Point Symbol Map

The first map displays every Singapore Pools outlet as a fixed-size red point.

::: panel-tabset

## View

```{r}
#| label: fixed-symbol-view
#| echo: false

tm_shape(sgpools_sf) +
  tm_bubbles(
    fill = "red",
    size = 1,
    col = "black",
    lwd = 1,
    popup.vars = c(
      "Name" = "NAME",
      "Address" = "ADDRESS",
      "Outlet type" = "OUTLET TYPE",
      "Winnings" = "Gp1Gp2 Winnings"
    )
  )
```

## Code

```{r}
#| label: fixed-symbol-code
#| eval: false

tm_shape(sgpools_sf) +
  tm_bubbles(
    fill = "red",
    size = 1,
    col = "black",
    lwd = 1
  )
```

:::

### 26.2 Let's Make It Proportional

To draw a proportional symbol map, `Gp1Gp2 Winnings` is assigned to the size visual attribute.

::: panel-tabset

## View

```{r}
#| label: proportional-symbol-view
#| echo: false

tm_shape(sgpools_sf) +
  tm_bubbles(
    fill = my_red,
    size = "Gp1Gp2 Winnings",
    col = "black",
    lwd = 1,
    popup.vars = c(
      "Name" = "NAME",
      "Address" = "ADDRESS",
      "Outlet type" = "OUTLET TYPE",
      "Winnings" = "Gp1Gp2 Winnings"
    )
  )
```

## Code

```{r}
#| label: proportional-symbol-code
#| eval: false

tm_shape(sgpools_sf) +
  tm_bubbles(
    fill = "red",
    size = "Gp1Gp2 Winnings",
    col = "black",
    lwd = 1
  )
```

:::

### 26.3 Let's Give It a Different Colour

The map below uses point size for winnings and point colour for outlet type.

::: panel-tabset

## View

```{r}
#| label: colour-symbol-view
#| echo: false

tm_shape(sgpools_sf) +
  tm_bubbles(
    fill = "OUTLET TYPE",
    size = "Gp1Gp2 Winnings",
    col = "black",
    lwd = 1,
    popup.vars = c(
      "Name" = "NAME",
      "Address" = "ADDRESS",
      "Outlet type" = "OUTLET TYPE",
      "Winnings" = "Gp1Gp2 Winnings"
    )
  )
```

## Code

```{r}
#| label: colour-symbol-code
#| eval: false

tm_shape(sgpools_sf) +
  tm_bubbles(
    fill = "OUTLET TYPE",
    size = "Gp1Gp2 Winnings",
    col = "black",
    lwd = 1
  )
```

:::

### 26.4 Faceted Interactive Maps

In view mode, `tm_facets()` can create synchronized maps. The `sync = TRUE` argument keeps zoom and pan settings linked across facets.

::: panel-tabset

## View

```{r}
#| label: faceted-symbol-view
#| echo: false

tm_shape(sgpools_sf) +
  tm_bubbles(
    fill = "OUTLET TYPE",
    size = "Gp1Gp2 Winnings",
    col = "black",
    lwd = 1,
    popup.vars = c(
      "Name" = "NAME",
      "Address" = "ADDRESS",
      "Outlet type" = "OUTLET TYPE",
      "Winnings" = "Gp1Gp2 Winnings"
    )
  ) +
  tm_facets(
    by = "OUTLET TYPE",
    nrow = 1,
    sync = TRUE
  )
```

## Code

```{r}
#| label: faceted-symbol-code
#| eval: false

tm_shape(sgpools_sf) +
  tm_bubbles(
    fill = "OUTLET TYPE",
    size = "Gp1Gp2 Winnings",
    col = "black",
    lwd = 1
  ) +
  tm_facets(
    by = "OUTLET TYPE",
    nrow = 1,
    sync = TRUE
  )
```

:::

Before ending the session, switch `tmap` back to plot mode.

```{r}
#| label: tmap-plot-mode

tmap_mode("plot")
```

## Additional Exploration

The bar chart below gives a non-map view of the same key variable. It shows that outlets generally have more records than branches, while both outlet types include some high-winning locations.

::: panel-tabset

## View

```{r}
#| label: winnings-by-type-view
#| echo: false

ggplot(sgpools, aes(`OUTLET TYPE`, `Gp1Gp2 Winnings`, fill = `OUTLET TYPE`)) +
  geom_boxplot(alpha = 0.75, width = 0.55, outlier.shape = 21) +
  scale_fill_manual(values = c("Branch" = my_teal, "Outlet" = my_warning)) +
  labs(
    title = "Group 1 and Group 2 Winnings by Outlet Type",
    x = NULL,
    y = "Gp1Gp2 Winnings"
  ) +
  theme_yy +
  theme(legend.position = "none")
```

## Code

```{r}
#| label: winnings-by-type-code
#| eval: false

ggplot(sgpools, aes(`OUTLET TYPE`, `Gp1Gp2 Winnings`, fill = `OUTLET TYPE`)) +
  geom_boxplot(alpha = 0.75, width = 0.55, outlier.shape = 21) +
  labs(
    title = "Group 1 and Group 2 Winnings by Outlet Type",
    x = NULL,
    y = "Gp1Gp2 Winnings"
  ) +
  theme_minimal()
```

:::

## Key Takeaways

::: {.callout-tip}
Point data becomes mappable when the coordinate columns are converted into geometry and assigned the correct CRS. In this exercise, the SVY21 coordinates are converted into an `sf` point layer with EPSG:3414 before being mapped with `tmap`.
:::

In this exercise:

- `read_csv()` imports the Singapore Pools outlet records;
- `st_as_sf()` converts the table into point features;
- `tmap_mode("view")` enables interactive maps;
- `tm_bubbles()` creates fixed-size and proportional point symbols; and
- `tm_facets(sync = TRUE)` creates synchronized interactive maps by outlet type.

## References

- [R4VA Chapter 22: Visualising Geospatial Point Data](https://r4va.netlify.app/chap22)
- [sf: Simple Features for R](https://cran.r-project.org/package=sf)
- [tmap: Thematic Maps in R](https://cran.r-project.org/package=tmap)

::: vaa-exercise-footer
### Portfolio note

**Key takeaway.** Point maps reveal location patterns that area summaries may hide.

**Practical use.** Use it when the distribution, clustering, or proximity of individual locations matters.

::: vaa-footer-nav
[Back to Gallery](../../hands-on-gallery.html) [Previous: Exercise 9a](../Hands-on_Ex09a/Hands-on_Ex09a.html) [Next: Exercise 9c](../Hands-on_Ex09c/Hands-on_Ex09c.html)
:::
:::

