---
title: "Hands-on Exercise 1:A Layered Grammar of Graphics - gdgplot2 methods"
author: "Yang Yang"
---

# 1.1 Overview

In this chapter, you will learn the fundamental principles and key components of ggplot2. You will also gain practical experience in using these components to create statistical graphics based on the Layered Grammar of Graphics. By the end of this chapter, you should be able to apply the main graphical elements in ggplot2 to produce statistical charts that are both visually clear and functionally effective.

# 1.2 Getting Started

## 1.2.1 Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If the are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## 1.2.2 Importing the data

```{r}
exam_data <- read_csv("Exam_data.csv")

```

## 1.3 Introducing ggplot

## 1.3.1 Plotting a simple bar chart

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) + 
  geom_bar()
```
