---
title: "Hands-on Exercise 5b - Visual Correlation Analysis"
author: "Yang Yang"
format:
  html:
    theme: sketchy
    toc: true
    toc-depth: 3
    number-sections: false
    code-copy: true
    code-overflow: wrap
editor: visual
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

if (!require(pacman)) {
  install.packages("pacman")
}

pacman::p_load(
  tidyverse,
  corrplot,
  ggstatsplot,
  DT,
  janitor,
  scales,
  plotly,
  knitr
)

my_teal <- "#2A6F6F"
my_soft_teal <- "#8AA6A3"
my_dark <- "#1F1F1F"
my_light_teal <- "#D8E8E6"
my_grey <- "#F3F4F4"
my_warning <- "#B07C3E"
my_red <- "#9A4B4B"

corr_cols <- colorRampPalette(
  c(my_red, my_light_teal, my_teal)
)(200)

theme_yy <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", colour = my_dark),
    plot.subtitle = element_text(colour = my_teal),
    axis.title = element_text(colour = my_dark),
    axis.text = element_text(colour = my_dark),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## 5.1 Overview

Correlation coefficient is used to measure the direction and strength of the linear relationship between two variables. A value close to 1 means a strong positive relationship, a value close to -1 means a strong negative relationship, and a value close to 0 means weak or no linear relationship.

When the dataset has many numerical variables, pairwise correlations can be displayed as a correlation matrix, scatterplot matrix, or corrgram. These methods help us identify which variables move together, which variables move in opposite directions, and which variables may create multicollinearity problems in later modelling.

In this hands-on exercise, we use the wine quality dataset. The dataset contains chemical measurements of wine, such as acidity, residual sugar, chlorides, sulphates, alcohol, and quality score.

By the end of this exercise, we will be able to:

- import and prepare the wine quality dataset,
- build scatterplot matrices using `pairs()`,
- add correlation coefficients to a scatterplot matrix,
- visualise correlation matrices using `ggcorrmat()`,
- compare red and white wine correlation structures,
- create corrgrams using `corrplot`,
- combine corrgrams with significance testing,
- reorder correlation matrices to reveal hidden structures,
- add small extensions to interpret wine quality drivers.

::: {.callout-important}
## Important

Correlation does not prove causation.  
It only describes the linear association between two variables.
:::

## 5.2 Installing and Launching R Packages

The main packages used in this exercise are `corrplot`, `ggstatsplot`, `tidyverse`, and `plotly`.

::: {.panel-tabset}

## View

```{r}
#| echo: false

package_table <- tibble(
  Package = c("tidyverse", "corrplot", "ggstatsplot", "plotly", "DT", "janitor"),
  Purpose = c(
    "Data import, wrangling, and ggplot2 workflow",
    "Corrgram and correlation matrix visualisation",
    "Professional statistical correlation matrix",
    "Interactive correlation heatmap",
    "Interactive HTML tables",
    "Cleaning column names"
  )
)

DT::datatable(
  package_table,
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t", pageLength = 6),
  width = "100%"
)
```

## Code

```{r}
#| eval: false

pacman::p_load(
  tidyverse,
  corrplot,
  ggstatsplot,
  plotly,
  DT,
  janitor
)
```

:::

## 5.3 Importing and Preparing the Data Set

The dataset used in this exercise is `wine_quality.csv`.

### 5.3.1 Importing data

::: {.panel-tabset}

## View

```{r}
#| echo: false

csv_file <- list.files(pattern = "wine_quality.*\\.csv$")[1]

wine_raw <- read_csv(csv_file, show_col_types = FALSE)

glimpse(wine_raw)
```

## Code

```{r}
csv_file <- list.files(pattern = "wine_quality.*\\.csv$")[1]

wine_raw <- read_csv(csv_file, show_col_types = FALSE)

glimpse(wine_raw)
```

:::

Using `list.files()` makes the code safer because it can still read filenames such as `wine_quality(2).csv`.

### 5.3.2 Cleaning variable names

::: {.panel-tabset}

## View

```{r}
#| echo: false

wine <- wine_raw %>%
  janitor::clean_names()

names(wine)
```

## Code

```{r}
wine <- wine_raw %>%
  janitor::clean_names()

names(wine)
```

:::

After cleaning, variables such as `fixed acidity`, `volatile acidity`, and `free sulfur dioxide` become easier to use in R.

### 5.3.3 Selecting numerical variables

::: {.panel-tabset}

## View

```{r}
#| echo: false

wine_numeric_vars <- c(
  "fixed_acidity",
  "volatile_acidity",
  "citric_acid",
  "residual_sugar",
  "chlorides",
  "free_sulfur_dioxide",
  "total_sulfur_dioxide",
  "density",
  "p_h",
  "sulphates",
  "alcohol"
)

wine_numeric <- wine %>%
  select(all_of(wine_numeric_vars)) %>%
  mutate(across(everything(), as.numeric)) %>%
  drop_na()

wine_for_analysis <- wine %>%
  mutate(across(all_of(wine_numeric_vars), as.numeric)) %>%
  drop_na(all_of(wine_numeric_vars))

DT::datatable(
  head(wine_for_analysis, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
wine_numeric_vars <- c(
  "fixed_acidity",
  "volatile_acidity",
  "citric_acid",
  "residual_sugar",
  "chlorides",
  "free_sulfur_dioxide",
  "total_sulfur_dioxide",
  "density",
  "p_h",
  "sulphates",
  "alcohol"
)

wine_numeric <- wine %>%
  select(all_of(wine_numeric_vars)) %>%
  mutate(across(everything(), as.numeric)) %>%
  drop_na()

wine_for_analysis <- wine %>%
  mutate(across(all_of(wine_numeric_vars), as.numeric)) %>%
  drop_na(all_of(wine_numeric_vars))
```

:::

Only the chemical measurement variables are selected for the correlation analysis.

## 5.4 Building Correlation Matrix: pairs() Method

The `pairs()` function creates a scatterplot matrix. It is useful because it shows the actual pairwise scatterplots, not only the correlation coefficients.

### 5.4.1 Building a basic correlation matrix

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 9

pairs(
  wine_numeric,
  col = adjustcolor(my_teal, alpha.f = 0.25),
  pch = 16,
  main = "Scatterplot matrix of wine chemical variables"
)
```

## Code

```{r}
pairs(
  wine_numeric,
  col = adjustcolor(my_teal, alpha.f = 0.25),
  pch = 16,
  main = "Scatterplot matrix of wine chemical variables"
)
```

:::

The scatterplot matrix gives a first view of pairwise relationships. However, because the dataset has many observations, the plot can become visually crowded.

### 5.4.2 Drawing the lower corner

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 9

pairs(
  wine_numeric,
  upper.panel = NULL,
  col = adjustcolor(my_teal, alpha.f = 0.25),
  pch = 16,
  main = "Lower-corner scatterplot matrix"
)
```

## Code

```{r}
pairs(
  wine_numeric,
  upper.panel = NULL,
  col = adjustcolor(my_teal, alpha.f = 0.25),
  pch = 16,
  main = "Lower-corner scatterplot matrix"
)
```

:::

A correlation matrix is symmetric, so showing only the lower half reduces repeated information.

### 5.4.3 Drawing the upper corner

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 9

pairs(
  wine_numeric,
  lower.panel = NULL,
  col = adjustcolor(my_teal, alpha.f = 0.25),
  pch = 16,
  main = "Upper-corner scatterplot matrix"
)
```

## Code

```{r}
pairs(
  wine_numeric,
  lower.panel = NULL,
  col = adjustcolor(my_teal, alpha.f = 0.25),
  pch = 16,
  main = "Upper-corner scatterplot matrix"
)
```

:::

This version shows the same relationships, but only in the upper triangle.

### 5.4.4 Including correlation coefficients

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 9

panel_cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
  usr <- par("usr")
  on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- cor(x, y, use = "complete.obs")
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste(prefix, txt, sep = "")
  if (missing(cex.cor)) {
    cex.cor <- 0.8 / strwidth(txt)
  }
  text(
    0.5,
    0.5,
    txt,
    cex = cex.cor * (1 + abs(r)) / 2,
    col = ifelse(r >= 0, my_teal, my_red)
  )
}

pairs(
  wine_numeric,
  upper.panel = panel_cor,
  col = adjustcolor(my_teal, alpha.f = 0.25),
  pch = 16,
  main = "Scatterplot matrix with correlation coefficients"
)
```

## Code

```{r}
panel_cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
  usr <- par("usr")
  on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- cor(x, y, use = "complete.obs")
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste(prefix, txt, sep = "")
  if (missing(cex.cor)) {
    cex.cor <- 0.8 / strwidth(txt)
  }
  text(
    0.5,
    0.5,
    txt,
    cex = cex.cor * (1 + abs(r)) / 2,
    col = ifelse(r >= 0, my_teal, my_red)
  )
}

pairs(
  wine_numeric,
  upper.panel = panel_cor,
  col = adjustcolor(my_teal, alpha.f = 0.25),
  pch = 16,
  main = "Scatterplot matrix with correlation coefficients"
)
```

:::

This version combines scatterplots and numerical correlation coefficients in one matrix.

::: {.callout-tip}
## What can we learn?

The scatterplot matrix shows the shape of the relationship, while the coefficient shows the strength of the linear relationship.  
Both are needed because a correlation coefficient alone can hide non-linear patterns.
:::

## 5.5 Visualising Correlation Matrix: ggcorrmat()

Scatterplot matrices can be cluttered when there are many observations. A corrgram gives a cleaner summary of pairwise correlations.

### 5.5.1 Basic ggcorrmat plot

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6

ggstatsplot::ggcorrmat(
  data = wine_for_analysis,
  cor.vars = all_of(wine_numeric_vars),
  ggcorrplot.args = list(
    outline.color = my_dark,
    hc.order = TRUE,
    tl.cex = 8
  ),
  title = "Correlogram for wine quality dataset",
  subtitle = "Correlation matrix of wine chemical measurements"
)
```

## Code

```{r}
ggstatsplot::ggcorrmat(
  data = wine_for_analysis,
  cor.vars = all_of(wine_numeric_vars),
  ggcorrplot.args = list(
    outline.color = my_dark,
    hc.order = TRUE,
    tl.cex = 8
  ),
  title = "Correlogram for wine quality dataset",
  subtitle = "Correlation matrix of wine chemical measurements"
)
```

:::

The `ggcorrmat()` output is more compact than the full scatterplot matrix and includes a statistical summary.

### 5.5.2 Adjusting text size and theme

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6

ggstatsplot::ggcorrmat(
  data = wine_for_analysis,
  cor.vars = all_of(wine_numeric_vars),
  ggcorrplot.args = list(
    outline.color = my_dark,
    hc.order = TRUE,
    tl.cex = 8
  ),
  ggplot.component = list(
    theme(
      text = element_text(size = 9),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8)
    )
  ),
  title = "Wine variable correlation matrix",
  subtitle = "Hierarchical ordering helps place similar variables near each other"
)
```

## Code

```{r}
ggstatsplot::ggcorrmat(
  data = wine_for_analysis,
  cor.vars = all_of(wine_numeric_vars),
  ggcorrplot.args = list(
    outline.color = my_dark,
    hc.order = TRUE,
    tl.cex = 8
  ),
  ggplot.component = list(
    theme(
      text = element_text(size = 9),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8)
    )
  ),
  title = "Wine variable correlation matrix",
  subtitle = "Hierarchical ordering helps place similar variables near each other"
)
```

:::

This version improves readability by adjusting label size.

## 5.6 Building Multiple Plots

The wine dataset contains both red and white wine. It is useful to compare whether the correlation structure is the same across wine type.

### 5.6.1 Checking wine type

::: {.panel-tabset}

## View

```{r}
#| echo: false

wine_type_summary <- wine_for_analysis %>%
  count(type, name = "records")

DT::datatable(
  wine_type_summary,
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t"),
  width = "100%"
)
```

## Code

```{r}
wine_type_summary <- wine_for_analysis %>%
  count(type, name = "records")
```

:::

The `type` variable allows the correlation matrix to be compared between red wine and white wine.

### 5.6.2 Grouped correlation matrix by wine type

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 6

ggstatsplot::grouped_ggcorrmat(
  data = wine_for_analysis,
  cor.vars = all_of(wine_numeric_vars),
  grouping.var = type,
  type = "robust",
  p.adjust.method = "holm",
  plotgrid.args = list(ncol = 2),
  ggcorrplot.args = list(
    outline.color = my_dark,
    hc.order = TRUE,
    tl.cex = 7
  ),
  annotation.args = list(
    title = "Grouped correlogram for red and white wine",
    subtitle = "Robust correlation is used to reduce sensitivity to outliers",
    caption = "Dataset: Wine quality data"
  )
)
```

## Code

```{r}
ggstatsplot::grouped_ggcorrmat(
  data = wine_for_analysis,
  cor.vars = all_of(wine_numeric_vars),
  grouping.var = type,
  type = "robust",
  p.adjust.method = "holm",
  plotgrid.args = list(ncol = 2),
  ggcorrplot.args = list(
    outline.color = my_dark,
    hc.order = TRUE,
    tl.cex = 7
  ),
  annotation.args = list(
    title = "Grouped correlogram for red and white wine",
    subtitle = "Robust correlation is used to reduce sensitivity to outliers",
    caption = "Dataset: Wine quality data"
  )
)
```

:::

The grouped correlogram makes it easier to see whether the same chemical relationships hold for both wine types.

## 5.7 Visualising Correlation Matrix Using corrplot Package

The `corrplot` package provides several visual styles for corrgrams.

### 5.7.1 Getting started with corrplot

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7

wine_cor <- cor(wine_numeric, use = "pairwise.complete.obs")

corrplot::corrplot(
  wine_cor,
  lower.col = corr_cols,
  upper.col = corr_cols,
  tl.col = my_dark,
  tl.srt = 45
)
```

## Code

```{r}
wine_cor <- cor(wine_numeric, use = "pairwise.complete.obs")

corrplot::corrplot(
  wine_cor,
  lower.col = corr_cols,
  upper.col = corr_cols,
  tl.col = my_dark,
  tl.srt = 45
)
```

:::

The default corrplot uses circles to represent correlation strength and direction.

### 5.7.2 Working with visual geometrics

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7

corrplot::corrplot(
  wine_cor,
  method = "ellipse",
  lower.col = corr_cols,
  upper.col = corr_cols,
  tl.col = my_dark,
  tl.srt = 45
)
```

## Code

```{r}
corrplot::corrplot(
  wine_cor,
  method = "ellipse",
  lower.col = corr_cols,
  upper.col = corr_cols,
  tl.col = my_dark,
  tl.srt = 45
)
```

:::

The ellipse method makes the sign and strength of each correlation visually clear.

### 5.7.3 Working with layout

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7

corrplot::corrplot(
  wine_cor,
  method = "ellipse",
  type = "lower",
  lower.col = corr_cols,
  upper.col = corr_cols,
  diag = FALSE,
  tl.col = my_dark,
  tl.srt = 45
)
```

## Code

```{r}
corrplot::corrplot(
  wine_cor,
  method = "ellipse",
  type = "lower",
  lower.col = corr_cols,
  upper.col = corr_cols,
  diag = FALSE,
  tl.col = my_dark,
  tl.srt = 45
)
```

:::

The lower-triangle layout avoids repeating the same correlation pairs.

### 5.7.4 Working with mixed layout

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7

corrplot::corrplot.mixed(
  wine_cor,
  lower = "ellipse",
  upper = "number",
  lower.col = corr_cols,
  upper.col = corr_cols,
  tl.pos = "lt",
  diag = "l",
  tl.col = my_dark,
  tl.srt = 45,
  number.cex = 0.65
)
```

## Code

```{r}
corrplot::corrplot.mixed(
  wine_cor,
  lower = "ellipse",
  upper = "number",
  lower.col = corr_cols,
  upper.col = corr_cols,
  tl.pos = "lt",
  diag = "l",
  tl.col = my_dark,
  tl.srt = 45,
  number.cex = 0.65
)
```

:::

The mixed layout is useful because the lower triangle gives visual pattern recognition, while the upper triangle provides exact numerical values.

### 5.7.5 Combining corrgram with significance test

::: {.panel-tabset}

## View

```{r}
#| echo: false

cor_mtest <- function(mat, conf.level = 0.95) {
  mat <- as.matrix(mat)
  n <- ncol(mat)
  p_mat <- low_ci_mat <- upp_ci_mat <- matrix(NA, n, n)
  diag(p_mat) <- 0
  diag(low_ci_mat) <- diag(upp_ci_mat) <- 1
  
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      test_result <- cor.test(mat[, i], mat[, j], conf.level = conf.level)
      p_mat[i, j] <- p_mat[j, i] <- test_result$p.value
      low_ci_mat[i, j] <- low_ci_mat[j, i] <- test_result$conf.int[1]
      upp_ci_mat[i, j] <- upp_ci_mat[j, i] <- test_result$conf.int[2]
    }
  }
  
  colnames(p_mat) <- rownames(p_mat) <- colnames(mat)
  colnames(low_ci_mat) <- rownames(low_ci_mat) <- colnames(mat)
  colnames(upp_ci_mat) <- rownames(upp_ci_mat) <- colnames(mat)
  
  list(p = p_mat, low_ci = low_ci_mat, upp_ci = upp_ci_mat)
}

wine_sig <- cor_mtest(wine_numeric, conf.level = 0.95)

DT::datatable(
  round(wine_sig$p, 4),
  class = "compact stripe hover",
  options = list(pageLength = 11, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
cor_mtest <- function(mat, conf.level = 0.95) {
  mat <- as.matrix(mat)
  n <- ncol(mat)
  p_mat <- low_ci_mat <- upp_ci_mat <- matrix(NA, n, n)
  diag(p_mat) <- 0
  diag(low_ci_mat) <- diag(upp_ci_mat) <- 1
  
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      test_result <- cor.test(mat[, i], mat[, j], conf.level = conf.level)
      p_mat[i, j] <- p_mat[j, i] <- test_result$p.value
      low_ci_mat[i, j] <- low_ci_mat[j, i] <- test_result$conf.int[1]
      upp_ci_mat[i, j] <- upp_ci_mat[j, i] <- test_result$conf.int[2]
    }
  }
  
  colnames(p_mat) <- rownames(p_mat) <- colnames(mat)
  colnames(low_ci_mat) <- rownames(low_ci_mat) <- colnames(mat)
  colnames(upp_ci_mat) <- rownames(upp_ci_mat) <- colnames(mat)
  
  list(p = p_mat, low_ci = low_ci_mat, upp_ci = upp_ci_mat)
}

wine_sig <- cor_mtest(wine_numeric, conf.level = 0.95)
```

:::

This step calculates p-values for each pairwise correlation.

### 5.7.6 Corrgram with insignificant correlations marked

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7

corrplot::corrplot(
  wine_cor,
  method = "number",
  type = "lower",
  lower.col = corr_cols,
  upper.col = corr_cols,
  diag = FALSE,
  tl.col = my_dark,
  tl.srt = 45,
  p.mat = wine_sig$p,
  sig.level = 0.05,
  insig = "pch",
  pch = 4,
  pch.col = my_dark,
  number.cex = 0.65
)
```

## Code

```{r}
corrplot::corrplot(
  wine_cor,
  method = "number",
  type = "lower",
  lower.col = corr_cols,
  upper.col = corr_cols,
  diag = FALSE,
  tl.col = my_dark,
  tl.srt = 45,
  p.mat = wine_sig$p,
  sig.level = 0.05,
  insig = "pch",
  pch = 4,
  pch.col = my_dark,
  number.cex = 0.65
)
```

:::

This plot marks correlations that are not statistically significant at the 5% level.

### 5.7.7 Reorder a corrgram

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7

corrplot::corrplot.mixed(
  wine_cor,
  lower = "ellipse",
  upper = "number",
  lower.col = corr_cols,
  upper.col = corr_cols,
  tl.pos = "lt",
  diag = "l",
  order = "AOE",
  tl.col = my_dark,
  tl.srt = 45,
  number.cex = 0.65
)
```

## Code

```{r}
corrplot::corrplot.mixed(
  wine_cor,
  lower = "ellipse",
  upper = "number",
  lower.col = corr_cols,
  upper.col = corr_cols,
  tl.pos = "lt",
  diag = "l",
  order = "AOE",
  tl.col = my_dark,
  tl.srt = 45,
  number.cex = 0.65
)
```

:::

Reordering helps place similar variables closer together, making hidden structure easier to detect.

### 5.7.8 Reordering a correlation matrix using hclust

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7

corrplot::corrplot(
  wine_cor,
  method = "ellipse",
  lower.col = corr_cols,
  upper.col = corr_cols,
  tl.pos = "lt",
  tl.col = my_dark,
  tl.srt = 45,
  order = "hclust",
  hclust.method = "ward.D",
  addrect = 3
)
```

## Code

```{r}
corrplot::corrplot(
  wine_cor,
  method = "ellipse",
  lower.col = corr_cols,
  upper.col = corr_cols,
  tl.pos = "lt",
  tl.col = my_dark,
  tl.srt = 45,
  order = "hclust",
  hclust.method = "ward.D",
  addrect = 3
)
```

:::

The rectangles show clusters of variables with similar correlation patterns.

## 5.8 Interactive Correlation Matrix

The PDF mentions interactive correlation analysis. Here, we create an interactive heatmap using `plotly`.

### 5.8.1 Preparing data for interactive heatmap

::: {.panel-tabset}

## View

```{r}
#| echo: false

wine_cor_plotly <- round(wine_cor, 3)

DT::datatable(
  wine_cor_plotly,
  class = "compact stripe hover",
  options = list(pageLength = 11, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
wine_cor_plotly <- round(wine_cor, 3)
```

:::

### 5.8.2 Plotting interactive heatmap

::: {.panel-tabset}

## View

```{r}
#| echo: false

plot_ly(
  x = colnames(wine_cor_plotly),
  y = rownames(wine_cor_plotly),
  z = wine_cor_plotly,
  type = "heatmap",
  colorscale = list(
    list(0, my_red),
    list(0.5, my_light_teal),
    list(1, my_teal)
  ),
  zmin = -1,
  zmax = 1,
  hovertemplate = paste(
    "Variable X: %{x}",
    "<br>Variable Y: %{y}",
    "<br>Correlation: %{z}",
    "<extra></extra>"
  )
) %>%
  layout(
    title = "Interactive correlation heatmap of wine variables",
    xaxis = list(title = "", tickangle = -45),
    yaxis = list(title = "")
  )
```

## Code

```{r}
plot_ly(
  x = colnames(wine_cor_plotly),
  y = rownames(wine_cor_plotly),
  z = wine_cor_plotly,
  type = "heatmap",
  colorscale = list(
    list(0, my_red),
    list(0.5, my_light_teal),
    list(1, my_teal)
  ),
  zmin = -1,
  zmax = 1,
  hovertemplate = paste(
    "Variable X: %{x}",
    "<br>Variable Y: %{y}",
    "<br>Correlation: %{z}",
    "<extra></extra>"
  )
) %>%
  layout(
    title = "Interactive correlation heatmap of wine variables",
    xaxis = list(title = "", tickangle = -45),
    yaxis = list(title = "")
  )
```

:::

The interactive version allows users to hover over each cell and read the exact correlation coefficient.

## 5.9 Extra Analysis: Which Variables Relate Most to Wine Quality?

This extension is directly useful because the dataset is not only about chemical variables. It also contains a `quality` score. We can check which variables are most associated with wine quality.

### 5.9.1 Computing correlation with quality

::: {.panel-tabset}

## View

```{r}
#| echo: false

quality_cor <- wine_for_analysis %>%
  mutate(quality = as.numeric(quality)) %>%
  select(quality, all_of(wine_numeric_vars)) %>%
  summarise(
    across(
      all_of(wine_numeric_vars),
      ~ cor(.x, quality, use = "pairwise.complete.obs")
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Correlation_with_quality"
  ) %>%
  arrange(desc(abs(Correlation_with_quality)))

DT::datatable(
  quality_cor %>%
    mutate(Correlation_with_quality = round(Correlation_with_quality, 3)),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t", scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
quality_cor <- wine_for_analysis %>%
  mutate(quality = as.numeric(quality)) %>%
  select(quality, all_of(wine_numeric_vars)) %>%
  summarise(
    across(
      all_of(wine_numeric_vars),
      ~ cor(.x, quality, use = "pairwise.complete.obs")
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Correlation_with_quality"
  ) %>%
  arrange(desc(abs(Correlation_with_quality)))
```

:::

### 5.9.2 Visualising correlation with quality

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5

ggplot(
  quality_cor,
  aes(
    x = fct_reorder(Variable, Correlation_with_quality),
    y = Correlation_with_quality,
    fill = Correlation_with_quality
  )
) +
  geom_col(
    colour = my_dark,
    linewidth = 0.3
  ) +
  coord_flip() +
  scale_fill_gradient2(
    low = my_red,
    mid = my_light_teal,
    high = my_teal,
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation between chemical variables and wine quality",
    subtitle = "Alcohol usually shows one of the clearer positive relationships with quality",
    x = NULL,
    y = "Correlation with quality",
    fill = "Correlation"
  ) +
  theme_yy
```

## Code

```{r}
ggplot(
  quality_cor,
  aes(
    x = fct_reorder(Variable, Correlation_with_quality),
    y = Correlation_with_quality,
    fill = Correlation_with_quality
  )
) +
  geom_col(
    colour = my_dark,
    linewidth = 0.3
  ) +
  coord_flip() +
  scale_fill_gradient2(
    low = my_red,
    mid = my_light_teal,
    high = my_teal,
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation between chemical variables and wine quality",
    subtitle = "Alcohol usually shows one of the clearer positive relationships with quality",
    x = NULL,
    y = "Correlation with quality",
    fill = "Correlation"
  ) +
  theme_yy
```

:::

::: {.callout-tip}
## Insight

This is more useful than only showing a full correlation matrix.  
The full matrix tells us how variables relate to each other, while this chart focuses on the business-style question: which chemical variables are most linked with wine quality?
:::

## 5.10 Extra Analysis: Red Wine vs White Wine Quality Drivers

The relationship between chemical variables and quality may differ by wine type. This extension compares red and white wine separately.

### 5.10.1 Computing quality correlation by wine type

::: {.panel-tabset}

## View

```{r}
#| echo: false

quality_cor_by_type <- wine_for_analysis %>%
  mutate(quality = as.numeric(quality)) %>%
  group_by(type) %>%
  summarise(
    across(
      all_of(wine_numeric_vars),
      ~ cor(.x, quality, use = "pairwise.complete.obs")
    ),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = all_of(wine_numeric_vars),
    names_to = "Variable",
    values_to = "Correlation_with_quality"
  )

DT::datatable(
  quality_cor_by_type %>%
    mutate(Correlation_with_quality = round(Correlation_with_quality, 3)),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 22, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
quality_cor_by_type <- wine_for_analysis %>%
  mutate(quality = as.numeric(quality)) %>%
  group_by(type) %>%
  summarise(
    across(
      all_of(wine_numeric_vars),
      ~ cor(.x, quality, use = "pairwise.complete.obs")
    ),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = all_of(wine_numeric_vars),
    names_to = "Variable",
    values_to = "Correlation_with_quality"
  )
```

:::

### 5.10.2 Plotting quality drivers by wine type

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5

ggplot(
  quality_cor_by_type,
  aes(
    x = fct_reorder(Variable, abs(Correlation_with_quality)),
    y = Correlation_with_quality,
    fill = type
  )
) +
  geom_col(
    position = "dodge",
    colour = my_dark,
    linewidth = 0.25
  ) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "red" = my_red,
      "white" = my_teal
    )
  ) +
  labs(
    title = "Quality correlation differs by wine type",
    subtitle = "The same chemical variable may matter differently for red and white wine",
    x = NULL,
    y = "Correlation with quality",
    fill = "Wine type"
  ) +
  theme_yy
```

## Code

```{r}
ggplot(
  quality_cor_by_type,
  aes(
    x = fct_reorder(Variable, abs(Correlation_with_quality)),
    y = Correlation_with_quality,
    fill = type
  )
) +
  geom_col(
    position = "dodge",
    colour = my_dark,
    linewidth = 0.25
  ) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "red" = my_red,
      "white" = my_teal
    )
  ) +
  labs(
    title = "Quality correlation differs by wine type",
    subtitle = "The same chemical variable may matter differently for red and white wine",
    x = NULL,
    y = "Correlation with quality",
    fill = "Wine type"
  ) +
  theme_yy
```

:::

This extension adds interpretation rather than just more charts. It links the correlation analysis to a practical question: which variables are more relevant when comparing wine quality within each wine type?

## 5.11 Extra Analysis: Strongest Correlation Pairs

The final extension extracts the strongest absolute correlations from the matrix.

### 5.11.1 Preparing strongest correlation pairs

::: {.panel-tabset}

## View

```{r}
#| echo: false

strong_pairs <- as.data.frame(as.table(wine_cor)) %>%
  rename(
    Variable_1 = Var1,
    Variable_2 = Var2,
    Correlation = Freq
  ) %>%
  filter(as.character(Variable_1) < as.character(Variable_2)) %>%
  mutate(abs_correlation = abs(Correlation)) %>%
  arrange(desc(abs_correlation)) %>%
  slice_head(n = 10)

DT::datatable(
  strong_pairs %>%
    mutate(
      Correlation = round(Correlation, 3),
      abs_correlation = round(abs_correlation, 3)
    ),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t", scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
strong_pairs <- as.data.frame(as.table(wine_cor)) %>%
  rename(
    Variable_1 = Var1,
    Variable_2 = Var2,
    Correlation = Freq
  ) %>%
  filter(as.character(Variable_1) < as.character(Variable_2)) %>%
  mutate(abs_correlation = abs(Correlation)) %>%
  arrange(desc(abs_correlation)) %>%
  slice_head(n = 10)
```

:::

### 5.11.2 Plotting strongest correlation pairs

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5

strong_pairs_plot <- strong_pairs %>%
  mutate(pair = paste(Variable_1, "vs", Variable_2))

ggplot(
  strong_pairs_plot,
  aes(
    x = fct_reorder(pair, abs_correlation),
    y = Correlation,
    fill = Correlation
  )
) +
  geom_col(
    colour = my_dark,
    linewidth = 0.25
  ) +
  coord_flip() +
  scale_fill_gradient2(
    low = my_red,
    mid = my_light_teal,
    high = my_teal,
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Top 10 strongest pairwise correlations",
    subtitle = "This helps identify possible multicollinearity before modelling",
    x = NULL,
    y = "Correlation coefficient",
    fill = "Correlation"
  ) +
  theme_yy
```

## Code

```{r}
strong_pairs_plot <- strong_pairs %>%
  mutate(pair = paste(Variable_1, "vs", Variable_2))

ggplot(
  strong_pairs_plot,
  aes(
    x = fct_reorder(pair, abs_correlation),
    y = Correlation,
    fill = Correlation
  )
) +
  geom_col(
    colour = my_dark,
    linewidth = 0.25
  ) +
  coord_flip() +
  scale_fill_gradient2(
    low = my_red,
    mid = my_light_teal,
    high = my_teal,
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Top 10 strongest pairwise correlations",
    subtitle = "This helps identify possible multicollinearity before modelling",
    x = NULL,
    y = "Correlation coefficient",
    fill = "Correlation"
  ) +
  theme_yy
```

:::

This is useful because strong correlations between predictors can make later regression results harder to interpret.

## 5.12 Final Takeaway

Visual correlation analysis helps summarise many pairwise relationships in a compact way. The `pairs()` method shows the raw scatterplot patterns, while `ggcorrmat()` and `corrplot()` provide cleaner correlation matrix views.

For the wine quality dataset, correlation analysis is useful for identifying chemical variables that move together and variables that may be more related to quality. However, correlation should be treated as exploratory evidence, not as causal proof.

::: {.callout-tip}
## Final takeaway

Use scatterplot matrices when you need to inspect raw pairwise patterns.  
Use corrgrams when the goal is to compare many correlations quickly and detect structure.
:::

## 5.13 References

This exercise covered:

- correlation coefficient interpretation,
- scatterplot matrix using `pairs()`,
- lower and upper triangle matrix display,
- custom correlation coefficient panels,
- `ggstatsplot::ggcorrmat()`,
- grouped correlation matrix by wine type,
- `corrplot()` visual geometrics,
- lower-triangle and mixed corrgram layouts,
- significance testing with p-values,
- correlation matrix reordering,
- interactive correlation heatmap using `plotly`,
- extensions on wine quality drivers and strongest correlation pairs.