---
title: "Hands-on Exercise 5d - Visual Multivariate Analysis with Parallel Coordinates Plot"
author: "Yang Yang"
format:
  html:
    theme: sketchy
    toc: true
    toc-depth: 3
    number-sections: false
    code-copy: true
    code-overflow: wrap
editor: visual
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

if (!require(pacman)) {
  install.packages("pacman")
}

pacman::p_load(
  tidyverse,
  GGally,
  parallelPlot,
  DT,
  janitor,
  scales,
  plotly,
  knitr
)

my_teal <- "#2A6F6F"
my_soft_teal <- "#8AA6A3"
my_dark <- "#1F1F1F"
my_light_teal <- "#D8E8E6"
my_grey <- "#F3F4F4"
my_warning <- "#B07C3E"
my_red <- "#9A4B4B"

theme_yy <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", colour = my_dark),
    plot.subtitle = element_text(colour = my_teal),
    axis.title = element_text(colour = my_dark),
    axis.text = element_text(colour = my_dark),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## 5.1 Overview

Parallel coordinates plot is designed for visualising multivariate numerical data. Instead of comparing only two variables at a time, it places many numerical variables on parallel vertical axes and draws one line for each observation.

In this exercise, we use the World Happiness 2018 dataset. Each country is represented by one line, and the variables include happiness score, GDP per capita, social support, healthy life expectancy, freedom, generosity, and perceptions of corruption.

By the end of this exercise, we will be able to:

-   create static parallel coordinates plots using `ggparcoord()` from the `GGally` package,
-   group parallel coordinate lines by geographical region,
-   improve readability using scaling, transparency, boxplots, facets, and rotated labels,
-   create interactive parallel coordinates plots using `parallelPlot`,
-   add a small extension to compare regional happiness profiles.

::: callout-important
## Important

Parallel coordinates plot is mainly an analytical visualisation.\
It is useful for exploring multivariate patterns, but it can become visually crowded when there are many observations.
:::

## 5.2 Installing and Launching R Packages

For this exercise, the main packages are `GGally`, `parallelPlot`, and `tidyverse`.

::: panel-tabset
## View

```{r}
#| echo: false

package_table <- tibble(
  Package = c("tidyverse", "GGally", "parallelPlot", "DT", "janitor", "scales"),
  Purpose = c(
    "Data import, wrangling, and ggplot2 workflow",
    "Static parallel coordinates plot using ggparcoord()",
    "Interactive parallel coordinates plot",
    "Interactive HTML tables",
    "Cleaning column names when needed",
    "Number formatting and rescaling"
  )
)

DT::datatable(
  package_table,
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t", pageLength = 6),
  width = "100%"
)
```

## Code

```{r}
#| eval: false

pacman::p_load(
  tidyverse,
  GGally,
  parallelPlot,
  DT,
  janitor,
  scales
)
```
:::

## 5.3 Data Preparation

The dataset used in this exercise is `WHData-2018.csv`. It contains country-level World Happiness indicators for 2018.

### 5.3.1 Importing data

::: panel-tabset
## View

```{r}
#| echo: false

csv_file <- list.files(pattern = "WHData-2018.*\\.csv$")[1]

wh <- read_csv(csv_file, show_col_types = FALSE)

glimpse(wh)
```

## Code

```{r}
csv_file <- list.files(pattern = "WHData-2018.*\\.csv$")[1]

wh <- read_csv(csv_file, show_col_types = FALSE)

glimpse(wh)
```
:::

Using `list.files()` makes the render safer because the code will still work if the file name contains a small suffix such as `(2)`.

### 5.3.2 Checking the data structure

::: panel-tabset
## View

```{r}
#| echo: false

DT::datatable(
  head(wh, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
DT::datatable(
  head(wh, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```
:::

The key numerical variables are located from column 7 to column 12:

-   GDP per capita
-   Social support
-   Healthy life expectancy
-   Freedom to make life choices
-   Generosity
-   Perceptions of corruption

### 5.3.3 Creating colour palette by region

::: panel-tabset
## View

```{r}
#| echo: false

region_levels <- sort(unique(wh$Region))

region_palette <- setNames(
  rep(
    c(
      my_teal,
      my_soft_teal,
      my_warning,
      my_red,
      "#4F7C82",
      "#7A9E7E",
      "#C9A66B",
      "#6E7582",
      "#9B7E93",
      "#3F5E5A"
    ),
    length.out = length(region_levels)
  ),
  region_levels
)

tibble(
  Region = names(region_palette),
  Colour = unname(region_palette)
) %>%
  DT::datatable(
    rownames = FALSE,
    class = "compact stripe hover",
    options = list(dom = "t", pageLength = 10),
    width = "100%"
  )
```

## Code

```{r}
region_levels <- sort(unique(wh$Region))

region_palette <- setNames(
  rep(
    c(
      my_teal,
      my_soft_teal,
      my_warning,
      my_red,
      "#4F7C82",
      "#7A9E7E",
      "#C9A66B",
      "#6E7582",
      "#9B7E93",
      "#3F5E5A"
    ),
    length.out = length(region_levels)
  ),
  region_levels
)
```
:::

The palette keeps the page consistent with the deep muted teal style while still giving enough colours for the different regions.

## 5.4 Plotting Static Parallel Coordinates Plot

This section follows the `ggparcoord()` method from the `GGally` package.

### 5.4.1 Plotting a simple parallel coordinates plot

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5

GGally::ggparcoord(
  data = wh,
  columns = c(7:12)
) +
  labs(
    title = "Simple parallel coordinates plot of World Happiness variables"
  ) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1),
    legend.position = "none"
  )
```

## Code

```{r}
GGally::ggparcoord(
  data = wh,
  columns = c(7:12)
) +
  labs(
    title = "Simple parallel coordinates plot of World Happiness variables"
  ) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1),
    legend.position = "none"
  )
```
:::

This first version is useful as a starting point, but it is visually crowded. Since all countries are drawn together, the regional pattern is not easy to see.

### 5.4.2 Plotting a parallel coordinates plot with boxplot

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5

GGally::ggparcoord(
  data = wh,
  columns = c(7:12),
  groupColumn = 2,
  scale = "uniminmax",
  alphaLines = 0.2,
  boxplot = TRUE,
  title = "Parallel Coordinates Plot of World Happiness Variables"
) +
  scale_colour_manual(values = region_palette) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1)
  )
```

## Code

```{r}
GGally::ggparcoord(
  data = wh,
  columns = c(7:12),
  groupColumn = 2,
  scale = "uniminmax",
  alphaLines = 0.2,
  boxplot = TRUE,
  title = "Parallel Coordinates Plot of World Happiness Variables"
) +
  scale_colour_manual(values = region_palette) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1)
  )
```
:::

The `groupColumn` argument colours the lines by region. The `scale = "uniminmax"` argument rescales each variable from 0 to 1, making variables with different units comparable.

::: callout-tip
## What can we learn?

The boxplot layer gives a quick view of the distribution of each variable.\
The lines show country-level profiles, while the boxplots summarise the overall spread.
:::

### 5.4.3 Parallel coordinates with facet

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 8

GGally::ggparcoord(
  data = wh,
  columns = c(7:12),
  groupColumn = 2,
  scale = "uniminmax",
  alphaLines = 0.2,
  boxplot = TRUE,
  title = "Multiple Parallel Coordinates Plots of World Happiness Variables by Region"
) +
  scale_colour_manual(values = region_palette) +
  facet_wrap(~ Region) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1),
    legend.position = "none"
  )
```

## Code

```{r}
GGally::ggparcoord(
  data = wh,
  columns = c(7:12),
  groupColumn = 2,
  scale = "uniminmax",
  alphaLines = 0.2,
  boxplot = TRUE,
  title = "Multiple Parallel Coordinates Plots of World Happiness Variables by Region"
) +
  scale_colour_manual(values = region_palette) +
  facet_wrap(~ Region) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1),
    legend.position = "none"
  )
```
:::

Faceting makes the plot easier to read because each region is shown separately. This is helpful when the full plot is too dense.

### 5.4.4 Rotating x-axis text labels

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 8

GGally::ggparcoord(
  data = wh,
  columns = c(7:12),
  groupColumn = 2,
  scale = "uniminmax",
  alphaLines = 0.2,
  boxplot = TRUE,
  title = "Parallel Coordinates Plot with Rotated Axis Labels"
) +
  scale_colour_manual(values = region_palette) +
  facet_wrap(~ Region) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 30),
    legend.position = "none"
  )
```

## Code

```{r}
GGally::ggparcoord(
  data = wh,
  columns = c(7:12),
  groupColumn = 2,
  scale = "uniminmax",
  alphaLines = 0.2,
  boxplot = TRUE,
  title = "Parallel Coordinates Plot with Rotated Axis Labels"
) +
  scale_colour_manual(values = region_palette) +
  facet_wrap(~ Region) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 30),
    legend.position = "none"
  )
```
:::

Rotating the x-axis labels reduces label overlap, but the text may still sit too close to the plot.

### 5.4.5 Adjusting the rotated x-axis text labels

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 8

GGally::ggparcoord(
  data = wh,
  columns = c(7:12),
  groupColumn = 2,
  scale = "uniminmax",
  alphaLines = 0.2,
  boxplot = TRUE,
  title = "Adjusted Parallel Coordinates Plot by Region"
) +
  scale_colour_manual(values = region_palette) +
  facet_wrap(~ Region) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none"
  )
```

## Code

```{r}
GGally::ggparcoord(
  data = wh,
  columns = c(7:12),
  groupColumn = 2,
  scale = "uniminmax",
  alphaLines = 0.2,
  boxplot = TRUE,
  title = "Adjusted Parallel Coordinates Plot by Region"
) +
  scale_colour_manual(values = region_palette) +
  facet_wrap(~ Region) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none"
  )
```
:::

The `hjust = 1` setting aligns the rotated labels better and makes the plot easier to read.

## 5.5 Plotting Interactive Parallel Coordinates Plot

The `parallelPlot` package creates interactive parallel coordinates plots using `htmlwidgets`.

### 5.5.1 Preparing data for interactive plot

::: panel-tabset
## View

```{r}
#| echo: false

wh_parallel <- wh %>%
  select(
    `Happiness score`,
    `GDP per capita`,
    `Social support`,
    `Healthy life expectancy`,
    `Freedom to make life choices`,
    Generosity,
    `Perceptions of corruption`
  )

DT::datatable(
  head(wh_parallel, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
wh_parallel <- wh %>%
  select(
    `Happiness score`,
    `GDP per capita`,
    `Social support`,
    `Healthy life expectancy`,
    `Freedom to make life choices`,
    Generosity,
    `Perceptions of corruption`
  )

DT::datatable(
  head(wh_parallel, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```
:::

Only numerical variables are selected for the interactive parallel coordinates plot.

### 5.5.2 The basic interactive plot

::: panel-tabset
## View

```{r}
#| echo: false

parallelPlot::parallelPlot(
  wh_parallel,
  width = 760,
  height = 420
)
```

## Code

```{r}
parallelPlot::parallelPlot(
  wh_parallel,
  width = 760,
  height = 420
)
```
:::

The basic interactive plot allows the reader to inspect patterns across several variables. However, some axis labels are long and can overlap.

### 5.5.3 Rotating axis labels

::: panel-tabset
## View

```{r}
#| echo: false

parallelPlot::parallelPlot(
  wh_parallel,
  rotateTitle = TRUE,
  width = 760,
  height = 420
)
```

## Code

```{r}
parallelPlot::parallelPlot(
  wh_parallel,
  rotateTitle = TRUE,
  width = 760,
  height = 420
)
```
:::

The `rotateTitle = TRUE` argument makes the axis labels easier to read.

### 5.5.4 Changing the colour scheme

::: panel-tabset
## View

```{r}
#| echo: false

parallelPlot::parallelPlot(
  wh_parallel,
  continuousCS = "YlOrRd",
  rotateTitle = TRUE,
  width = 760,
  height = 420
)
```

## Code

```{r}
parallelPlot::parallelPlot(
  wh_parallel,
  continuousCS = "YlOrRd",
  rotateTitle = TRUE,
  width = 760,
  height = 420
)
```
:::

The `continuousCS` argument changes the continuous colour scheme. This helps when the reader wants stronger colour contrast during interactive inspection.

### 5.5.5 Parallel coordinates plot with histogram

::: panel-tabset
## View

```{r}
#| echo: false

histoVisibility <- rep(TRUE, ncol(wh_parallel))

parallelPlot::parallelPlot(
  wh_parallel,
  rotateTitle = TRUE,
  histoVisibility = histoVisibility,
  width = 760,
  height = 420
)
```

## Code

```{r}
histoVisibility <- rep(TRUE, ncol(wh_parallel))

parallelPlot::parallelPlot(
  wh_parallel,
  rotateTitle = TRUE,
  histoVisibility = histoVisibility,
  width = 760,
  height = 420
)
```
:::

The histogram layer shows the distribution of each variable along its axis. This makes the interactive plot more informative.

::: callout-tip
## What can we learn?

Interactive parallel coordinates are useful because users can inspect relationships dynamically.\
The histogram option adds distribution information to each axis.
:::

## 5.6 Extra Analysis: Regional Average Profile

This small extension summarises the average World Happiness profile by region. It is still within the parallel coordinates topic because it reduces country-level noise and compares regional profiles.

### 5.6.1 Preparing regional average data

::: panel-tabset
## View

```{r}
#| echo: false

region_profile <- wh %>%
  group_by(Region) %>%
  summarise(
    `Happiness score` = mean(`Happiness score`, na.rm = TRUE),
    `GDP per capita` = mean(`GDP per capita`, na.rm = TRUE),
    `Social support` = mean(`Social support`, na.rm = TRUE),
    `Healthy life expectancy` = mean(`Healthy life expectancy`, na.rm = TRUE),
    `Freedom to make life choices` = mean(`Freedom to make life choices`, na.rm = TRUE),
    Generosity = mean(Generosity, na.rm = TRUE),
    `Perceptions of corruption` = mean(`Perceptions of corruption`, na.rm = TRUE),
    .groups = "drop"
  )

DT::datatable(
  region_profile %>%
    mutate(across(where(is.numeric), ~ round(.x, 3))),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 10, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
region_profile <- wh %>%
  group_by(Region) %>%
  summarise(
    `Happiness score` = mean(`Happiness score`, na.rm = TRUE),
    `GDP per capita` = mean(`GDP per capita`, na.rm = TRUE),
    `Social support` = mean(`Social support`, na.rm = TRUE),
    `Healthy life expectancy` = mean(`Healthy life expectancy`, na.rm = TRUE),
    `Freedom to make life choices` = mean(`Freedom to make life choices`, na.rm = TRUE),
    Generosity = mean(Generosity, na.rm = TRUE),
    `Perceptions of corruption` = mean(`Perceptions of corruption`, na.rm = TRUE),
    .groups = "drop"
  )

DT::datatable(
  region_profile %>%
    mutate(across(where(is.numeric), ~ round(.x, 3))),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 10, scrollX = TRUE),
  width = "100%"
)
```
:::

### 5.6.2 Parallel coordinates plot of regional averages

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5

GGally::ggparcoord(
  data = region_profile,
  columns = c(2:8),
  groupColumn = 1,
  scale = "uniminmax",
  alphaLines = 0.8,
  title = "Parallel Coordinates Plot of Regional Average Happiness Profiles"
) +
  scale_colour_manual(values = region_palette) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

## Code

```{r}
GGally::ggparcoord(
  data = region_profile,
  columns = c(2:8),
  groupColumn = 1,
  scale = "uniminmax",
  alphaLines = 0.8,
  title = "Parallel Coordinates Plot of Regional Average Happiness Profiles"
) +
  scale_colour_manual(values = region_palette) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```
:::

This regional version is less crowded than the country-level plot. It helps compare the broad profile of each region.

## 5.7 Extra Analysis: Top and Bottom Happiness Countries

This extension compares the highest and lowest happiness score countries. It is useful because parallel coordinates can show whether high-happiness countries share similar profiles across GDP, health, social support, and freedom.

### 5.7.1 Preparing top and bottom countries

::: panel-tabset
## View

```{r}
#| echo: false

top_bottom_countries <- bind_rows(
  wh %>%
    arrange(desc(`Happiness score`)) %>%
    slice_head(n = 8) %>%
    mutate(Group = "Top 8"),
  wh %>%
    arrange(`Happiness score`) %>%
    slice_head(n = 8) %>%
    mutate(Group = "Bottom 8")
)

DT::datatable(
  top_bottom_countries %>%
    select(
      Country,
      Region,
      Group,
      `Happiness score`,
      `GDP per capita`,
      `Social support`,
      `Healthy life expectancy`,
      `Freedom to make life choices`,
      Generosity,
      `Perceptions of corruption`
    ),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 16, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
top_bottom_countries <- bind_rows(
  wh %>%
    arrange(desc(`Happiness score`)) %>%
    slice_head(n = 8) %>%
    mutate(Group = "Top 8"),
  wh %>%
    arrange(`Happiness score`) %>%
    slice_head(n = 8) %>%
    mutate(Group = "Bottom 8")
)
```
:::

### 5.7.2 Parallel coordinates plot for top and bottom countries

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5

GGally::ggparcoord(
  data = top_bottom_countries,
  columns = c(4, 7:12),
  groupColumn = "Group",
  scale = "uniminmax",
  alphaLines = 0.75,
  boxplot = TRUE,
  title = "Top and Bottom Happiness Countries: Parallel Coordinates Comparison"
) +
  scale_colour_manual(
    values = c(
      "Top 8" = my_teal,
      "Bottom 8" = my_red
    )
  ) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

## Code

```{r}
GGally::ggparcoord(
  data = top_bottom_countries,
  columns = c(4, 7:12),
  groupColumn = "Group",
  scale = "uniminmax",
  alphaLines = 0.75,
  boxplot = TRUE,
  title = "Top and Bottom Happiness Countries: Parallel Coordinates Comparison"
) +
  scale_colour_manual(
    values = c(
      "Top 8" = my_teal,
      "Bottom 8" = my_red
    )
  ) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```
:::

The top and bottom group comparison makes the multivariate contrast clearer than looking at all 156 countries at once.

## 5.8 Extra Analysis: Relationship Strength with Happiness Score

This final extension supports the parallel coordinates analysis by showing which variables are most associated with happiness score.

### 5.8.1 Computing correlations with happiness score

::: panel-tabset
## View

```{r}
#| echo: false

happiness_cor <- wh %>%
  select(
    `Happiness score`,
    `GDP per capita`,
    `Social support`,
    `Healthy life expectancy`,
    `Freedom to make life choices`,
    Generosity,
    `Perceptions of corruption`
  ) %>%
  summarise(
    across(
      -`Happiness score`,
      ~ cor(.x, `Happiness score`, use = "pairwise.complete.obs")
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Correlation"
  ) %>%
  arrange(desc(Correlation))

DT::datatable(
  happiness_cor %>%
    mutate(Correlation = round(Correlation, 3)),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t", scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
happiness_cor <- wh %>%
  select(
    `Happiness score`,
    `GDP per capita`,
    `Social support`,
    `Healthy life expectancy`,
    `Freedom to make life choices`,
    Generosity,
    `Perceptions of corruption`
  ) %>%
  summarise(
    across(
      -`Happiness score`,
      ~ cor(.x, `Happiness score`, use = "pairwise.complete.obs")
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Correlation"
  ) %>%
  arrange(desc(Correlation))
```
:::

### 5.8.2 Bar chart of correlations with happiness score

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5

ggplot(
  happiness_cor,
  aes(
    x = fct_reorder(Variable, Correlation),
    y = Correlation
  )
) +
  geom_col(
    fill = my_teal,
    colour = my_dark,
    linewidth = 0.3
  ) +
  coord_flip() +
  labs(
    title = "Correlation with Happiness Score",
    subtitle = "This extension helps interpret which axes matter more in the parallel coordinates plot",
    x = NULL,
    y = "Correlation"
  ) +
  theme_yy
```

## Code

```{r}
ggplot(
  happiness_cor,
  aes(
    x = fct_reorder(Variable, Correlation),
    y = Correlation
  )
) +
  geom_col(
    fill = my_teal,
    colour = my_dark,
    linewidth = 0.3
  ) +
  coord_flip() +
  labs(
    title = "Correlation with Happiness Score",
    subtitle = "This extension helps interpret which axes matter more in the parallel coordinates plot",
    x = NULL,
    y = "Correlation"
  ) +
  theme_yy
```
:::

This bar chart is not a replacement for the parallel coordinates plot. It is a supporting view that helps explain which variables are most closely related to happiness score.

## 5.9 Final Takeaway

Parallel coordinates plots are useful for exploring high-dimensional numerical data. They allow many variables to be compared in one visual frame.

In this exercise, the static `ggparcoord()` plots help compare World Happiness indicators by region. The interactive `parallelPlot` charts allow more flexible inspection. The extra regional and top-bottom comparisons reduce visual clutter and make the interpretation clearer.

::: callout-tip
## Final takeaway

Use parallel coordinates plots when the key question is about multivariate profiles.\
They are strongest for exploration, comparison, and pattern detection across many numerical variables.
:::

## 5.10 References

This exercise covered:

-   `ggparcoord()` from the `GGally` package,
-   static parallel coordinates plots,
-   grouped and faceted parallel coordinates plots,
-   rotated and adjusted axis labels,
-   interactive parallel coordinates plots using `parallelPlot`,
-   histogram-enhanced interactive plots,
-   regional average profiles,
-   top-versus-bottom happiness country comparison.

::: vaa-exercise-footer
### Portfolio note

**Key takeaway.** Parallel coordinates expose multivariate profiles and trade-offs across dimensions.

**Practical use.** Use it when individual observations need to be compared over many variables.

::: vaa-footer-nav
[Back to Gallery](../../hands-on-gallery.html) [Previous: Exercise 5c](../Hands-on_Ex05c/Hands-on_Ex05c.html) [Next: Exercise 5e](../Hands-on_Ex05e/Hands-on_Ex05e.html)
:::
:::

