---
title: "Hands-on Exercise 10 - Information Dashboard Design: R methods"
author: "Yang Yang"
format:
  html:
    theme: sketchy
    toc: true
    toc-depth: 3
    number-sections: false
    code-copy: true
    code-overflow: wrap
editor: visual
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

user_lib <- Sys.getenv("R_LIBS_USER")
if (nzchar(user_lib)) {
  dir.create(user_lib, recursive = TRUE, showWarnings = FALSE)
  .libPaths(c(user_lib, .libPaths()))
}

if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

pacman::p_load(
  lubridate,
  ggthemes,
  reactable,
  reactablefmtr,
  gt,
  gtExtras,
  tidyverse,
  RODBC,
  dataui,
  DT,
  knitr,
  scales
)

my_teal <- "#2A6F6F"
my_soft_teal <- "#8AA6A3"
my_dark <- "#1F1F1F"
my_light_teal <- "#D8E8E6"
my_grey <- "#F3F4F4"
my_warning <- "#B07C3E"
my_red <- "#9A4B4B"
my_blue <- "#4C78A8"

find_data_file <- function(file_name, subfolder = NULL) {
  candidates <- if (is.null(subfolder)) {
    c(
      file.path("data", file_name),
      file.path("Hands-on_Ex", "Hands-on_Ex10", "data", file_name)
    )
  } else {
    c(
      file.path("data", subfolder, file_name),
      file.path("Hands-on_Ex", "Hands-on_Ex10", "data", subfolder, file_name)
    )
  }

  found <- candidates[file.exists(candidates)]

  if (length(found) == 0) {
    stop(
      "Missing required data file: ", file_name,
      ". Put it inside Hands-on_Ex/Hands-on_Ex10/data/ before rendering."
    )
  }

  found[1]
}

theme_yy <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", colour = my_dark),
    plot.subtitle = element_text(colour = my_teal),
    axis.title = element_text(colour = my_dark),
    axis.text = element_text(colour = my_dark),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## 31.1 Overview

This hands-on exercise follows [Chapter 31 of R4VA](https://r4va.netlify.app/chap31), *Information Dashboard Design: R methods*. The exercise uses the Coffee Chain Microsoft Access database to build compact dashboard components.

By the end of the exercise, the following tasks are practised:

- importing data from a Microsoft Access database;
- creating bullet charts with `ggplot2`;
- creating sparklines with `ggplot2`;
- building static dashboard tables with `gt` and `gtExtras`; and
- building interactive dashboard tables with `reactable`, `reactablefmtr`, and `dataui`.

## 31.2 Getting Started

The packages used in this exercise are loaded in the setup chunk.

```{r}
#| label: package-code
#| eval: false

if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

pacman::p_load(
  lubridate,
  ggthemes,
  reactable,
  reactablefmtr,
  gt,
  gtExtras,
  tidyverse,
  RODBC,
  dataui,
  DT,
  knitr,
  scales
)
```

```{r}
#| label: package-table

tribble(
  ~Package, ~Purpose,
  "tidyverse", "Data import, wrangling, summarising, and ggplot2 graphics.",
  "lubridate", "Extracting date parts such as month and year.",
  "ggthemes", "Using theme_tufte() for compact sparkline plots.",
  "gt", "Creating static dashboard tables.",
  "gtExtras", "Adding bullet charts and sparklines to gt tables.",
  "reactable", "Creating interactive HTML tables.",
  "reactablefmtr", "Adding formatted sparklines and bars to reactable tables.",
  "dataui", "Underlying sparkline widgets used by reactablefmtr examples.",
  "RODBC", "Original Chapter 31 Access database connection method.",
  "DT", "Interactive preview tables."
) |>
  datatable(
    rownames = FALSE,
    options = list(pageLength = 10, dom = "tip")
  )
```

## 31.3 Importing Microsoft Access Database

### 31.3.1 The Data Set

The data source is a Microsoft Access database named `Coffee Chain.mdb`. The saved query referenced in the chapter is `CoffeeChain Query`. On macOS, the 32-bit Windows `RODBC::odbcConnectAccess2007()` workflow is not available, so this exercise uses `mdbtools` to read the Access tables and reconstruct the saved query using its selected fields and join keys.

### 31.3.2 Importing Database into R

The original Chapter 31 code is shown below for reference.

```{r}
#| label: access-import-windows-code
#| eval: false

library(RODBC)
con <- odbcConnectAccess2007("data/Coffee Chain.mdb")
coffeechain <- sqlFetch(con, "CoffeeChain Query")
write_rds(coffeechain, "data/rds/CoffeeChain.rds")
odbcClose(con)
```

The evaluated code below is the macOS-compatible import used in this document. It reads the three base tables from the MDB, joins them by `Area Code` and `ProductId`, then saves the result as `data/rds/CoffeeChain.rds`.

```{r}
#| label: access-import-macos

mdb_file <- find_data_file("Coffee Chain.mdb")
rds_file <- file.path("data", "rds", "CoffeeChain.rds")
dir.create(dirname(rds_file), recursive = TRUE, showWarnings = FALSE)

read_mdb_table <- function(table_name) {
  if (Sys.which("mdb-export") == "") {
    stop(
      "mdb-export was not found. Install mdbtools before rebuilding CoffeeChain.rds."
    )
  }

  exported <- system2(
    "mdb-export",
    c(shQuote(mdb_file), shQuote(table_name)),
    stdout = TRUE
  )

  read_csv(I(paste(exported, collapse = "\n")), show_col_types = FALSE)
}

fact_table <- read_mdb_table("factTable")
location_table <- read_mdb_table("Location")
product_table <- read_mdb_table("Product")

coffeechain <- fact_table |>
  left_join(location_table, by = "Area Code") |>
  left_join(product_table, by = "ProductId") |>
  transmute(
    Profit,
    Margin,
    Sales,
    COGS,
    `Total Expenses`,
    Marketing,
    Inventory,
    `Budget Profit`,
    `Budget Margin`,
    `Budget Sales`,
    `Budget COGS`,
    Date = mdy_hms(Date),
    Market,
    State,
    `Area Code`,
    `Market Size`,
    `Product Type`,
    Product,
    Type
  )

write_rds(coffeechain, rds_file)
```

### 31.3.3 Data Preparation

The generated RDS file is imported for the dashboard work.

```{r}
#| label: read-coffeechain-rds

coffeechain <- read_rds(find_data_file("CoffeeChain.rds", "rds"))

coffeechain |>
  slice_head(n = 10) |>
  datatable(
    rownames = FALSE,
    options = list(pageLength = 10, scrollX = TRUE, dom = "tip")
  )
```

```{r}
#| label: coffeechain-summary

tibble(
  measure = c(
    "Rows",
    "Columns",
    "Products",
    "Markets",
    "Date range"
  ),
  value = c(
    nrow(coffeechain),
    ncol(coffeechain),
    n_distinct(coffeechain$Product),
    n_distinct(coffeechain$Market),
    paste(min(coffeechain$Date), "to", max(coffeechain$Date))
  )
) |>
  kable()
```

The code below aggregates actual and budgeted sales at product level.

```{r}
#| label: product-summary

product <- coffeechain |>
  group_by(Product) |>
  summarise(
    target = sum(`Budget Sales`),
    current = sum(Sales),
    .groups = "drop"
  )

product |>
  datatable(
    rownames = FALSE,
    options = list(pageLength = 13, dom = "tip")
  ) |>
  formatCurrency(columns = c("target", "current"), currency = "$", digits = 0)
```

### 31.3.4 Bullet Chart in ggplot2

::: panel-tabset

## View

```{r}
#| label: ggplot-bullet-view
#| echo: false
#| fig-width: 9
#| fig-height: 6

ggplot(product, aes(Product, current)) +
  geom_col(
    aes(y = max(target) * 1.01),
    fill = "grey85",
    width = 0.85
  ) +
  geom_col(
    aes(y = target * 0.75),
    fill = "grey60",
    width = 0.85
  ) +
  geom_col(
    aes(y = target * 0.50),
    fill = "grey50",
    width = 0.85
  ) +
  geom_col(
    aes(y = current),
    width = 0.35,
    fill = "black"
  ) +
  geom_errorbar(
    aes(y = target, ymin = target, ymax = target),
    width = 0.4,
    colour = my_red,
    linewidth = 1
  ) +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Product Sales Against Budget Sales",
    x = NULL,
    y = "Sales"
  ) +
  theme_yy
```

## Code

```{r}
#| label: ggplot-bullet-code
#| eval: false

ggplot(product, aes(Product, current)) +
  geom_col(aes(y = max(target) * 1.01), fill = "grey85", width = 0.85) +
  geom_col(aes(y = target * 0.75), fill = "grey60", width = 0.85) +
  geom_col(aes(y = target * 0.50), fill = "grey50", width = 0.85) +
  geom_col(aes(y = current), width = 0.35, fill = "black") +
  geom_errorbar(
    aes(y = target, ymin = target, ymax = target),
    width = 0.4,
    colour = "red",
    linewidth = 1
  ) +
  coord_flip()
```

:::

## 31.4 Plotting Sparklines Using ggplot2

### 31.4.1 Preparing the Data

```{r}
#| label: prepare-sparkline-data

sales_report <- coffeechain |>
  filter(Date >= as.Date("2013-01-01")) |>
  mutate(Month = month(Date)) |>
  group_by(Month, Product) |>
  summarise(Sales = sum(Sales), .groups = "drop") |>
  select(Month, Product, Sales)

mins <- sales_report |>
  group_by(Product) |>
  slice_min(Sales, n = 1, with_ties = FALSE) |>
  ungroup()

maxs <- sales_report |>
  group_by(Product) |>
  slice_max(Sales, n = 1, with_ties = FALSE) |>
  ungroup()

ends <- sales_report |>
  group_by(Product) |>
  filter(Month == max(Month)) |>
  ungroup()

quarts <- sales_report |>
  group_by(Product) |>
  summarise(
    quart1 = quantile(Sales, 0.25),
    quart2 = quantile(Sales, 0.75),
    .groups = "drop"
  ) |>
  right_join(sales_report, by = "Product")
```

### 31.4.2 Sparklines in ggplot2

::: panel-tabset

## View

```{r}
#| label: ggplot-sparklines-view
#| echo: false
#| fig-width: 9
#| fig-height: 7

ggplot(sales_report, aes(x = Month, y = Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(
    data = quarts,
    aes(ymin = quart1, ymax = quart2),
    fill = "grey90"
  ) +
  geom_line(linewidth = 0.3) +
  geom_point(data = mins, colour = my_red, size = 0.9) +
  geom_point(data = maxs, colour = my_blue, size = 0.9) +
  geom_text(data = mins, aes(label = Sales), vjust = -1, size = 2) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5, size = 2) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5, size = 2) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0, size = 2) +
  expand_limits(
    x = max(sales_report$Month) +
      (0.25 * (max(sales_report$Month) - min(sales_report$Month)))
  ) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 8, base_family = "Helvetica") +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    strip.text = element_blank()
  )
```

## Code

```{r}
#| label: ggplot-sparklines-code
#| eval: false

ggplot(sales_report, aes(x = Month, y = Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(
    data = quarts,
    aes(ymin = quart1, ymax = quart2),
    fill = "grey90"
  ) +
  geom_line(linewidth = 0.3) +
  geom_point(data = mins, colour = "red") +
  geom_point(data = maxs, colour = "blue") +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  theme_tufte(base_size = 8, base_family = "Helvetica")
```

:::

## 31.5 Static Information Dashboard Design: gt and gtExtras Methods

### 31.5.1 Plotting a Simple Bullet Chart

```{r}
#| label: gt-bullet

product |>
  gt(id = "product_bullet") |>
  gt_plt_bullet(
    column = current,
    target = target,
    width = 60,
    palette = c("lightblue", "black")
  ) |>
  gt_theme_538(quiet = TRUE)
```

## 31.6 Sparklines: gtExtras Method

### 31.6.1 Preparing the CoffeeChain Sales Report

```{r}
#| label: prepare-gt-report

report <- coffeechain |>
  mutate(Year = year(Date)) |>
  filter(Year == 2013) |>
  mutate(
    Month = month(Date, label = TRUE, abbr = TRUE)
  ) |>
  group_by(Product, Month) |>
  summarise(Sales = sum(Sales), .groups = "drop") |>
  arrange(Product, Month)

spark <- report |>
  group_by(Product) |>
  summarise(
    `Monthly Sales` = list(Sales),
    .groups = "drop"
  )

spark |>
  datatable(
    rownames = FALSE,
    options = list(pageLength = 13, dom = "tip")
  )
```

### 31.6.2 Plotting CoffeeChain Sales Report

```{r}
#| label: gt-sparkline-simple

spark |>
  gt(id = "sparkline_simple") |>
  gt_plt_sparkline(
    `Monthly Sales`,
    same_limit = FALSE
  )
```

### 31.6.3 Adding Statistics

```{r}
#| label: sales-stats

sales <- report |>
  group_by(Product) |>
  summarise(
    Min = min(Sales, na.rm = TRUE),
    Max = max(Sales, na.rm = TRUE),
    Average = mean(Sales, na.rm = TRUE),
    .groups = "drop"
  )

sales |>
  gt(id = "sales_stats") |>
  fmt_number(columns = Average, decimals = 2)
```

### 31.6.4 Combining the Data Frame

```{r}
#| label: combine-sales-spark

sales_data <- sales |>
  left_join(spark, by = "Product")

sales_data |>
  gt(id = "sales_data") |>
  gt_plt_sparkline(
    `Monthly Sales`,
    same_limit = FALSE
  ) |>
  fmt_number(columns = c(Min, Max, Average), decimals = 2)
```

### 31.6.5 Combining Bullet Chart and Sparklines

```{r}
#| label: combine-bullet-spark

bullet <- coffeechain |>
  filter(Date >= as.Date("2013-01-01")) |>
  group_by(Product) |>
  summarise(
    Target = sum(`Budget Sales`),
    Actual = sum(Sales),
    .groups = "drop"
  )

sales_dashboard <- sales_data |>
  left_join(bullet, by = "Product")

sales_dashboard |>
  gt(id = "sales_dashboard") |>
  gt_plt_sparkline(`Monthly Sales`) |>
  gt_plt_bullet(
    column = Actual,
    target = Target,
    width = 28,
    palette = c("lightblue", "black")
  ) |>
  fmt_number(columns = c(Min, Max, Average, Actual, Target), decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

## 31.7 Interactive Information Dashboard Design: reactable and reactablefmtr Methods

The interactive dashboard examples use a list-column version of the monthly sales report.

```{r}
#| label: prepare-interactive-report

interactive_report <- report |>
  group_by(Product) |>
  summarise(
    `Monthly Sales` = list(Sales),
    .groups = "drop"
  )
```

### 31.7.1 Plotting Interactive Sparklines

```{r}
#| label: reactable-sparkline

reactable(
  interactive_report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(interactive_report)
    )
  )
)
```

### 31.7.2 Changing the Page Size

```{r}
#| label: reactable-pagesize

reactable(
  interactive_report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(interactive_report)
    )
  )
)
```

### 31.7.3 Adding Points and Labels

```{r}
#| label: reactable-points-labels

reactable(
  interactive_report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        interactive_report,
        highlight_points = highlight_points(
          min = "red",
          max = "blue"
        ),
        labels = c("first", "last")
      )
    )
  )
)
```

### 31.7.4 Adding Reference Line

```{r}
#| label: reactable-reference-line

reactable(
  interactive_report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        interactive_report,
        highlight_points = highlight_points(
          min = "red",
          max = "blue"
        ),
        statline = "mean"
      )
    )
  )
)
```

### 31.7.5 Adding Bandline

```{r}
#| label: reactable-bandline

reactable(
  interactive_report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        interactive_report,
        highlight_points = highlight_points(
          min = "red",
          max = "blue"
        ),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
      )
    )
  )
)
```

### 31.7.6 Changing from Sparkline to Sparkbar

```{r}
#| label: reactable-sparkbar

reactable(
  interactive_report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        interactive_report,
        highlight_bars = highlight_bars(
          min = "red",
          max = "blue"
        ),
        bandline = "innerquartiles",
        statline = "mean"
      )
    )
  )
)
```

## Key Takeaways

::: {.callout-tip}
Dashboard design works best when each compact display has a clear role: bullet charts compare actual against target, sparklines show trend shape, and interactive tables let readers inspect the detail without leaving the dashboard.
:::

In this exercise:

- `mdbtools` is used to import the Access database reliably on macOS;
- `ggplot2` builds bullet charts and sparklines from prepared summaries;
- `gtExtras` embeds bullet charts and sparklines in static `gt` tables; and
- `reactablefmtr` embeds interactive sparklines and sparkbars in `reactable` tables.

## References

- [R4VA Chapter 31: Information Dashboard Design: R methods](https://r4va.netlify.app/chap31)
- [gt](https://gt.rstudio.com/)
- [gtExtras](https://jthomasmock.github.io/gtExtras/)
- [reactable](https://glin.github.io/reactable/)
- [reactablefmtr](https://kcuilla.github.io/reactablefmtr/)

::: vaa-exercise-footer
### Portfolio note

**Key takeaway.** Dashboard design turns multiple indicators into a compact monitoring view.

**Practical use.** Use it when decision makers need concise, repeatable, and scannable visual reporting.

::: vaa-footer-nav
[Back to Gallery](../../hands-on-gallery.html) [Previous: Exercise 9c](../Hands-on_Ex09c/Hands-on_Ex09c.html)
:::
:::

