---
title: "Hands-on Exercise 8 - Visualising and Analysing Time-oriented Data"
author: "Yang Yang"
format:
  html:
    theme: sketchy
    toc: true
    toc-depth: 3
    number-sections: false
    code-copy: true
    code-overflow: wrap
editor: visual
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

user_lib <- Sys.getenv("R_LIBS_USER")
if (nzchar(user_lib)) {
  dir.create(user_lib, recursive = TRUE, showWarnings = FALSE)
  .libPaths(c(user_lib, .libPaths()))
}

my_teal <- "#2A6F6F"
my_soft_teal <- "#8AA6A3"
my_dark <- "#1F1F1F"
my_light_teal <- "#D8E8E6"
my_grey <- "#F3F4F4"
my_warning <- "#B07C3E"
my_red <- "#9A4B4B"
my_blue <- "#4C78A8"

find_data_file <- function(file_name) {
  candidates <- c(
    file.path("data", file_name),
    file.path("Hands-on_Ex", "Hands-on_Ex08", "data", file_name)
  )
  found <- candidates[file.exists(candidates)]

  if (length(found) == 0) {
    stop(
      "Missing required data file: ", file_name,
      ". Put it inside Hands-on_Ex/Hands-on_Ex08/data/ before rendering."
    )
  }

  found[1]
}
```

## 17.1 Learning Outcome

This hands-on exercise follows [Chapter 17 of R4VA](https://r4va.netlify.app/chap17), *Visualising and Analysing Time-oriented Data*. By the end of this exercise, the following time-oriented visualisations will be created:

- calendar heatmap using `ggplot2`;
- multiple calendar heatmaps for the top source countries;
- cycle plot using `ggplot2`; and
- slopegraph for comparing two time points.

## 17.2 Getting Started

The code chunk below checks whether `pacman` is installed. If it is missing, it installs it from CRAN. This is included as executable code so the package setup is reproducible.

```{r}
#| label: install-pacman

if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}
```

## 17.3 Do It Yourself

The code chunk below checks, installs, and launches the packages needed for this exercise. The original Chapter 17 package list includes `CGPfunctions` for `newggslopegraph()`. That package has been removed from live CRAN and currently depends on an archived package that does not load cleanly with this R/ggplot2 setup. For this reason, the evaluated package setup installs and loads the packages used by the runnable exercise, and the slopegraph section uses a `ggplot2` fallback instead of failing.

```{r}
#| label: install-and-load-packages

required_packages <- c(
  "scales",
  "viridis",
  "lubridate",
  "ggthemes",
  "gridExtra",
  "readxl",
  "knitr",
  "data.table",
  "ggHoriPlot",
  "tidyverse",
  "DT"
)

pacman::p_load(char = required_packages)

cgp_available <- requireNamespace("CGPfunctions", quietly = TRUE)
```

The archived `CGPfunctions` installation command is shown below for reference. It is not evaluated in this document because its archived dependency `ggmosaic` does not currently load cleanly with this R/ggplot2 setup.

```{r}
#| label: optional-cgpfunctions-install-code
#| eval: false

install.packages(
  "https://cran.r-project.org/src/contrib/Archive/CGPfunctions/CGPfunctions_0.6.3.tar.gz",
  repos = NULL,
  type = "source"
)
```

```{r}
#| label: package-status

package_status <- tibble(
  package = c(required_packages, "CGPfunctions"),
  status = map_chr(
    package,
    ~ if (requireNamespace(.x, quietly = TRUE)) "Available" else "Not available"
  )
)

DT::datatable(
  package_status,
  rownames = FALSE,
  options = list(pageLength = 12, dom = "tip")
)
```

```{r}
#| label: theme
#| include: false

theme_yy <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", colour = my_dark),
    plot.subtitle = element_text(colour = my_teal),
    axis.title = element_text(colour = my_dark),
    axis.text = element_text(colour = my_dark),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", colour = my_dark),
    legend.position = "bottom"
  )
```

## 17.4 Plotting Calendar Heatmap

This section uses `eventlog.csv`, which contains cyber attack records by timestamp, source country, and time zone.

### 17.4.1 The Data

```{r}
#| label: import-eventlog

attacks_raw <- read_csv(find_data_file("eventlog.csv"), show_col_types = FALSE)

glimpse(attacks_raw)
```

### 17.4.2 Examining the Data Structure

```{r}
#| label: eventlog-head

kable(head(attacks_raw))
```

There are three columns in the data:

- `timestamp`: date-time value of the cyber attack record;
- `source_country`: ISO 3166-1 alpha-2 country code of the attack source; and
- `tz`: time zone of the source IP address.

### 17.4.3 Data Preparation

The calendar heatmap needs weekday and hour fields. The function below converts the timestamp by time zone, then derives the weekday and hour.

```{r}
#| label: make-hour-weekday

make_hr_wkday <- function(ts, sc, tz) {
  real_times <- ymd_hms(
    format(ts, "%Y-%m-%d %H:%M:%S"),
    tz = tz[1],
    quiet = TRUE
  )

  data.table(
    source_country = sc,
    wkday = weekdays(real_times),
    hour = hour(real_times)
  )
}
```

```{r}
#| label: prepare-attacks

wkday_levels <- c(
  "Saturday", "Friday", "Thursday", "Wednesday",
  "Tuesday", "Monday", "Sunday"
)

attacks <- attacks_raw |>
  group_by(tz) |>
  group_modify(
    ~ make_hr_wkday(
      ts = .x$timestamp,
      sc = .x$source_country,
      tz = .y$tz
    )
  ) |>
  ungroup() |>
  mutate(
    wkday = factor(wkday, levels = wkday_levels),
    hour = factor(hour, levels = 0:23)
  )

kable(head(attacks))
```

### 17.4.4 Building the Calendar Heatmap

::: panel-tabset

## View

```{r}
#| label: calendar-heatmap-view
#| echo: false
#| fig-width: 10
#| fig-height: 5.8

grouped <- attacks |>
  count(wkday, hour) |>
  ungroup() |>
  na.omit()

ggplot(
  grouped,
  aes(hour, wkday, fill = n)
) +
  geom_tile(color = "white", linewidth = 0.1) +
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_gradient(
    name = "# of attacks",
    low = "sky blue",
    high = "dark blue",
    labels = comma
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Attacks by Weekday and Time of Day"
  ) +
  theme(
    axis.ticks = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6)
  )
```

## Code

```{r}
#| label: calendar-heatmap-code
#| eval: false

grouped <- attacks |>
  count(wkday, hour) |>
  ungroup() |>
  na.omit()

ggplot(grouped, aes(hour, wkday, fill = n)) +
  geom_tile(color = "white", linewidth = 0.1) +
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_gradient(
    name = "# of attacks",
    low = "sky blue",
    high = "dark blue"
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Attacks by Weekday and Time of Day"
  )
```

:::

### 17.4.5 Building Multiple Calendar Heatmaps

The top four source countries are identified by attack count, then the calendar heatmap is faceted by country.

```{r}
#| label: attacks-by-country

attacks_by_country <- count(attacks, source_country) |>
  mutate(percent = percent(n / sum(n))) |>
  arrange(desc(n))

DT::datatable(
  attacks_by_country,
  rownames = FALSE,
  options = list(pageLength = 10, dom = "tip")
)
```

```{r}
#| label: prepare-top4-attacks

top4 <- attacks_by_country$source_country[1:4]

top4_attacks <- attacks |>
  filter(source_country %in% top4) |>
  count(source_country, wkday, hour) |>
  ungroup() |>
  mutate(source_country = factor(source_country, levels = top4)) |>
  na.omit()
```

::: panel-tabset

## View

```{r}
#| label: multiple-calendar-heatmap-view
#| echo: false
#| fig-width: 10
#| fig-height: 8

ggplot(
  top4_attacks,
  aes(hour, wkday, fill = n)
) +
  geom_tile(color = "white", linewidth = 0.1) +
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_gradient(
    name = "# of attacks",
    low = "sky blue",
    high = "dark blue",
    labels = comma
  ) +
  facet_wrap(~source_country, ncol = 2) +
  labs(
    x = NULL,
    y = NULL,
    title = "Attacks on Top 4 Countries by Weekday and Time of Day"
  ) +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_text(size = 7),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6)
  )
```

## Code

```{r}
#| label: multiple-calendar-heatmap-code
#| eval: false

ggplot(top4_attacks, aes(hour, wkday, fill = n)) +
  geom_tile(color = "white", linewidth = 0.1) +
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_gradient(
    name = "# of attacks",
    low = "sky blue",
    high = "dark blue"
  ) +
  facet_wrap(~source_country, ncol = 2) +
  labs(
    x = NULL,
    y = NULL,
    title = "Attacks on Top 4 Countries by Weekday and Time of Day"
  )
```

:::

## 17.5 Plotting Cycle Plot

This section uses `arrivals_by_air.xlsx` to visualise visitor arrivals from Vietnam by month and year.

### 17.5.1 Step 1: Data Import

```{r}
#| label: import-air

air <- read_excel(find_data_file("arrivals_by_air.xlsx"))

kable(head(air[, 1:6]))
```

### 17.5.2 Step 2: Deriving Month and Year Fields

```{r}
#| label: derive-air-date-fields

air <- air |>
  mutate(
    month = factor(
      month(`Month-Year`),
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    ),
    year = year(ymd(`Month-Year`))
  )
```

### 17.5.3 Step 4: Extracting the Target Country

```{r}
#| label: extract-vietnam

Vietnam <- air |>
  select(Vietnam, month, year) |>
  filter(year >= 2010)

kable(head(Vietnam))
```

### 17.5.4 Step 5: Computing Year Average Arrivals by Month

```{r}
#| label: vietnam-month-average

hline.data <- Vietnam |>
  group_by(month) |>
  summarise(avgvalue = mean(Vietnam), .groups = "drop")

kable(hline.data)
```

### 17.5.5 Step 6: Plotting the Cycle Plot

::: panel-tabset

## View

```{r}
#| label: cycle-plot-view
#| echo: false
#| fig-width: 12
#| fig-height: 5.8

ggplot() +
  geom_line(
    data = Vietnam,
    aes(x = year, y = Vietnam, group = month),
    colour = my_dark,
    linewidth = 0.45
  ) +
  geom_hline(
    aes(yintercept = avgvalue),
    data = hline.data,
    linetype = 6,
    colour = my_red,
    linewidth = 0.5
  ) +
  facet_grid(~month) +
  labs(
    title = "Visitor Arrivals from Vietnam by Air, Jan 2010-Dec 2019",
    x = NULL,
    y = "No. of Visitors"
  ) +
  scale_y_continuous(labels = comma) +
  theme_tufte(base_family = "Helvetica") +
  theme(
    axis.text.x = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

## Code

```{r}
#| label: cycle-plot-code
#| eval: false

ggplot() +
  geom_line(
    data = Vietnam,
    aes(x = year, y = Vietnam, group = month),
    colour = "black"
  ) +
  geom_hline(
    aes(yintercept = avgvalue),
    data = hline.data,
    linetype = 6,
    colour = "red",
    linewidth = 0.5
  ) +
  facet_grid(~month) +
  labs(
    title = "Visitor arrivals from Vietnam by air, Jan 2010-Dec 2019",
    x = "",
    y = "No. of Visitors"
  ) +
  theme_tufte(base_family = "Helvetica")
```

:::

## 17.6 Plotting Slopegraph

Chapter 17 uses `newggslopegraph()` from `CGPfunctions`. Because `CGPfunctions` is not currently installable cleanly in this R setup, this section creates the same type of two-year comparison using `ggplot2`, which is already available and keeps the exercise runnable.

### 17.6.1 Step 1: Data Import

```{r}
#| label: import-rice

rice <- read_csv(find_data_file("rice.csv"), show_col_types = FALSE)

kable(head(rice))
```

### 17.6.2 Step 2: Plotting the Slopegraph

```{r}
#| label: prepare-rice-slope

rice_slope <- rice |>
  mutate(Year = factor(Year)) |>
  filter(Year %in% c("1961", "1980")) |>
  group_by(Country) |>
  filter(n_distinct(Year) == 2) |>
  ungroup()
```

::: panel-tabset

## View

```{r}
#| label: slopegraph-view
#| echo: false
#| fig-width: 10
#| fig-height: 7

left_labels <- rice_slope |>
  filter(Year == "1961")

right_labels <- rice_slope |>
  filter(Year == "1980")

ggplot(
  rice_slope,
  aes(x = Year, y = Yield, group = Country, colour = Country)
) +
  geom_line(linewidth = 0.8, alpha = 0.85) +
  geom_point(size = 2.4) +
  geom_text(
    data = left_labels,
    aes(label = Country),
    hjust = 1.12,
    size = 3.2,
    show.legend = FALSE
  ) +
  geom_text(
    data = right_labels,
    aes(label = Country),
    hjust = -0.12,
    size = 3.2,
    show.legend = FALSE
  ) +
  scale_y_continuous(labels = comma) +
  scale_x_discrete(expand = expansion(mult = 0.22)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Rice Yield of Top 11 Asian Countries",
    subtitle = "1961-1980",
    caption = "Prepared by: Yang Yang",
    x = NULL,
    y = "Yield"
  ) +
  theme_yy +
  theme(
    legend.position = "none",
    plot.margin = margin(10, 60, 10, 60)
  )
```

## Code

```{r}
#| label: slopegraph-code
#| eval: false

rice_slope <- rice |>
  mutate(Year = factor(Year)) |>
  filter(Year %in% c("1961", "1980")) |>
  group_by(Country) |>
  filter(n_distinct(Year) == 2) |>
  ungroup()

ggplot(
  rice_slope,
  aes(x = Year, y = Yield, group = Country, colour = Country)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.4) +
  geom_text(
    data = filter(rice_slope, Year == "1961"),
    aes(label = Country),
    hjust = 1.12,
    show.legend = FALSE
  ) +
  geom_text(
    data = filter(rice_slope, Year == "1980"),
    aes(label = Country),
    hjust = -0.12,
    show.legend = FALSE
  ) +
  coord_cartesian(clip = "off") +
  theme_yy +
  theme(legend.position = "none")
```

:::

## Key Takeaways

::: {.callout-tip}
Time-oriented visualisation works best when the date-time fields are converted into meaningful analysis fields such as weekday, hour, month, and year. Once these fields are derived, `ggplot2` can be used to build calendar heatmaps, cycle plots, and slopegraphs.
:::

In this exercise:

- calendar heatmaps reveal attack intensity by weekday and hour;
- faceted calendar heatmaps compare temporal attack patterns across countries;
- cycle plots compare monthly seasonal patterns across years; and
- slopegraphs show how country-level rice yield changed between 1961 and 1980.

## References

- [R4VA Chapter 17: Visualising and Analysing Time-oriented Data](https://r4va.netlify.app/chap17)
- [CRAN: CGPfunctions package status](https://cran.r-project.org/package=CGPfunctions)
- [CRAN Archive: CGPfunctions](https://cran.r-project.org/src/contrib/Archive/CGPfunctions/)

::: vaa-exercise-footer
### Portfolio note

**Key takeaway.** Time-oriented graphics organise event patterns so change can be read as a sequence.

**Practical use.** Use it when timing, order, seasonality, or duration affects the analytical question.

::: vaa-footer-nav
[Back to Gallery](../../hands-on-gallery.html) [Previous: Exercise 7b](../Hands-on_Ex07b/Hands-on_Ex07b.html) [Next: Exercise 9a](../Hands-on_Ex09a/Hands-on_Ex09a.html)
:::
:::

