---
title: "Hands-on Exercise 2 - Beyond ggplot2 Fundamentals"
author: "Yang Yang"
---

## 2.1 Overview

This chapter introduces several ggplot2 extensions that enhance the clarity and visual quality of statistical graphics. By the end of this exercise, you will be able to:

-   Manage annotation placement on graphs using the **ggrepel** package,

-   Design publication-ready figures with the **ggthemes** and **hrbrthemes** packages,

-   Combine multiple ggplot2 graphs into a single composite figure using the **patchwork** package.

## 2.2 Getting started

### 2.2.1 Installing and loading the required libraries

In this exercise, beside tidyverse, four R packages will be used. They are:

-   ggrepel: an R package provides geoms for ggplot2 to repel overlapping text labels.

-   ggthemes: an R package provides some extra themes, geoms, and scales for ‘ggplot2’.

-   hrbrthemes: an R package provides typography-centric themes and theme components for ggplot2.

-   patchwork: an R package for preparing composite figure created using ggplot2.

The code chunk below checks whether the required packages are already installed and loads them into your working R environment.

```{r}
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse) 
```

### 2.2.2 Importing data

For the purpose of this exercise, a data file called *Exam_data* will be used. It consists of year end examination grades of a cohort of primary 3 students from a local school. It is in csv file format.

The code chunk below imports *exam_data.csv* into R environment by using [*read_csv()*](https://readr.tidyverse.org/reference/read_delim.html) function of [**readr**](https://readr.tidyverse.org/) package. **readr** is one of the tidyverse package.

```{r}
exam_data <- read_csv("Exam_data.csv")
```

The **exam_data** tibble contains seven attributes in total — four categorical and three continuous.

-   The categorical attributes are: **ID, CLASS, GENDER** and **RACE**.

-   The continuous attributes are: **MATHS, ENGLISH** and **SCIENCE**.

## 2.2.3 Understanding the Data

Before diving into the visualisations, it is good practice to get a quick overview of the dataset.

::: panel-tabset
## Structure

```{r}
glimpse(exam_data)
```

## Maths

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 fill = "steelblue",
                 color = "white") +
  labs(title = "Distribution of Maths Scores",
       x = "Maths Score",
       y = "Count") +
  theme_minimal()
```

## English

```{r}
ggplot(data = exam_data,
       aes(x = ENGLISH)) +
  geom_histogram(bins = 20,
                 fill = "salmon",
                 color = "white") +
  labs(title = "Distribution of English Scores",
       x = "English Score",
       y = "Count") +
  theme_minimal()
```

## Science

```{r}
ggplot(data = exam_data,
       aes(x = SCIENCE)) +
  geom_histogram(bins = 20,
                 fill = "seagreen",
                 color = "white") +
  labs(title = "Distribution of Science Scores",
       x = "Science Score",
       y = "Count") +
  theme_minimal()
```
:::

This keeps it short and purposeful — `glimpse()` gives a structural overview, and the histogram gives an immediate sense of the score distribution before any advanced plotting begins. You can swap `MATHS` for `ENGLISH` or `SCIENCE` to explore the other subjects too.
