---
title: "Hands-on Exercise 5c - Heatmap for Visualising and Analysing Multivariate Data"
author: "Yang Yang"
format:
  html:
    theme: sketchy
    toc: true
    toc-depth: 3
    number-sections: false
    code-copy: true
    code-overflow: wrap
editor: visual
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

if (!require(pacman)) {
  install.packages("pacman")
}

pacman::p_load(
  tidyverse,
  heatmaply,
  dendextend,
  seriation,
  DT,
  janitor,
  scales,
  cluster,
  plotly,
  knitr
)

my_teal <- "#2A6F6F"
my_soft_teal <- "#8AA6A3"
my_dark <- "#1F1F1F"
my_light_teal <- "#D8E8E6"
my_grey <- "#F3F4F4"
my_warning <- "#B07C3E"
my_red <- "#9A4B4B"

heat_cols <- colorRampPalette(
  c(my_light_teal, my_soft_teal, my_teal, my_dark)
)(256)

theme_yy <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", colour = my_dark),
    plot.subtitle = element_text(colour = my_teal),
    axis.title = element_text(colour = my_dark),
    axis.text = element_text(colour = my_dark),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## 5.1 Overview

Heatmaps visualise data through colour intensity. When applied to a table, heatmaps are useful for comparing many observations across many variables at the same time.

In this hands-on exercise, we use the World Happiness 2018 dataset. Each row represents a country, while the columns represent happiness-related indicators such as happiness score, GDP per capita, social support, healthy life expectancy, freedom, generosity, and perceptions of corruption.

By the end of this exercise, we will be able to:

- import and prepare World Happiness 2018 data,
- transform a data frame into a matrix for heatmap analysis,
- plot static heatmaps using base R,
- create interactive heatmaps using `heatmaply`,
- apply scaling, normalising, and percentising transformations,
- compare clustering and seriation methods,
- add small extensions using regional averages and correlation heatmaps.

::: {.callout-important}
## Important

Heatmaps are most useful when the variables are comparable.  
If variables have very different ranges, transformation is needed before interpretation.
:::

## 5.2 Installing and Launching R Packages

The main packages used in this exercise are `heatmaply`, `seriation`, `dendextend`, and `tidyverse`.

::: {.panel-tabset}

## View

```{r}
#| echo: false

package_table <- tibble(
  Package = c("tidyverse", "heatmaply", "dendextend", "seriation", "DT", "cluster"),
  Purpose = c(
    "Data import, wrangling, and plotting",
    "Interactive heatmap and clustering heatmap",
    "Dendrogram and clustering helper functions",
    "Optimal row and column ordering",
    "Interactive HTML tables",
    "Silhouette analysis for cluster checking"
  )
)

DT::datatable(
  package_table,
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t", pageLength = 6),
  width = "100%"
)
```

## Code

```{r}
#| eval: false

pacman::p_load(
  tidyverse,
  heatmaply,
  dendextend,
  seriation,
  DT,
  cluster
)
```

:::

## 5.3 Importing and Preparing the Data Set

The dataset used in this exercise is `WHData-2018.csv`.

### 5.3.1 Importing the data set

::: {.panel-tabset}

## View

```{r}
#| echo: false

csv_file <- list.files(pattern = "WHData-2018.*\\.csv$")[1]

wh <- read_csv(csv_file, show_col_types = FALSE)

glimpse(wh)
```

## Code

```{r}
csv_file <- list.files(pattern = "WHData-2018.*\\.csv$")[1]

wh <- read_csv(csv_file, show_col_types = FALSE)

glimpse(wh)
```

:::

Using `list.files()` makes the code safer because it can still read a file name such as `WHData-2018(3).csv`.

### 5.3.2 Preparing the data

For heatmap analysis, the country name should become the row name. The numerical variables are then selected for analysis.

::: {.panel-tabset}

## View

```{r}
#| echo: false

heatmap_vars <- c(
  "Happiness score",
  "Dystopia",
  "GDP per capita",
  "Social support",
  "Healthy life expectancy",
  "Freedom to make life choices",
  "Generosity",
  "Perceptions of corruption"
)

wh_selected <- wh %>%
  select(
    Country,
    Region,
    all_of(heatmap_vars)
  ) %>%
  drop_na()

DT::datatable(
  head(wh_selected, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
heatmap_vars <- c(
  "Happiness score",
  "Dystopia",
  "GDP per capita",
  "Social support",
  "Healthy life expectancy",
  "Freedom to make life choices",
  "Generosity",
  "Perceptions of corruption"
)

wh_selected <- wh %>%
  select(
    Country,
    Region,
    all_of(heatmap_vars)
  ) %>%
  drop_na()
```

:::

### 5.3.3 Transforming the data frame into a matrix

A heatmap needs a numerical matrix. Therefore, the selected numerical columns are converted into matrix format.

::: {.panel-tabset}

## View

```{r}
#| echo: false

wh_matrix_df <- wh_selected %>%
  select(all_of(heatmap_vars)) %>%
  as.data.frame()

row.names(wh_matrix_df) <- make.unique(wh_selected$Country)

wh_matrix <- as.matrix(wh_matrix_df)

class(wh_matrix)
dim(wh_matrix)
```

## Code

```{r}
wh_matrix_df <- wh_selected %>%
  select(all_of(heatmap_vars)) %>%
  as.data.frame()

row.names(wh_matrix_df) <- make.unique(wh_selected$Country)

wh_matrix <- as.matrix(wh_matrix_df)

class(wh_matrix)
dim(wh_matrix)
```

:::

The matrix contains countries as rows and happiness-related indicators as columns.

## 5.4 Static Heatmap

Static heatmaps can be created using the base R `heatmap()` function.

### 5.4.1 heatmap() without clustering

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 9

heatmap(
  wh_matrix,
  Rowv = NA,
  Colv = NA,
  col = heat_cols,
  scale = "none",
  margins = c(10, 4),
  cexRow = 0.45,
  cexCol = 0.8,
  main = "Static heatmap without clustering"
)
```

## Code

```{r}
heatmap(
  wh_matrix,
  Rowv = NA,
  Colv = NA,
  col = heat_cols,
  scale = "none",
  margins = c(10, 4),
  cexRow = 0.45,
  cexCol = 0.8,
  main = "Static heatmap without clustering"
)
```

:::

This heatmap follows the original row and column order. However, because the variables have different ranges, the larger-scale variables dominate the colour pattern.

### 5.4.2 Cluster heatmap

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 9

heatmap(
  wh_matrix,
  col = heat_cols,
  scale = "none",
  margins = c(10, 4),
  cexRow = 0.45,
  cexCol = 0.8,
  main = "Cluster heatmap of World Happiness indicators"
)
```

## Code

```{r}
heatmap(
  wh_matrix,
  col = heat_cols,
  scale = "none",
  margins = c(10, 4),
  cexRow = 0.45,
  cexCol = 0.8,
  main = "Cluster heatmap of World Happiness indicators"
)
```

:::

By default, `heatmap()` performs hierarchical clustering. The row and column order changes because similar countries and similar variables are placed closer together.

### 5.4.3 Column-scaled heatmap

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 9

heatmap(
  wh_matrix,
  col = heat_cols,
  scale = "column",
  margins = c(10, 4),
  cexRow = 0.45,
  cexCol = 0.8,
  main = "Column-scaled heatmap of World Happiness indicators"
)
```

## Code

```{r}
heatmap(
  wh_matrix,
  col = heat_cols,
  scale = "column",
  margins = c(10, 4),
  cexRow = 0.45,
  cexCol = 0.8,
  main = "Column-scaled heatmap of World Happiness indicators"
)
```

:::

Column scaling makes each variable more comparable by standardising values within each column.

::: {.callout-tip}
## What can we learn?

Without scaling, variables with larger numerical ranges can dominate the heatmap.  
Column scaling reduces this issue and makes cross-variable comparison more meaningful.
:::

## 5.5 Creating Interactive Heatmap

The `heatmaply` package creates interactive cluster heatmaps that can be embedded in an HTML page.

### 5.5.1 Working with heatmaply

::: {.panel-tabset}

## View

```{r}
#| echo: false

heatmaply::heatmaply(
  wh_matrix,
  colors = heat_cols,
  main = "Interactive heatmap of World Happiness indicators",
  xlab = "Variables",
  ylab = "Countries"
)
```

## Code

```{r}
heatmaply::heatmaply(
  wh_matrix,
  colors = heat_cols,
  main = "Interactive heatmap of World Happiness indicators",
  xlab = "Variables",
  ylab = "Countries"
)
```

:::

Unlike the base heatmap, `heatmaply()` creates an interactive output. Users can hover over cells and inspect the values.

## 5.6 Data Transformation

When the variables have different measurement ranges, transformation is needed before clustering and interpretation.

### 5.6.1 Scaling method

::: {.panel-tabset}

## View

```{r}
#| echo: false

heatmaply::heatmaply(
  wh_matrix,
  scale = "column",
  colors = heat_cols,
  main = "Interactive heatmap with column scaling",
  xlab = "Variables",
  ylab = "Countries"
)
```

## Code

```{r}
heatmaply::heatmaply(
  wh_matrix,
  scale = "column",
  colors = heat_cols,
  main = "Interactive heatmap with column scaling",
  xlab = "Variables",
  ylab = "Countries"
)
```

:::

Scaling subtracts the mean and divides by the standard deviation. It is useful when variables are assumed to be roughly comparable after standardisation.

### 5.6.2 Normalising method

::: {.panel-tabset}

## View

```{r}
#| echo: false

wh_normalised <- heatmaply::normalize(wh_matrix)

heatmaply::heatmaply(
  wh_normalised,
  colors = heat_cols,
  main = "Interactive heatmap with normalised values",
  xlab = "Variables",
  ylab = "Countries"
)
```

## Code

```{r}
wh_normalised <- heatmaply::normalize(wh_matrix)

heatmaply::heatmaply(
  wh_normalised,
  colors = heat_cols,
  main = "Interactive heatmap with normalised values",
  xlab = "Variables",
  ylab = "Countries"
)
```

:::

Normalising brings each variable into a 0 to 1 range. This preserves the relative shape of each variable while making the variables easier to compare.

### 5.6.3 Percentising method

::: {.panel-tabset}

## View

```{r}
#| echo: false

wh_percentised <- heatmaply::percentize(wh_matrix)

heatmaply::heatmaply(
  wh_percentised,
  colors = heat_cols,
  main = "Interactive heatmap with percentised values",
  xlab = "Variables",
  ylab = "Countries"
)
```

## Code

```{r}
wh_percentised <- heatmaply::percentize(wh_matrix)

heatmaply::heatmaply(
  wh_percentised,
  colors = heat_cols,
  main = "Interactive heatmap with percentised values",
  xlab = "Variables",
  ylab = "Countries"
)
```

:::

Percentising converts values into empirical percentiles. This makes each cell easier to interpret as a relative position within its variable.

## 5.7 Clustering Algorithm

Hierarchical clustering groups similar countries and variables together. The result depends on the distance method and clustering method selected.

### 5.7.1 Manual approach

::: {.panel-tabset}

## View

```{r}
#| echo: false

heatmaply::heatmaply(
  wh_normalised,
  dist_method = "euclidean",
  hclust_method = "ward.D",
  colors = heat_cols,
  main = "Manual clustering: Euclidean distance and ward.D method",
  xlab = "Variables",
  ylab = "Countries"
)
```

## Code

```{r}
heatmaply::heatmaply(
  wh_normalised,
  dist_method = "euclidean",
  hclust_method = "ward.D",
  colors = heat_cols,
  main = "Manual clustering: Euclidean distance and ward.D method",
  xlab = "Variables",
  ylab = "Countries"
)
```

:::

The manual approach fixes the distance method and clustering method directly.

### 5.7.2 Statistical approach

Instead of choosing the number of clusters by guessing, we can use average silhouette width as a simple diagnostic.

::: {.panel-tabset}

## View

```{r}
#| echo: false

wh_dist <- dist(wh_normalised, method = "euclidean")
wh_clust <- hclust(wh_dist, method = "average")

silhouette_result <- map_dfr(
  2:8,
  function(k_value) {
    cluster_group <- cutree(wh_clust, k = k_value)
    silhouette_score <- mean(cluster::silhouette(cluster_group, wh_dist)[, 3])
    
    tibble(
      k = k_value,
      average_silhouette_width = silhouette_score
    )
  }
)

best_k <- silhouette_result$k[which.max(silhouette_result$average_silhouette_width)]

DT::datatable(
  silhouette_result %>%
    mutate(average_silhouette_width = round(average_silhouette_width, 3)),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t", pageLength = 7),
  width = "100%"
)
```

## Code

```{r}
wh_dist <- dist(wh_normalised, method = "euclidean")
wh_clust <- hclust(wh_dist, method = "average")

silhouette_result <- map_dfr(
  2:8,
  function(k_value) {
    cluster_group <- cutree(wh_clust, k = k_value)
    silhouette_score <- mean(cluster::silhouette(cluster_group, wh_dist)[, 3])
    
    tibble(
      k = k_value,
      average_silhouette_width = silhouette_score
    )
  }
)

best_k <- silhouette_result$k[which.max(silhouette_result$average_silhouette_width)]
```

:::

### 5.7.3 Plotting the silhouette result

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4

ggplot(
  silhouette_result,
  aes(
    x = k,
    y = average_silhouette_width
  )
) +
  geom_line(colour = my_teal, linewidth = 0.8) +
  geom_point(size = 2.5, colour = my_dark) +
  geom_point(
    data = silhouette_result %>% filter(k == best_k),
    aes(x = k, y = average_silhouette_width),
    colour = my_warning,
    size = 4
  ) +
  scale_x_continuous(breaks = 2:8) +
  labs(
    title = "Estimating the number of clusters",
    subtitle = "Higher average silhouette width suggests a better cluster separation",
    x = "Number of clusters",
    y = "Average silhouette width"
  ) +
  theme_yy
```

## Code

```{r}
ggplot(
  silhouette_result,
  aes(
    x = k,
    y = average_silhouette_width
  )
) +
  geom_line(colour = my_teal, linewidth = 0.8) +
  geom_point(size = 2.5, colour = my_dark) +
  geom_point(
    data = silhouette_result %>% filter(k == best_k),
    aes(x = k, y = average_silhouette_width),
    colour = my_warning,
    size = 4
  ) +
  scale_x_continuous(breaks = 2:8) +
  labs(
    title = "Estimating the number of clusters",
    subtitle = "Higher average silhouette width suggests a better cluster separation",
    x = "Number of clusters",
    y = "Average silhouette width"
  ) +
  theme_yy
```

:::

### 5.7.4 Heatmap with selected number of clusters

::: {.panel-tabset}

## View

```{r}
#| echo: false

heatmaply::heatmaply(
  wh_normalised,
  dist_method = "euclidean",
  hclust_method = "average",
  k_row = best_k,
  colors = heat_cols,
  main = paste("Interactive heatmap with", best_k, "country clusters"),
  xlab = "Variables",
  ylab = "Countries"
)
```

## Code

```{r}
heatmaply::heatmaply(
  wh_normalised,
  dist_method = "euclidean",
  hclust_method = "average",
  k_row = best_k,
  colors = heat_cols,
  main = paste("Interactive heatmap with", best_k, "country clusters"),
  xlab = "Variables",
  ylab = "Countries"
)
```

:::

This version uses the selected number of clusters from the silhouette check.

## 5.8 Seriation

Seriation changes the ordering of rows and columns so that similar items are placed closer together. It can make the heatmap easier to read.

### 5.8.1 Optimal Leaf Ordering

::: {.panel-tabset}

## View

```{r}
#| echo: false

heatmaply::heatmaply(
  wh_normalised,
  seriate = "OLO",
  colors = heat_cols,
  main = "Heatmap using Optimal Leaf Ordering",
  xlab = "Variables",
  ylab = "Countries"
)
```

## Code

```{r}
heatmaply::heatmaply(
  wh_normalised,
  seriate = "OLO",
  colors = heat_cols,
  main = "Heatmap using Optimal Leaf Ordering",
  xlab = "Variables",
  ylab = "Countries"
)
```

:::

`OLO` stands for Optimal Leaf Ordering. It aims to arrange the leaves of the dendrogram in a clearer sequence.

### 5.8.2 Gruvaeus and Wainer ordering

::: {.panel-tabset}

## View

```{r}
#| echo: false

heatmaply::heatmaply(
  wh_normalised,
  seriate = "GW",
  colors = heat_cols,
  main = "Heatmap using Gruvaeus and Wainer ordering",
  xlab = "Variables",
  ylab = "Countries"
)
```

## Code

```{r}
heatmaply::heatmaply(
  wh_normalised,
  seriate = "GW",
  colors = heat_cols,
  main = "Heatmap using Gruvaeus and Wainer ordering",
  xlab = "Variables",
  ylab = "Countries"
)
```

:::

`GW` is a faster heuristic ordering method. It is useful when the dataset is larger and optimal ordering becomes more expensive.

## 5.9 Extra Analysis: Regional Average Heatmap

This small extension summarises the indicators by region. It reduces country-level noise and gives a clearer comparison between regions.

### 5.9.1 Preparing regional averages

::: {.panel-tabset}

## View

```{r}
#| echo: false

region_average <- wh_selected %>%
  group_by(Region) %>%
  summarise(
    across(
      all_of(heatmap_vars),
      ~ mean(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  )

region_matrix_df <- region_average %>%
  select(all_of(heatmap_vars)) %>%
  as.data.frame()

row.names(region_matrix_df) <- make.unique(region_average$Region)

region_matrix <- as.matrix(region_matrix_df)

DT::datatable(
  region_average %>%
    mutate(across(where(is.numeric), ~ round(.x, 3))),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 10, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
region_average <- wh_selected %>%
  group_by(Region) %>%
  summarise(
    across(
      all_of(heatmap_vars),
      ~ mean(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  )

region_matrix_df <- region_average %>%
  select(all_of(heatmap_vars)) %>%
  as.data.frame()

row.names(region_matrix_df) <- make.unique(region_average$Region)

region_matrix <- as.matrix(region_matrix_df)
```

:::

### 5.9.2 Regional average heatmap

::: {.panel-tabset}

## View

```{r}
#| echo: false

heatmaply::heatmaply(
  heatmaply::normalize(region_matrix),
  colors = heat_cols,
  main = "Regional average heatmap of World Happiness indicators",
  xlab = "Variables",
  ylab = "Regions"
)
```

## Code

```{r}
heatmaply::heatmaply(
  heatmaply::normalize(region_matrix),
  colors = heat_cols,
  main = "Regional average heatmap of World Happiness indicators",
  xlab = "Variables",
  ylab = "Regions"
)
```

:::

The regional heatmap is easier to interpret because it has fewer rows than the country-level heatmap.

## 5.10 Extra Analysis: Correlation Heatmap

A heatmap can also be used to visualise correlation between variables.

### 5.10.1 Computing correlation matrix

::: {.panel-tabset}

## View

```{r}
#| echo: false

cor_matrix <- cor(
  wh_matrix,
  use = "pairwise.complete.obs"
)

round(cor_matrix, 3)
```

## Code

```{r}
cor_matrix <- cor(
  wh_matrix,
  use = "pairwise.complete.obs"
)

round(cor_matrix, 3)
```

:::

### 5.10.2 Plotting correlation heatmap

::: {.panel-tabset}

## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6

cor_long <- as.data.frame(as.table(cor_matrix)) %>%
  rename(
    Variable_1 = Var1,
    Variable_2 = Var2,
    Correlation = Freq
  )

ggplot(
  cor_long,
  aes(
    x = Variable_1,
    y = Variable_2,
    fill = Correlation
  )
) +
  geom_tile(colour = "white", linewidth = 0.35) +
  scale_fill_gradient2(
    low = my_red,
    mid = my_light_teal,
    high = my_teal,
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation heatmap of World Happiness indicators",
    subtitle = "This extension shows similarity between variables",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  ) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    panel.grid = element_blank()
  )
```

## Code

```{r}
cor_long <- as.data.frame(as.table(cor_matrix)) %>%
  rename(
    Variable_1 = Var1,
    Variable_2 = Var2,
    Correlation = Freq
  )

ggplot(
  cor_long,
  aes(
    x = Variable_1,
    y = Variable_2,
    fill = Correlation
  )
) +
  geom_tile(colour = "white", linewidth = 0.35) +
  scale_fill_gradient2(
    low = my_red,
    mid = my_light_teal,
    high = my_teal,
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation heatmap of World Happiness indicators",
    subtitle = "This extension shows similarity between variables",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  ) +
  theme_yy +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    panel.grid = element_blank()
  )
```

:::

The correlation heatmap helps identify which indicators move together. It supports the main heatmap analysis by showing variable similarity directly.

## 5.11 Extra Analysis: Top and Bottom Happiness Countries

This extension compares the highest and lowest happiness score countries using a smaller heatmap.

### 5.11.1 Preparing top and bottom countries

::: {.panel-tabset}

## View

```{r}
#| echo: false

top_bottom_heatmap_data <- bind_rows(
  wh_selected %>%
    arrange(desc(`Happiness score`)) %>%
    slice_head(n = 10) %>%
    mutate(Group = "Top 10"),
  wh_selected %>%
    arrange(`Happiness score`) %>%
    slice_head(n = 10) %>%
    mutate(Group = "Bottom 10")
)

top_bottom_matrix_df <- top_bottom_heatmap_data %>%
  select(all_of(heatmap_vars)) %>%
  as.data.frame()

row.names(top_bottom_matrix_df) <- make.unique(
  paste(top_bottom_heatmap_data$Group, top_bottom_heatmap_data$Country, sep = " - ")
)

top_bottom_matrix <- as.matrix(top_bottom_matrix_df)

DT::datatable(
  top_bottom_heatmap_data %>%
    select(Group, Country, Region, all_of(heatmap_vars)),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 20, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
top_bottom_heatmap_data <- bind_rows(
  wh_selected %>%
    arrange(desc(`Happiness score`)) %>%
    slice_head(n = 10) %>%
    mutate(Group = "Top 10"),
  wh_selected %>%
    arrange(`Happiness score`) %>%
    slice_head(n = 10) %>%
    mutate(Group = "Bottom 10")
)

top_bottom_matrix_df <- top_bottom_heatmap_data %>%
  select(all_of(heatmap_vars)) %>%
  as.data.frame()

row.names(top_bottom_matrix_df) <- make.unique(
  paste(top_bottom_heatmap_data$Group, top_bottom_heatmap_data$Country, sep = " - ")
)

top_bottom_matrix <- as.matrix(top_bottom_matrix_df)
```

:::

### 5.11.2 Heatmap of top and bottom countries

::: {.panel-tabset}

## View

```{r}
#| echo: false

heatmaply::heatmaply(
  heatmaply::normalize(top_bottom_matrix),
  colors = heat_cols,
  main = "Heatmap comparison of top and bottom happiness countries",
  xlab = "Variables",
  ylab = "Countries"
)
```

## Code

```{r}
heatmaply::heatmaply(
  heatmaply::normalize(top_bottom_matrix),
  colors = heat_cols,
  main = "Heatmap comparison of top and bottom happiness countries",
  xlab = "Variables",
  ylab = "Countries"
)
```

:::

This smaller heatmap is easier to read than the full country-level heatmap. It highlights how the strongest and weakest happiness-score countries differ across multiple indicators.

## 5.12 Final Takeaway

Heatmaps are useful for exploring multivariate data because they show many observations and variables in one compact view. However, interpretation depends heavily on transformation, clustering, and row-column ordering.

In this exercise, static heatmaps provide the basic structure, while `heatmaply` adds interactivity. Scaling, normalising, and percentising help make variables comparable. Clustering and seriation improve pattern detection.

::: {.callout-tip}
## Final takeaway

Use heatmaps when the main task is to compare many observations across many variables.  
For World Happiness data, heatmaps help reveal country clusters, regional profiles, and relationships between happiness-related indicators.
:::

## 5.13 References

This exercise covered:

- static heatmaps using base R `heatmap()`,
- interactive heatmaps using `heatmaply`,
- column scaling,
- normalising,
- percentising,
- manual and silhouette-based clustering,
- seriation using `OLO` and `GW`,
- regional average heatmaps,
- correlation heatmaps,
- top-bottom country comparison heatmaps.

::: vaa-exercise-footer
### Portfolio note

**Key takeaway.** Heatmaps compress multivariate structure into readable colour patterns.

**Practical use.** Use it when the goal is to compare many values, clusters, or high-low contrasts at once.

::: vaa-footer-nav
[Back to Gallery](../../hands-on-gallery.html) [Previous: Exercise 5b](../Hands-on_Ex05b/Hands-on_Ex05b.html) [Next: Exercise 5d](../Hands-on_Ex05d/Hands-on_Ex05d.html)
:::
:::

