---
title: "Hands-on Exercise 5e - Treemap Visualisation with R"
author: "Yang Yang"
format:
  html:
    theme: sketchy
    toc: true
    toc-depth: 3
    number-sections: false
    code-copy: true
    code-overflow: wrap
editor: visual
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

if (!require(pacman)) {
  install.packages("pacman")
}

pacman::p_load(
  tidyverse,
  treemap,
  treemapify,
  DT,
  janitor,
  lubridate,
  plotly,
  scales,
  knitr
)

my_teal <- "#2A6F6F"
my_soft_teal <- "#8AA6A3"
my_dark <- "#1F1F1F"
my_light_teal <- "#D8E8E6"
my_grey <- "#F3F4F4"
my_warning <- "#B07C3E"
my_red <- "#9A4B4B"

theme_yy <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", colour = my_dark),
    plot.subtitle = element_text(colour = my_teal),
    axis.title = element_text(colour = my_dark),
    axis.text = element_text(colour = my_dark),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## 5.1 Overview

Treemap visualisation is useful when we need to display hierarchical data. Instead of showing values in a normal bar chart, a treemap uses rectangles. The size of each rectangle represents one quantitative variable, while the colour can represent another variable.

In this hands-on exercise, we use `realis2018.csv`, which contains Singapore private residential property transaction records in 2018. The hierarchy used in this exercise is:

-   Planning Region
-   Planning Area
-   Project Name

By the end of this exercise, we will be able to:

-   import and clean private residential transaction data,
-   summarise transaction records into a treemap-ready structure,
-   design static treemaps using the `treemap` package,
-   compare different colour settings in treemap design,
-   build treemaps using `treemapify`,
-   create a simple interactive treemap using `plotly`,
-   add a small extension to compare resale condominium activity by planning area.

::: callout-important
## Important

A treemap works best when the data has a clear hierarchy.\
For this exercise, the hierarchy is geographic: Planning Region → Planning Area → Project Name.
:::

## 5.2 Installing and Launching R Packages

Before creating the treemap, the required R packages need to be loaded. The main packages are `treemap`, `treemapify`, and `tidyverse`.

::: panel-tabset
## View

```{r}
#| echo: false

package_table <- tibble(
  Package = c("tidyverse", "treemap", "treemapify", "DT", "janitor", "plotly"),
  Purpose = c(
    "Data import, wrangling, and plotting",
    "Static treemap visualisation",
    "Treemap visualisation using ggplot2 grammar",
    "Interactive HTML tables",
    "Cleaning column names",
    "Interactive treemap visualisation"
  )
)

DT::datatable(
  package_table,
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(dom = "t", pageLength = 6),
  width = "100%"
)
```

## Code

```{r}
#| eval: false

pacman::p_load(
  tidyverse,
  treemap,
  treemapify,
  DT,
  janitor,
  plotly
)
```
:::

::: callout-note
## Package note

This version uses `plotly` for the interactive treemap instead of `d3treeR`.\
`d3treeR` is not always stable because it is installed from GitHub, not directly from CRAN.
:::

## 5.3 Data Wrangling

The raw transaction data is too detailed for a treemap. Each row is one transaction record, but a treemap needs summarised values by hierarchy.

### 5.3.1 Importing the data set

::: panel-tabset
## View

```{r}
#| echo: false

csv_file <- list.files(pattern = "realis2018.*\\.csv$")[1]

realis_raw <- read_csv(csv_file, show_col_types = FALSE)

glimpse(realis_raw)
```

## Code

```{r}
#| eval: false

csv_file <- list.files(pattern = "realis2018.*\\.csv$")[1]

realis_raw <- read_csv(csv_file, show_col_types = FALSE)

glimpse(realis_raw)
```
:::

The data is imported from the current exercise folder. Using `list.files()` avoids path errors when the website is rendered.

### 5.3.2 Cleaning the data

::: panel-tabset
## View

```{r}
#| echo: false

realis2018 <- realis_raw %>%
  janitor::clean_names() %>%
  mutate(
    project_name = as.character(project_name),
    planning_region = as.character(planning_region),
    planning_area = as.character(planning_area),
    property_type = as.character(property_type),
    type_of_sale = as.character(type_of_sale),
    no_of_units = as.numeric(no_of_units),
    area_sqm = as.numeric(area_sqm),
    unit_price_psm = as.numeric(unit_price_psm),
    transacted_price = as.numeric(transacted_price),
    sale_date = lubridate::dmy(sale_date)
  )

DT::datatable(
  head(realis2018, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
realis2018 <- realis_raw %>%
  janitor::clean_names() %>%
  mutate(
    project_name = as.character(project_name),
    planning_region = as.character(planning_region),
    planning_area = as.character(planning_area),
    property_type = as.character(property_type),
    type_of_sale = as.character(type_of_sale),
    no_of_units = as.numeric(no_of_units),
    area_sqm = as.numeric(area_sqm),
    unit_price_psm = as.numeric(unit_price_psm),
    transacted_price = as.numeric(transacted_price),
    sale_date = lubridate::dmy(sale_date)
  )

DT::datatable(
  head(realis2018, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```
:::

Column names are cleaned into a consistent format. For example, `Project Name` becomes `project_name`, and `Unit Price ($ psm)` becomes `unit_price_psm`.

### 5.3.3 Grouped summaries without the pipe

The basic idea is to group the transaction records by project and location, then calculate summary measures.

::: panel-tabset
## View

```{r}
#| echo: false

realis2018_grouped <- group_by(
  realis2018,
  project_name,
  planning_region,
  planning_area,
  property_type,
  type_of_sale
)

realis2018_summarised_no_pipe <- summarise(
  realis2018_grouped,
  total_unit_sold = sum(no_of_units, na.rm = TRUE),
  total_area = sum(area_sqm, na.rm = TRUE),
  median_unit_price_psm = median(unit_price_psm, na.rm = TRUE),
  median_transacted_price = median(transacted_price, na.rm = TRUE),
  .groups = "drop"
)

DT::datatable(
  head(realis2018_summarised_no_pipe, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
realis2018_grouped <- group_by(
  realis2018,
  project_name,
  planning_region,
  planning_area,
  property_type,
  type_of_sale
)

realis2018_summarised_no_pipe <- summarise(
  realis2018_grouped,
  total_unit_sold = sum(no_of_units, na.rm = TRUE),
  total_area = sum(area_sqm, na.rm = TRUE),
  median_unit_price_psm = median(unit_price_psm, na.rm = TRUE),
  median_transacted_price = median(transacted_price, na.rm = TRUE),
  .groups = "drop"
)

DT::datatable(
  head(realis2018_summarised_no_pipe, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```
:::

This approach works, but it creates an intermediate grouped object. The pipe version is cleaner.

### 5.3.4 Grouped summaries with the pipe

::: panel-tabset
## View

```{r}
#| echo: false

realis2018_summarised <- realis2018 %>%
  group_by(
    project_name,
    planning_region,
    planning_area,
    property_type,
    type_of_sale
  ) %>%
  summarise(
    total_unit_sold = sum(no_of_units, na.rm = TRUE),
    total_area = sum(area_sqm, na.rm = TRUE),
    median_unit_price_psm = median(unit_price_psm, na.rm = TRUE),
    median_transacted_price = median(transacted_price, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(
    total_unit_sold > 0,
    !is.na(median_unit_price_psm),
    !is.na(planning_region),
    !is.na(planning_area),
    !is.na(project_name)
  )

DT::datatable(
  head(realis2018_summarised, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
realis2018_summarised <- realis2018 %>%
  group_by(
    project_name,
    planning_region,
    planning_area,
    property_type,
    type_of_sale
  ) %>%
  summarise(
    total_unit_sold = sum(no_of_units, na.rm = TRUE),
    total_area = sum(area_sqm, na.rm = TRUE),
    median_unit_price_psm = median(unit_price_psm, na.rm = TRUE),
    median_transacted_price = median(transacted_price, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(
    total_unit_sold > 0,
    !is.na(median_unit_price_psm),
    !is.na(planning_region),
    !is.na(planning_area),
    !is.na(project_name)
  )

DT::datatable(
  head(realis2018_summarised, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```
:::

::: callout-tip
## What can we learn?

The pipe version is easier to read because each operation follows the previous one.\
This is useful for treemap preparation because the data needs to be grouped, summarised, filtered, and checked.
:::

## 5.4 Designing Treemap with treemap Package

The `treemap` package is designed specifically for treemap visualisation. The key arguments are:

-   `index`: hierarchy of rectangles,
-   `vSize`: variable used for rectangle size,
-   `vColor`: variable used for rectangle colour,
-   `type`: how colour values should be interpreted.

### 5.4.1 Selecting resale condominium records

::: panel-tabset
## View

```{r}
#| echo: false

realis2018_selected <- realis2018_summarised %>%
  filter(
    property_type == "Condominium",
    type_of_sale == "Resale"
  )

DT::datatable(
  head(realis2018_selected, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
realis2018_selected <- realis2018_summarised %>%
  filter(
    property_type == "Condominium",
    type_of_sale == "Resale"
  )

DT::datatable(
  head(realis2018_selected, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```
:::

The selected data focuses only on resale condominium transactions. This keeps the treemap closer to one property market segment.

### 5.4.2 Using basic treemap arguments

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 7

treemap(
  realis2018_selected,
  index = c("planning_region", "planning_area", "project_name"),
  vSize = "total_unit_sold",
  vColor = "median_unit_price_psm",
  title = "Resale Condominium by Planning Region and Area, 2018",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```

## Code

```{r}
treemap(
  realis2018_selected,
  index = c("planning_region", "planning_area", "project_name"),
  vSize = "total_unit_sold",
  vColor = "median_unit_price_psm",
  title = "Resale Condominium by Planning Region and Area, 2018",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```
:::

This first treemap uses the hierarchy and two quantitative variables, but the colour setting is not yet ideal.

::: callout-note
## Reading the treemap

The larger rectangles represent projects with more units sold.\
The colour is controlled by median unit price, but the default colour behaviour needs to be adjusted.
:::

### 5.4.3 Working with vColor and type arguments

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 7

treemap(
  realis2018_selected,
  index = c("planning_region", "planning_area", "project_name"),
  vSize = "total_unit_sold",
  vColor = "median_unit_price_psm",
  type = "value",
  palette = "Greens",
  title = "Resale Condominium by Planning Region and Area, 2018",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```

## Code

```{r}
treemap(
  realis2018_selected,
  index = c("planning_region", "planning_area", "project_name"),
  vSize = "total_unit_sold",
  vColor = "median_unit_price_psm",
  type = "value",
  palette = "Greens",
  title = "Resale Condominium by Planning Region and Area, 2018",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```
:::

With `type = "value"`, the colour intensity reflects median unit price more directly.

## 5.5 Colours in treemap Package

Colour choice matters because a poor palette can make the treemap difficult to interpret. A diverging palette is more suitable when values have both negative and positive sides. For positive values such as unit price, a single-direction palette is easier to read.

### 5.5.1 The value type treemap

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 7

treemap(
  realis2018_selected,
  index = c("planning_region", "planning_area", "project_name"),
  vSize = "total_unit_sold",
  vColor = "median_unit_price_psm",
  type = "value",
  palette = "RdYlBu",
  title = "Value Type Treemap: Resale Condominium, 2018",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```

## Code

```{r}
treemap(
  realis2018_selected,
  index = c("planning_region", "planning_area", "project_name"),
  vSize = "total_unit_sold",
  vColor = "median_unit_price_psm",
  type = "value",
  palette = "RdYlBu",
  title = "Value Type Treemap: Resale Condominium, 2018",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```
:::

The value type treats the colour scale as a value-based mapping. Since unit prices are positive, the diverging palette may not fully use both ends of the colour range.

### 5.5.2 The manual type treemap

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 7

treemap(
  realis2018_selected,
  index = c("planning_region", "planning_area", "project_name"),
  vSize = "total_unit_sold",
  vColor = "median_unit_price_psm",
  type = "manual",
  palette = "Blues",
  title = "Manual Type Treemap: Resale Condominium, 2018",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```

## Code

```{r}
treemap(
  realis2018_selected,
  index = c("planning_region", "planning_area", "project_name"),
  vSize = "total_unit_sold",
  vColor = "median_unit_price_psm",
  type = "manual",
  palette = "Blues",
  title = "Manual Type Treemap: Resale Condominium, 2018",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```
:::

The manual type maps the value range more directly to the colour palette. A single-colour palette is cleaner here because price is a positive continuous variable.

### 5.5.3 Small extension: using a custom teal ggplot version

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 6

ggplot(
  realis2018_selected,
  aes(
    area = total_unit_sold,
    fill = median_unit_price_psm,
    label = project_name
  )
) +
  geom_treemap(colour = "white", linewidth = 0.25) +
  geom_treemap_text(
    colour = my_dark,
    place = "centre",
    grow = FALSE,
    reflow = TRUE,
    min.size = 3
  ) +
  scale_fill_gradient(
    low = my_light_teal,
    high = my_teal,
    labels = comma
  ) +
  labs(
    title = "Custom teal treemap of resale condominium projects",
    subtitle = "Rectangle size shows total units sold; colour shows median unit price",
    fill = "Median unit price\n(S$ psm)"
  ) +
  theme_yy
```

## Code

```{r}
ggplot(
  realis2018_selected,
  aes(
    area = total_unit_sold,
    fill = median_unit_price_psm,
    label = project_name
  )
) +
  geom_treemap(colour = "white", linewidth = 0.25) +
  geom_treemap_text(
    colour = my_dark,
    place = "centre",
    grow = FALSE,
    reflow = TRUE,
    min.size = 3
  ) +
  scale_fill_gradient(
    low = my_light_teal,
    high = my_teal,
    labels = comma
  ) +
  labs(
    title = "Custom teal treemap of resale condominium projects",
    subtitle = "Rectangle size shows total units sold; colour shows median unit price",
    fill = "Median unit price\n(S$ psm)"
  ) +
  theme_yy
```
:::

This extension uses the same treemap idea but applies the customised teal colour style.

## 5.6 Treemap Layout

The layout algorithm affects how rectangles are arranged. The `treemap` package supports different algorithms, including `squarified` and `pivotSize`.

### 5.6.1 Working with algorithm argument

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 7

treemap(
  realis2018_selected,
  index = c("planning_region", "planning_area", "project_name"),
  vSize = "total_unit_sold",
  vColor = "median_unit_price_psm",
  type = "manual",
  palette = "Blues",
  algorithm = "squarified",
  title = "Squarified Treemap: Resale Condominium, 2018",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```

## Code

```{r}
treemap(
  realis2018_selected,
  index = c("planning_region", "planning_area", "project_name"),
  vSize = "total_unit_sold",
  vColor = "median_unit_price_psm",
  type = "manual",
  palette = "Blues",
  algorithm = "squarified",
  title = "Squarified Treemap: Resale Condominium, 2018",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```
:::

The squarified layout tries to make rectangles closer to square shapes, which can improve readability.

### 5.6.2 Using sortID

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 7

treemap(
  realis2018_selected,
  index = c("planning_region", "planning_area", "project_name"),
  vSize = "total_unit_sold",
  vColor = "median_unit_price_psm",
  type = "manual",
  palette = "Blues",
  algorithm = "pivotSize",
  sortID = "median_transacted_price",
  title = "Treemap Sorted by Median Transacted Price, 2018",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```

## Code

```{r}
treemap(
  realis2018_selected,
  index = c("planning_region", "planning_area", "project_name"),
  vSize = "total_unit_sold",
  vColor = "median_unit_price_psm",
  type = "manual",
  palette = "Blues",
  algorithm = "pivotSize",
  sortID = "median_transacted_price",
  title = "Treemap Sorted by Median Transacted Price, 2018",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```
:::

The `sortID` argument controls the rectangle ordering when the selected layout supports sorting.

::: callout-tip
## What can we learn?

Treemap design is not only about colour.\
Hierarchy, size variable, colour variable, layout, and sorting all affect how the reader interprets the chart.
:::

## 5.7 Designing Treemap using treemapify Package

The `treemapify` package allows treemaps to be created using the grammar of graphics. This makes it easier to combine treemap geometry with `ggplot2` themes and scales.

### 5.7.1 Designing a basic treemap

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6

ggplot(
  realis2018_selected,
  aes(
    area = total_unit_sold,
    fill = median_unit_price_psm
  )
) +
  geom_treemap(colour = "white", linewidth = 0.25) +
  scale_fill_gradient(
    low = my_light_teal,
    high = my_teal,
    labels = comma
  ) +
  labs(
    title = "Basic treemap using treemapify",
    subtitle = "Area represents total units sold; fill represents median unit price",
    fill = "Median unit price\n(S$ psm)"
  ) +
  theme_yy
```

## Code

```{r}
ggplot(
  realis2018_selected,
  aes(
    area = total_unit_sold,
    fill = median_unit_price_psm
  )
) +
  geom_treemap(colour = "white", linewidth = 0.25) +
  scale_fill_gradient(
    low = my_light_teal,
    high = my_teal,
    labels = comma
  ) +
  labs(
    title = "Basic treemap using treemapify",
    subtitle = "Area represents total units sold; fill represents median unit price",
    fill = "Median unit price\n(S$ psm)"
  ) +
  theme_yy
```
:::

### 5.7.2 Defining hierarchy by Planning Region

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6

ggplot(
  realis2018_selected,
  aes(
    area = total_unit_sold,
    fill = median_unit_price_psm,
    subgroup = planning_region
  )
) +
  geom_treemap(colour = "white", linewidth = 0.25) +
  geom_treemap_subgroup_border(colour = my_dark, linewidth = 1) +
  geom_treemap_subgroup_text(
    colour = my_dark,
    place = "centre",
    grow = FALSE,
    alpha = 0.8
  ) +
  scale_fill_gradient(
    low = my_light_teal,
    high = my_teal,
    labels = comma
  ) +
  labs(
    title = "Treemap grouped by Planning Region",
    subtitle = "Subgroup borders separate the highest geographic level",
    fill = "Median unit price\n(S$ psm)"
  ) +
  theme_yy
```

## Code

```{r}
ggplot(
  realis2018_selected,
  aes(
    area = total_unit_sold,
    fill = median_unit_price_psm,
    subgroup = planning_region
  )
) +
  geom_treemap(colour = "white", linewidth = 0.25) +
  geom_treemap_subgroup_border(colour = my_dark, linewidth = 1) +
  geom_treemap_subgroup_text(
    colour = my_dark,
    place = "centre",
    grow = FALSE,
    alpha = 0.8
  ) +
  scale_fill_gradient(
    low = my_light_teal,
    high = my_teal,
    labels = comma
  ) +
  labs(
    title = "Treemap grouped by Planning Region",
    subtitle = "Subgroup borders separate the highest geographic level",
    fill = "Median unit price\n(S$ psm)"
  ) +
  theme_yy
```
:::

### 5.7.3 Defining hierarchy by Planning Area

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6

ggplot(
  realis2018_selected,
  aes(
    area = total_unit_sold,
    fill = median_unit_price_psm,
    subgroup = planning_region,
    subgroup2 = planning_area
  )
) +
  geom_treemap(colour = "white", linewidth = 0.2) +
  geom_treemap_subgroup2_border(colour = my_soft_teal, linewidth = 0.6) +
  geom_treemap_subgroup_border(colour = my_dark, linewidth = 1) +
  scale_fill_gradient(
    low = my_light_teal,
    high = my_teal,
    labels = comma
  ) +
  labs(
    title = "Treemap grouped by Planning Region and Planning Area",
    subtitle = "Two levels of subgroup borders make the hierarchy clearer",
    fill = "Median unit price\n(S$ psm)"
  ) +
  theme_yy
```

## Code

```{r}
ggplot(
  realis2018_selected,
  aes(
    area = total_unit_sold,
    fill = median_unit_price_psm,
    subgroup = planning_region,
    subgroup2 = planning_area
  )
) +
  geom_treemap(colour = "white", linewidth = 0.2) +
  geom_treemap_subgroup2_border(colour = my_soft_teal, linewidth = 0.6) +
  geom_treemap_subgroup_border(colour = my_dark, linewidth = 1) +
  scale_fill_gradient(
    low = my_light_teal,
    high = my_teal,
    labels = comma
  ) +
  labs(
    title = "Treemap grouped by Planning Region and Planning Area",
    subtitle = "Two levels of subgroup borders make the hierarchy clearer",
    fill = "Median unit price\n(S$ psm)"
  ) +
  theme_yy
```
:::

### 5.7.4 Adding boundary line and labels

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 7

ggplot(
  realis2018_selected,
  aes(
    area = total_unit_sold,
    fill = median_unit_price_psm,
    subgroup = planning_region,
    subgroup2 = planning_area,
    label = planning_area
  )
) +
  geom_treemap(colour = "white", linewidth = 0.2) +
  geom_treemap_subgroup2_border(colour = my_soft_teal, linewidth = 0.6) +
  geom_treemap_subgroup_border(colour = my_dark, linewidth = 1.1) +
  geom_treemap_subgroup_text(
    colour = my_dark,
    place = "centre",
    grow = FALSE,
    alpha = 0.85
  ) +
  scale_fill_gradient(
    low = my_light_teal,
    high = my_teal,
    labels = comma
  ) +
  labs(
    title = "Treemap with hierarchy boundary lines",
    subtitle = "Region and planning area boundaries improve readability",
    fill = "Median unit price\n(S$ psm)"
  ) +
  theme_yy
```

## Code

```{r}
ggplot(
  realis2018_selected,
  aes(
    area = total_unit_sold,
    fill = median_unit_price_psm,
    subgroup = planning_region,
    subgroup2 = planning_area,
    label = planning_area
  )
) +
  geom_treemap(colour = "white", linewidth = 0.2) +
  geom_treemap_subgroup2_border(colour = my_soft_teal, linewidth = 0.6) +
  geom_treemap_subgroup_border(colour = my_dark, linewidth = 1.1) +
  geom_treemap_subgroup_text(
    colour = my_dark,
    place = "centre",
    grow = FALSE,
    alpha = 0.85
  ) +
  scale_fill_gradient(
    low = my_light_teal,
    high = my_teal,
    labels = comma
  ) +
  labs(
    title = "Treemap with hierarchy boundary lines",
    subtitle = "Region and planning area boundaries improve readability",
    fill = "Median unit price\n(S$ psm)"
  ) +
  theme_yy
```
:::

This version is visually cleaner because the hierarchy is easier to identify.

## 5.8 Designing Interactive Treemap

The original chapter uses `d3treeR` for interactive treemap design. To avoid installation problems, this exercise uses `plotly`, which is more stable in Quarto.

### 5.8.1 Preparing data for interactive treemap

::: panel-tabset
## View

```{r}
#| echo: false

region_nodes <- realis2018_selected %>%
  group_by(planning_region) %>%
  summarise(
    label = first(planning_region),
    value = sum(total_unit_sold, na.rm = TRUE),
    color_value = median(median_unit_price_psm, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    id = paste0("region|", label),
    parent = "Singapore"
  ) %>%
  select(id, label, parent, value, color_value)

area_nodes <- realis2018_selected %>%
  group_by(planning_region, planning_area) %>%
  summarise(
    label = first(planning_area),
    value = sum(total_unit_sold, na.rm = TRUE),
    color_value = median(median_unit_price_psm, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    id = paste0("area|", planning_region, "|", label),
    parent = paste0("region|", planning_region)
  ) %>%
  select(id, label, parent, value, color_value)

project_nodes <- realis2018_selected %>%
  mutate(
    id = paste0("project|", planning_region, "|", planning_area, "|", project_name),
    label = project_name,
    parent = paste0("area|", planning_region, "|", planning_area),
    value = total_unit_sold,
    color_value = median_unit_price_psm
  ) %>%
  select(id, label, parent, value, color_value)

root_node <- tibble(
  id = "Singapore",
  label = "Singapore",
  parent = "",
  value = sum(region_nodes$value, na.rm = TRUE),
  color_value = median(region_nodes$color_value, na.rm = TRUE)
)

treemap_plotly_data <- bind_rows(
  root_node,
  region_nodes,
  area_nodes,
  project_nodes
)

DT::datatable(
  head(treemap_plotly_data, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
region_nodes <- realis2018_selected %>%
  group_by(planning_region) %>%
  summarise(
    label = first(planning_region),
    value = sum(total_unit_sold, na.rm = TRUE),
    color_value = median(median_unit_price_psm, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    id = paste0("region|", label),
    parent = "Singapore"
  ) %>%
  select(id, label, parent, value, color_value)

area_nodes <- realis2018_selected %>%
  group_by(planning_region, planning_area) %>%
  summarise(
    label = first(planning_area),
    value = sum(total_unit_sold, na.rm = TRUE),
    color_value = median(median_unit_price_psm, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    id = paste0("area|", planning_region, "|", label),
    parent = paste0("region|", planning_region)
  ) %>%
  select(id, label, parent, value, color_value)

project_nodes <- realis2018_selected %>%
  mutate(
    id = paste0("project|", planning_region, "|", planning_area, "|", project_name),
    label = project_name,
    parent = paste0("area|", planning_region, "|", planning_area),
    value = total_unit_sold,
    color_value = median_unit_price_psm
  ) %>%
  select(id, label, parent, value, color_value)

root_node <- tibble(
  id = "Singapore",
  label = "Singapore",
  parent = "",
  value = sum(region_nodes$value, na.rm = TRUE),
  color_value = median(region_nodes$color_value, na.rm = TRUE)
)

treemap_plotly_data <- bind_rows(
  root_node,
  region_nodes,
  area_nodes,
  project_nodes
)

DT::datatable(
  head(treemap_plotly_data, 12),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```
:::

### 5.8.2 Plotting interactive treemap

::: panel-tabset
## View

```{r}
#| echo: false

plot_ly(
  data = treemap_plotly_data,
  type = "treemap",
  ids = ~ id,
  labels = ~ label,
  parents = ~ parent,
  values = ~ value,
  branchvalues = "total",
  textinfo = "label+value",
  marker = list(
    colors = treemap_plotly_data$color_value,
    colorscale = list(
      list(0, my_light_teal),
      list(1, my_teal)
    ),
    showscale = TRUE,
    colorbar = list(title = "Median price")
  ),
  hovertemplate = paste(
    "<b>%{label}</b>",
    "<br>Total units sold: %{value}",
    "<br>Median unit price: %{color:.0f}",
    "<extra></extra>"
  )
) %>%
  layout(
    title = "Interactive Treemap of Resale Condominium Transactions, 2018"
  )
```

## Code

```{r}
plot_ly(
  data = treemap_plotly_data,
  type = "treemap",
  ids = ~ id,
  labels = ~ label,
  parents = ~ parent,
  values = ~ value,
  branchvalues = "total",
  textinfo = "label+value",
  marker = list(
    colors = treemap_plotly_data$color_value,
    colorscale = list(
      list(0, my_light_teal),
      list(1, my_teal)
    ),
    showscale = TRUE,
    colorbar = list(title = "Median price")
  ),
  hovertemplate = paste(
    "<b>%{label}</b>",
    "<br>Total units sold: %{value}",
    "<br>Median unit price: %{color:.0f}",
    "<extra></extra>"
  )
) %>%
  layout(
    title = "Interactive Treemap of Resale Condominium Transactions, 2018"
  )
```
:::

The interactive treemap allows the reader to move through the hierarchy and inspect details by hovering.

## 5.9 Extra Analysis: Top Planning Areas

This small extension summarises the largest planning areas by total units sold. It supports the treemap by making the ranking easier to read.

### 5.9.1 Ranking planning areas

::: panel-tabset
## View

```{r}
#| echo: false

planning_area_rank <- realis2018_selected %>%
  group_by(planning_region, planning_area) %>%
  summarise(
    total_unit_sold = sum(total_unit_sold, na.rm = TRUE),
    median_unit_price_psm = median(median_unit_price_psm, na.rm = TRUE),
    median_transacted_price = median(median_transacted_price, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_unit_sold))

DT::datatable(
  planning_area_rank %>%
    mutate(
      median_unit_price_psm = round(median_unit_price_psm, 0),
      median_transacted_price = round(median_transacted_price, 0)
    ),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```

## Code

```{r}
planning_area_rank <- realis2018_selected %>%
  group_by(planning_region, planning_area) %>%
  summarise(
    total_unit_sold = sum(total_unit_sold, na.rm = TRUE),
    median_unit_price_psm = median(median_unit_price_psm, na.rm = TRUE),
    median_transacted_price = median(median_transacted_price, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_unit_sold))

DT::datatable(
  planning_area_rank %>%
    mutate(
      median_unit_price_psm = round(median_unit_price_psm, 0),
      median_transacted_price = round(median_transacted_price, 0)
    ),
  rownames = FALSE,
  class = "compact stripe hover",
  options = list(pageLength = 12, scrollX = TRUE),
  width = "100%"
)
```
:::

### 5.9.2 Bar chart of top planning areas

::: panel-tabset
## View

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 6

top_area_rank <- planning_area_rank %>%
  slice_head(n = 15)

ggplot(
  top_area_rank,
  aes(
    x = fct_reorder(planning_area, total_unit_sold),
    y = total_unit_sold,
    fill = median_unit_price_psm
  )
) +
  geom_col(
    colour = my_dark,
    linewidth = 0.25
  ) +
  coord_flip() +
  scale_fill_gradient(
    low = my_light_teal,
    high = my_teal,
    labels = comma
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Top 15 planning areas by total units sold",
    subtitle = "This supporting chart makes the treemap ranking easier to read",
    x = NULL,
    y = "Total units sold",
    fill = "Median unit price\n(S$ psm)"
  ) +
  theme_yy
```

## Code

```{r}
top_area_rank <- planning_area_rank %>%
  slice_head(n = 15)

ggplot(
  top_area_rank,
  aes(
    x = fct_reorder(planning_area, total_unit_sold),
    y = total_unit_sold,
    fill = median_unit_price_psm
  )
) +
  geom_col(
    colour = my_dark,
    linewidth = 0.25
  ) +
  coord_flip() +
  scale_fill_gradient(
    low = my_light_teal,
    high = my_teal,
    labels = comma
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Top 15 planning areas by total units sold",
    subtitle = "This supporting chart makes the treemap ranking easier to read",
    x = NULL,
    y = "Total units sold",
    fill = "Median unit price\n(S$ psm)"
  ) +
  theme_yy
```
:::

## 5.10 Final Takeaway

Treemap visualisation is useful when the data has a hierarchy and when both size and colour can carry meaning. In this exercise, total units sold controls the rectangle size, while median unit price controls colour intensity.

The `treemap` package gives a quick way to create traditional treemaps. The `treemapify` package gives more control through the ggplot2 grammar. The interactive version helps readers inspect the hierarchy more directly.

::: callout-tip
## Final takeaway

Use treemaps when the main question is about hierarchical composition.\
For property transactions, treemaps help show how sales volume and price differ across region, area, and project level.
:::

## 5.11 Reference

This exercise covered:

-   importing and cleaning `realis2018.csv`,
-   summarising transaction records for treemap design,
-   creating static treemaps using `treemap`,
-   adjusting `vColor`, `type`, palette, algorithm, and `sortID`,
-   creating ggplot-style treemaps using `treemapify`,
-   creating an interactive treemap using `plotly`,
-   adding a simple top planning area ranking as a small extension.
