---
title: "Hands-on Exercise 9c - Analytical Mapping"
author: "Yang Yang"
format:
  html:
    theme: sketchy
    toc: true
    toc-depth: 3
    number-sections: false
    code-copy: true
    code-overflow: wrap
editor: visual
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

user_lib <- Sys.getenv("R_LIBS_USER")
if (nzchar(user_lib)) {
  dir.create(user_lib, recursive = TRUE, showWarnings = FALSE)
  .libPaths(c(user_lib, .libPaths()))
}

if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

pacman::p_load(
  sf,
  tmap,
  tidyverse,
  DT,
  knitr,
  scales
)

tmap_mode("plot")

my_teal <- "#2A6F6F"
my_soft_teal <- "#8AA6A3"
my_dark <- "#1F1F1F"
my_light_teal <- "#D8E8E6"
my_grey <- "#F3F4F4"
my_warning <- "#B07C3E"
my_red <- "#9A4B4B"
my_blue <- "#4C78A8"

find_data_file <- function(file_name) {
  candidates <- c(
    file.path("data", "rds", file_name),
    file.path("Hands-on_Ex", "Hands-on_Ex09c", "data", "rds", file_name)
  )
  found <- candidates[file.exists(candidates)]

  if (length(found) == 0) {
    stop(
      "Missing required data file: ", file_name,
      ". Put it inside Hands-on_Ex/Hands-on_Ex09c/data/rds/ before rendering."
    )
  }

  found[1]
}

theme_yy <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", colour = my_dark),
    plot.subtitle = element_text(colour = my_teal),
    axis.title = element_text(colour = my_dark),
    axis.text = element_text(colour = my_dark),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## 23.1 Overview

This hands-on exercise follows [Chapter 23 of R4VA](https://r4va.netlify.app/chap23), *Analytical Mapping*. The exercise uses a prepared polygon simple feature data frame of Nigeria water points at Local Government Area (LGA) level.

The uploaded file is named `NGA_wp (1).rds`, so this document reads that exact filename from the local `data/rds` folder.

### 23.1.1 Objectives

The objective is to gain hands-on experience using R methods to create analytical maps.

### 23.1.2 Learning Outcome

By the end of this exercise, the following tasks are practised:

- importing geospatial data in RDS format;
- creating choropleth maps with `tmap`;
- mapping counts and rates;
- creating percentile maps; and
- creating box maps to highlight extreme values.

## 23.2 Getting Started

### 23.2.1 Installing and Loading Packages

The setup chunk installs `pacman` if needed and loads the required packages.

```{r}
#| label: package-code
#| eval: false

if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

pacman::p_load(tmap, tidyverse, sf, DT, knitr, scales)
```

```{r}
#| label: package-table

tribble(
  ~Package, ~Purpose,
  "sf", "Read and handle simple feature polygon data.",
  "tmap", "Create choropleth, percentile, and box maps.",
  "tidyverse", "Wrangle the water point attributes and derive rates.",
  "DT", "Display interactive tables.",
  "knitr", "Create compact summary tables.",
  "scales", "Format rates and proportions."
) |>
  datatable(
    rownames = FALSE,
    options = list(pageLength = 10, dom = "tip")
  )
```

### 23.2.2 Importing Data

The data file is an RDS object. It is already a polygon `sf` data frame.

```{r}
#| label: import-nga

NGA_wp <- read_rds(find_data_file("NGA_wp (1).rds"))

NGA_wp
```

```{r}
#| label: data-metadata

tibble(
  measure = c(
    "Number of LGAs",
    "Number of columns",
    "Geometry type",
    "Coordinate reference system",
    "Minimum total water points",
    "Maximum total water points"
  ),
  value = c(
    nrow(NGA_wp),
    ncol(NGA_wp),
    paste(unique(st_geometry_type(NGA_wp)), collapse = ", "),
    st_crs(NGA_wp)$input,
    min(NGA_wp$total_wp, na.rm = TRUE),
    max(NGA_wp$total_wp, na.rm = TRUE)
  )
) |>
  kable()
```

```{r}
#| label: data-preview

NGA_wp |>
  st_drop_geometry() |>
  slice_head(n = 10) |>
  datatable(
    rownames = FALSE,
    options = list(pageLength = 10, scrollX = TRUE, dom = "tip")
  )
```

## 23.3 Basic Choropleth Mapping

### 23.3.1 Visualising Distribution of Functional and Total Water Points

The first analytical maps compare two count variables:

- `total_wp`: total number of water points in each LGA;
- `wp_functional`: number of functional water points in each LGA.

::: panel-tabset

## View

```{r}
#| label: count-maps-view
#| echo: false
#| fig-width: 11
#| fig-height: 5.5

p1 <- tm_shape(NGA_wp) +
  tm_polygons(
    fill = "wp_functional",
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 10,
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      title = "Functional",
      position = c("right", "bottom")
    )
  ) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Distribution of Functional Water Points by LGA")

p2 <- tm_shape(NGA_wp) +
  tm_polygons(
    fill = "total_wp",
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 10,
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      title = "Total",
      position = c("right", "bottom")
    )
  ) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Distribution of Total Water Points by LGA")

tmap_arrange(p2, p1, nrow = 1)
```

## Code

```{r}
#| label: count-maps-code
#| eval: false

p1 <- tm_shape(NGA_wp) +
  tm_polygons(
    fill = "wp_functional",
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 10,
      values = "brewer.blues"
    )
  ) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Distribution of Functional Water Points by LGA")

p2 <- tm_shape(NGA_wp) +
  tm_polygons(
    fill = "total_wp",
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 10,
      values = "brewer.blues"
    )
  ) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Distribution of Total Water Points by LGA")

tmap_arrange(p2, p1, nrow = 1)
```

:::

::: {.callout-note}
Count maps are useful, but they can over-emphasise LGAs with many water points. The rate maps in the next section adjust for the total number of water points in each LGA.
:::

## 23.4 Choropleth Map for Rates

### 23.4.1 Deriving Proportion of Functional and Non-Functional Water Points

Rates are derived by dividing by `total_wp`. LGAs with zero total water points are assigned `NA` for rate fields to avoid division by zero.

```{r}
#| label: derive-rates

NGA_wp <- NGA_wp |>
  mutate(
    pct_functional = if_else(
      total_wp > 0,
      wp_functional / total_wp,
      NA_real_
    ),
    pct_nonfunctional = if_else(
      total_wp > 0,
      wp_nonfunctional / total_wp,
      NA_real_
    )
  )

NGA_wp |>
  st_drop_geometry() |>
  summarise(
    lgas = n(),
    lgas_without_water_points = sum(total_wp == 0),
    mean_pct_functional = mean(pct_functional, na.rm = TRUE),
    mean_pct_nonfunctional = mean(pct_nonfunctional, na.rm = TRUE)
  ) |>
  mutate(
    mean_pct_functional = percent(mean_pct_functional, accuracy = 0.1),
    mean_pct_nonfunctional = percent(mean_pct_nonfunctional, accuracy = 0.1)
  ) |>
  kable()
```

### 23.4.2 Plotting Map of Rate

::: panel-tabset

## View

```{r}
#| label: rate-map-view
#| echo: false
#| fig-width: 8
#| fig-height: 7

tm_shape(NGA_wp) +
  tm_polygons(
    fill = "pct_functional",
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 10,
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      title = "Functional rate",
      position = c("right", "bottom")
    )
  ) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Rate Map of Functional Water Points by LGA")
```

## Code

```{r}
#| label: rate-map-code
#| eval: false

tm_shape(NGA_wp) +
  tm_polygons(
    fill = "pct_functional",
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 10,
      values = "brewer.blues"
    )
  ) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Rate Map of Functional Water Points by LGA")
```

:::

## 23.5 Extreme Value Maps

Extreme value maps are choropleth maps designed to highlight unusually low or unusually high values. They are useful for spatial exploratory data analysis.

### 23.5.1 Percentile Map

The percentile map uses six classes: 0-1%, 1-10%, 10-50%, 50-90%, 90-99%, and 99-100%.

#### 23.5.1.1 Data Preparation

```{r}
#| label: clean-data

NGA_wp_clean <- NGA_wp |>
  drop_na(pct_functional, pct_nonfunctional)

tibble(
  measure = c(
    "LGAs before dropping NA rates",
    "LGAs after dropping NA rates"
  ),
  value = c(nrow(NGA_wp), nrow(NGA_wp_clean))
) |>
  kable()
```

#### 23.5.1.2 Creating Customised Classification

```{r}
#| label: percentile-breaks

percent <- c(0, .01, .1, .5, .9, .99, 1)

total_wp_values <- NGA_wp_clean["total_wp"] |>
  st_set_geometry(NULL)

quantile(total_wp_values[, 1], percent)
```

#### 23.5.1.3 Creating the `get_var()` Function

```{r}
#| label: get-var-function

get_var <- function(vname, df) {
  v <- df[vname] |>
    st_set_geometry(NULL)
  v <- unname(v[, 1])
  return(v)
}
```

#### 23.5.1.4 A Percentile Mapping Function

```{r}
#| label: percentile-map-function

percentmap <- function(vnam, df, legtitle = NA, mtitle = "Percentile Map") {
  percent <- c(0, .01, .1, .5, .9, .99, 1)
  var <- get_var(vnam, df)
  bperc <- quantile(var, percent, na.rm = TRUE)

  tm_shape(df) +
    tm_polygons(
      fill = vnam,
      fill.scale = tm_scale_intervals(
        breaks = bperc,
        values = "brewer.blues",
        labels = c(
          "< 1%",
          "1% - 10%",
          "10% - 50%",
          "50% - 90%",
          "90% - 99%",
          "> 99%"
        )
      ),
      fill.legend = tm_legend(title = legtitle)
    ) +
    tm_borders(lwd = 0.1, fill_alpha = 1) +
    tm_title(mtitle)
}
```

#### 23.5.1.5 Test Drive the Percentile Mapping Function

::: panel-tabset

## View

```{r}
#| label: percentile-map-view
#| echo: false
#| fig-width: 8
#| fig-height: 7

percentmap(
  "total_wp",
  NGA_wp_clean,
  legtitle = "Total water points",
  mtitle = "Percentile Map of Total Water Points"
)
```

## Code

```{r}
#| label: percentile-map-code
#| eval: false

percentmap(
  "total_wp",
  NGA_wp_clean,
  legtitle = "Total water points",
  mtitle = "Percentile Map of Total Water Points"
)
```

:::

### 23.5.2 Box Map

A box map uses the logic of a boxplot to identify lower and upper spatial outliers.

```{r}
#| label: boxplot-view
#| fig-width: 5
#| fig-height: 4

ggplot(
  data = NGA_wp_clean |> st_drop_geometry(),
  aes(x = "", y = wp_nonfunctional)
) +
  geom_boxplot(fill = my_light_teal, colour = my_dark) +
  labs(
    title = "Boxplot of Non-Functional Water Points",
    x = NULL,
    y = "Non-functional water points"
  ) +
  theme_yy
```

#### 23.5.2.1 Creating the `boxbreaks()` Function

```{r}
#| label: boxbreaks-function

boxbreaks <- function(v, mult = 1.5) {
  qv <- unname(quantile(v, na.rm = TRUE))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr

  bb <- vector(mode = "numeric", length = 7)

  if (lofence < qv[1]) {
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }

  if (upfence > qv[5]) {
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }

  bb[3:5] <- qv[2:4]
  return(bb)
}
```

#### 23.5.2.2 Test Drive the Break Function

```{r}
#| label: boxbreaks-test

var <- get_var("wp_nonfunctional", NGA_wp_clean)
boxbreaks(var)
```

#### 23.5.2.3 Box Map Function

```{r}
#| label: boxmap-function

boxmap <- function(vnam, df, legtitle = NA, mtitle = "Box Map", mult = 1.5) {
  var <- get_var(vnam, df)
  bb <- boxbreaks(var, mult)

  tm_shape(df) +
    tm_polygons(
      fill = vnam,
      fill.scale = tm_scale_intervals(
        breaks = bb,
        values = "brewer.blues",
        labels = c(
          "Lower outlier",
          "< 25%",
          "25% - 50%",
          "50% - 75%",
          "> 75%",
          "Upper outlier"
        )
      ),
      fill.legend = tm_legend(title = legtitle)
    ) +
    tm_borders(lwd = 0.1, fill_alpha = 1) +
    tm_title(mtitle)
}
```

::: panel-tabset

## View

```{r}
#| label: boxmap-view
#| echo: false
#| fig-width: 8
#| fig-height: 7

boxmap(
  "wp_nonfunctional",
  NGA_wp_clean,
  legtitle = "Non-functional water points",
  mtitle = "Box Map of Non-Functional Water Points"
)
```

## Code

```{r}
#| label: boxmap-code
#| eval: false

boxmap(
  "wp_nonfunctional",
  NGA_wp_clean,
  legtitle = "Non-functional water points",
  mtitle = "Box Map of Non-Functional Water Points"
)
```

:::

## Additional Exploration

The table below lists LGAs with the highest non-functional water point rates among LGAs with at least 10 total water points.

```{r}
#| label: high-nonfunctional-table

NGA_wp |>
  st_drop_geometry() |>
  filter(total_wp >= 10) |>
  arrange(desc(pct_nonfunctional)) |>
  select(
    ADM1_EN,
    ADM2_EN,
    total_wp,
    wp_functional,
    wp_nonfunctional,
    pct_nonfunctional
  ) |>
  slice_head(n = 15) |>
  datatable(
    rownames = FALSE,
    options = list(pageLength = 10, scrollX = TRUE, dom = "tip")
  ) |>
  formatPercentage(columns = "pct_nonfunctional", digits = 1)
```

## Key Takeaways

::: {.callout-tip}
Analytical maps should be selected to match the question. Count maps show magnitude, rate maps control for the denominator, and extreme value maps highlight outliers.
:::

In this exercise:

- `read_rds()` imports the Nigeria water point polygon layer;
- `tm_polygons()` creates count and rate choropleth maps;
- functional and non-functional rates are derived safely by checking `total_wp > 0`;
- percentile maps highlight values in the top and bottom tails; and
- box maps use boxplot logic to identify lower and upper spatial outliers.

## References

- [R4VA Chapter 23: Analytical Mapping](https://r4va.netlify.app/chap23)
- [tmap: Thematic Maps in R](https://cran.r-project.org/package=tmap)
- [sf: Simple Features for R](https://cran.r-project.org/package=sf)
