---
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: true
    preview-links: auto
    width: 1600
    height: 900
    margin: 0.055
    footer: "ISSS608 Take-home Exercise 1"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

required_packages <- c(
  "tidyverse", "readxl", "janitor", "scales", "ggdist", "ggridges",
  "patchwork", "broom", "glue", "forcats", "stringr"
)

missing_packages <- required_packages[!required_packages %in% rownames(installed.packages())]
if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}

invisible(lapply(required_packages, library, character.only = TRUE))

my_blue <- "#4F91D9"
my_dark_blue <- "#1F4E79"
my_mid_blue <- "#6EA6D9"
my_light_blue <- "#D9E8F6"
my_dark <- "#1F1F1F"
my_grey <- "#666666"

theme_set(
  theme_minimal(base_size = 18) +
    theme(
      plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
      plot.subtitle = element_text(size = 14, colour = "grey35", hjust = 0.5),
      axis.title = element_text(size = 14),
      axis.text = element_text(size = 12),
      legend.position = "bottom",
      legend.title = element_text(size = 11),
      legend.text = element_text(size = 10),
      panel.grid.minor = element_blank(),
      plot.margin = margin(8, 8, 8, 8)
    )
)

safe_ratio <- function(numerator, denominator) {
  if_else(!is.na(denominator) & denominator > 0, numerator / denominator, NA_real_)
}

winsorise_upper <- function(x, p = 0.99) {
  cap <- quantile(x, probs = p, na.rm = TRUE)
  pmin(x, cap)
}

clean_lab <- function(x) {
  stringr::str_replace_all(as.character(x), "_", " ")
}

data_dir <- "data"

if (!dir.exists(data_dir)) {
  stop("The data folder was not found. Put the CSV and Excel dictionary inside Take-home_Ex/Take-home_Ex01/data/")
}

csv_file <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE)
xlsx_file <- list.files(data_dir, pattern = "\\.xlsx$", full.names = TRUE)

if (length(csv_file) == 0) stop("No CSV file found in the data folder.")
if (length(xlsx_file) == 0) stop("No Excel dictionary file found in the data folder.")

portfolio_raw <- readr::read_delim(
  csv_file[1],
  delim = ";",
  show_col_types = FALSE,
  trim_ws = TRUE
) |>
  janitor::clean_names()

dictionary_raw <- readxl::read_excel(
  xlsx_file[1],
  sheet = "Cartera"
) |>
  janitor::clean_names()

portfolio <- portfolio_raw |>
  mutate(
    across(
      c(policy_type, policy_status, business_type, payment_frequency,
        bonus_score, fuel_type, vehicle_brand, municipality_type, circulation_area),
      ~ forcats::fct_na_value_to_level(as.factor(.x), level = "Unknown")
    ),
    claim_flag = total_claims > 0,
    loss_ratio = safe_ratio(total_incurred, total_premium),
    profit = total_premium - total_incurred,
    claim_frequency = safe_ratio(total_claims, total_exposure),
    premium_rate = safe_ratio(total_premium, total_exposure),
    claim_severity = safe_ratio(total_incurred, total_claims),
    loss_ratio_w99 = winsorise_upper(loss_ratio, 0.99),
    incurred_positive = if_else(total_incurred > 0, total_incurred, NA_real_),
    driver_age_band = cut(
      driver_age,
      breaks = c(-Inf, 25, 35, 45, 55, 65, Inf),
      labels = c("≤25", "26–35", "36–45", "46–55", "56–65", "66+")
    ),
    vehicle_age_band = cut(
      vehicle_age,
      breaks = c(-Inf, 3, 7, 12, 20, Inf),
      labels = c("≤3", "4–7", "8–12", "13–20", "21+")
    )
  )

segment_summary <- function(data, segment_var) {
  data |>
    group_by({{ segment_var }}) |>
    summarise(
      policies = n(),
      exposure = sum(total_exposure, na.rm = TRUE),
      premium = sum(total_premium, na.rm = TRUE),
      incurred = sum(total_incurred, na.rm = TRUE),
      claims = sum(total_claims, na.rm = TRUE),
      portfolio_loss_ratio = incurred / premium,
      claim_frequency = claims / exposure,
      premium_rate = premium / exposure,
      avg_severity = incurred / claims,
      claim_share = claims / sum(data$total_claims, na.rm = TRUE),
      exposure_share = exposure / sum(data$total_exposure, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(desc(portfolio_loss_ratio))
}

annual_summary <- portfolio |>
  group_by(year) |>
  summarise(
    policies = n(),
    exposure = sum(total_exposure, na.rm = TRUE),
    premium = sum(total_premium, na.rm = TRUE),
    incurred = sum(total_incurred, na.rm = TRUE),
    claims = sum(total_claims, na.rm = TRUE),
    portfolio_loss_ratio = incurred / premium,
    claim_frequency = claims / exposure,
    premium_rate = premium / exposure,
    avg_severity = incurred / claims,
    .groups = "drop"
  )

policy_type_summary <- segment_summary(portfolio, policy_type)
bonus_summary <- segment_summary(portfolio, bonus_score)
payment_summary <- segment_summary(portfolio, payment_frequency)
status_summary <- segment_summary(portfolio, policy_status)
fuel_summary <- segment_summary(portfolio, fuel_type)
area_summary <- segment_summary(portfolio, circulation_area)

overall <- portfolio |>
  summarise(
    records = n(),
    exposure = sum(total_exposure, na.rm = TRUE),
    premium = sum(total_premium, na.rm = TRUE),
    incurred = sum(total_incurred, na.rm = TRUE),
    claims = sum(total_claims, na.rm = TRUE),
    portfolio_loss_ratio = incurred / premium,
    claim_frequency = claims / exposure,
    premium_rate = premium / exposure,
    avg_severity = incurred / claims,
    zero_claim_share = mean(total_claims == 0, na.rm = TRUE),
    positive_claim_share = mean(total_claims > 0, na.rm = TRUE)
  )

annual_change <- annual_summary |>
  arrange(year) |>
  summarise(
    premium_growth = last(premium) / first(premium) - 1,
    incurred_growth = last(incurred) / first(incurred) - 1,
    premium_rate_change = last(premium_rate) / first(premium_rate) - 1,
    loss_ratio_change_pp = (last(portfolio_loss_ratio) - first(portfolio_loss_ratio)) * 100,
    claim_frequency_change = last(claim_frequency) / first(claim_frequency) - 1
  )

worst_policy <- policy_type_summary |> slice_max(portfolio_loss_ratio, n = 1, with_ties = FALSE)
best_policy <- policy_type_summary |> slice_min(portfolio_loss_ratio, n = 1, with_ties = FALSE)
cancelled <- status_summary |> filter(policy_status == "C")
active <- status_summary |> filter(policy_status == "A")
worst_payment <- payment_summary |> slice_max(portfolio_loss_ratio, n = 1, with_ties = FALSE)
highest_freq_payment <- payment_summary |> slice_max(claim_frequency, n = 1, with_ties = FALSE)
highest_bonus <- bonus_summary |> slice_max(portfolio_loss_ratio, n = 1, with_ties = FALSE)
lowest_bonus <- bonus_summary |> slice_min(portfolio_loss_ratio, n = 1, with_ties = FALSE)

missing_tbl <- portfolio_raw |>
  summarise(across(everything(), ~sum(is.na(.x))/n())) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_pct") |>
  filter(missing_pct > 0) |>
  arrange(desc(missing_pct))

claim_count_tbl <- portfolio |>
  mutate(claim_count_label = if_else(total_claims >= 5, "5+", as.character(total_claims))) |>
  count(claim_count_label) |>
  mutate(
    claim_count_label = factor(claim_count_label, levels = c("0", "1", "2", "3", "4", "5+")),
    pct = n / sum(n)
  )

tail_tbl <- portfolio |>
  filter(total_incurred > 0) |>
  arrange(desc(total_incurred)) |>
  mutate(
    rank = row_number(),
    cum_claim_share = cumsum(total_incurred) / sum(total_incurred),
    policy_share = rank / n()
  )

top1_loss_share <- tail_tbl |> filter(policy_share <= 0.01) |> summarise(v = max(cum_claim_share)) |> pull(v)
top5_loss_share <- tail_tbl |> filter(policy_share <= 0.05) |> summarise(v = max(cum_claim_share)) |> pull(v)

method_tbl <- tibble(
  Stage = c("Distribution", "Fair comparison", "Interaction", "Validation"),
  Method = c("Raincloud / ridgeline", "Funnel plot", "Heatmap", "Kruskal-Wallis"),
  Why = c(
    "Shows skew, median, IQR and tail concentration",
    "Controls for exposure size before flagging outliers",
    "Finds risk pockets created by factor combinations",
    "Checks whether visual differences are statistically visible"
  )
)
```

## Cover {.cover-slide}
<div class="cover-container">
  <div class="cover-title">Motor Insurance Portfolio Profitability and Volatility</div>
  <div class="cover-subtitle">
    ISSS608 Take-home Exercise 1 · Visual Analytics for Data Discovery
  </div>
  <div class="cover-meta">
    Yang Yang · 23 May 2026
  </div>
</div>

<style>
/* =========================================================
   Global RevealJS slide styling
   ========================================================= */

.reveal {
  font-size: 28px;
}

/* Normal slide titles */
.reveal h1 {
  text-align: center;
  font-weight: 700;
  font-size: 1.55em;
  line-height: 1.10;
}

.reveal h2 {
  text-align: center;
  font-weight: 700;
  font-size: 1.35em;
  line-height: 1.08;
}

/* Normal paragraph and bullet text */
.reveal p,
.reveal li {
  font-size: 0.68em;
  line-height: 1.25;
}

/* Subtitle used in normal slides */
.reveal .subtitle {
  font-size: 0.78em;
  text-align: center;
  color: #444444;
  margin-top: -10px;
}

/* Observation text under charts */
.reveal .obs {
  font-size: 0.58em;
  line-height: 1.28;
  margin-top: 8px;
}

/* Smaller notes */
.reveal .small-note {
  font-size: 0.50em;
  line-height: 1.20;
}

/* Two-column layout */
.reveal .two-col {
  display: grid;
  grid-template-columns: 1fr 1fr;
  gap: 34px;
  align-items: start;
}

/* Card layout */
.reveal .content-card {
  border-left: 7px solid #4F91D9;
  background: #F6F9FC;
  padding: 16px 20px;
  margin-bottom: 14px;
}

.reveal .content-card strong,
.reveal strong {
  color: #1F4E79;
}

/* =========================================================
   Manual cover slide
   ========================================================= */

/* Force manual cover slide to use full slide height */
.reveal .slides section.cover-slide,
.reveal section.cover-slide {
  top: 0 !important;
  left: 0 !important;
  transform: none !important;

  width: 100% !important;
  height: 100vh !important;
  min-height: 100vh !important;

  padding: 0 !important;
  margin: 0 !important;

  display: flex !important;
  flex-direction: column !important;
  justify-content: center !important;
  align-items: center !important;

  text-align: center !important;
  box-sizing: border-box !important;
}

/* Hide the dummy slide heading "Cover" */
.reveal section.cover-slide h2 {
  display: none !important;
}

/* Centre title block */
.reveal section.cover-slide .cover-container {
  width: 84% !important;
  max-width: 1200px !important;
  margin: 0 auto !important;

  display: flex !important;
  flex-direction: column !important;
  justify-content: center !important;
  align-items: center !important;

  text-align: center !important;
  transform: translateY(-2vh);
}

/* Main cover title */
.reveal section.cover-slide .cover-title {
  font-size: 1.85em !important;
  font-weight: 700 !important;
  line-height: 1.12 !important;
  text-align: center !important;
  color: #111111 !important;
  margin-bottom: 0.55em !important;
}

/* Cover subtitle */
.reveal section.cover-slide .cover-subtitle {
  font-size: 0.78em !important;
  font-weight: 400 !important;
  line-height: 1.35 !important;
  text-align: center !important;
  color: #333333 !important;
  margin-bottom: 0.45em !important;
}

/* Cover author/date */
.reveal section.cover-slide .cover-meta {
  font-size: 0.68em !important;
  font-weight: 400 !important;
  line-height: 1.30 !important;
  text-align: center !important;
  color: #333333 !important;
}

/* =========================================================
   Fallback: if Quarto auto title slide exists
   ========================================================= */

.reveal section#title-slide {
  top: 0 !important;
  left: 0 !important;
  transform: none !important;
  height: 100vh !important;
  min-height: 100vh !important;
  width: 100% !important;

  display: flex !important;
  flex-direction: column !important;
  justify-content: center !important;
  align-items: center !important;

  text-align: center !important;
  padding: 0 7% !important;
  box-sizing: border-box !important;
}

.reveal section#title-slide .quarto-title-block,
.reveal section#title-slide .quarto-title,
.reveal section#title-slide .title {
  text-align: center !important;
  margin-left: auto !important;
  margin-right: auto !important;
}

.reveal section#title-slide h1.title,
.reveal section#title-slide h1 {
  font-size: 2.05em !important;
  line-height: 1.12 !important;
  font-weight: 700 !important;
  text-align: center !important;
  margin-top: 0 !important;
  margin-bottom: 0.35em !important;
}

.reveal section#title-slide .subtitle,
.reveal section#title-slide p.subtitle {
  font-size: 0.82em !important;
  line-height: 1.25 !important;
  color: #444444 !important;
  text-align: center !important;
  margin-top: 0 !important;
  margin-bottom: 1.2em !important;
}

.reveal section#title-slide .author,
.reveal section#title-slide .date,
.reveal section#title-slide p.author,
.reveal section#title-slide p.date {
  font-size: 0.68em !important;
  line-height: 1.20 !important;
  text-align: center !important;
  color: #333333 !important;
  margin-top: 0.20em !important;
  margin-bottom: 0.20em !important;
}
</style>

::: notes
This executive summary presents the portfolio-level and segment-level findings from the motor insurance portfolio dataset. The analysis follows the assignment requirement: data preparation, univariate visual analytics, bivariate visual analytics, statistical validation, and management implications. The focus is profitability and volatility, measured through premium adequacy, claim frequency, claim severity, and tail-risk behaviour.
:::

## Contents

::::::::::: columns
:::::: {.column width="50%"}
::: content-card
<strong>1. Data and KPI design</strong><br> Policy-year unit, exposure adjustment, and profitability metrics.
:::

::: content-card
<strong>2. Portfolio-level diagnostics</strong><br> Growth, loss-ratio trend, zero claims, and loss-tail behaviour.
:::

::: content-card
<strong>3. Segment-level visual discovery</strong><br> Product, bonus score, payment frequency, policy status, and geography.
:::
::::::

:::::: {.column width="50%"}
::: content-card
<strong>4. Advanced visual analytics</strong><br> Heatmap, funnel plot, raincloud plot, and ridgeline comparison.
:::

::: content-card
<strong>5. Statistical support</strong><br> Non-parametric validation of segment differences.
:::

::: content-card
<strong>6. Management implications</strong><br> Where underwriting and pricing attention should go first.
:::
::::::
:::::::::::

::: notes
The presentation is structured as a decision-oriented visual analysis, not a chart gallery. Each section answers a specific management question: whether the portfolio is profitable, where volatility appears, which segments explain the pressure, and which findings are strong enough to guide further underwriting or pricing review.
:::

## Observation 1 — The dataset supports exposure-adjusted insurance analysis

```{r}
#| fig-width: 13
#| fig-height: 4.8

p_records <- annual_summary |>
  ggplot(aes(x = factor(year), y = policies)) +
  geom_col(fill = my_blue, colour = my_dark, linewidth = 0.35) +
  geom_text(aes(label = comma(policies)), vjust = -0.35, size = 4.5) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale()), expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Policy-year records", x = NULL, y = "Records")

p_exposure <- annual_summary |>
  ggplot(aes(x = factor(year), y = exposure)) +
  geom_col(fill = my_mid_blue, colour = my_dark, linewidth = 0.35) +
  geom_text(aes(label = comma(round(exposure, 0))), vjust = -0.35, size = 4.5) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale()), expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Total exposure", x = NULL, y = "Exposure years")

if (nrow(missing_tbl) > 0) {
  p_missing <- missing_tbl |>
    slice_head(n = 6) |>
    mutate(variable = fct_reorder(variable, missing_pct)) |>
    ggplot(aes(x = variable, y = missing_pct)) +
    geom_col(fill = my_light_blue, colour = my_dark, linewidth = 0.35) +
    geom_text(aes(label = percent(missing_pct, accuracy = 0.1)), hjust = -0.05, size = 4) +
    coord_flip() +
    scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.25))) +
    labs(title = "Missingness check", x = NULL, y = "Missing share")
} else {
  p_missing <- ggplot() +
    annotate("text", x = 1, y = 1, label = "No missing values detected", size = 6, colour = my_dark_blue) +
    theme_void() +
    labs(title = "Missingness check")
}

p_records + p_exposure + p_missing
```

::: obs
-   Dataset contains **`r comma(overall$records)` policy-year records** and **\`r comma(round(overall\$exposure, 0)) exposure years**.
-   The analytical unit is appropriate because premium, claims, incurred losses and exposure are all available.
-   **Implication: profitability must be analysed using exposure-adjusted and portfolio-weighted metrics, not raw averages.**
:::

::: notes
This slide validates the data foundation. The records are policy-year observations, which is suitable for insurance analysis because each row links exposure, premium, claim counts, and incurred losses. The most important methodological choice is to use exposure-adjusted metrics. Without exposure adjustment, a short-duration policy and a full-year policy would be treated as equally informative, which would distort both profitability and frequency analysis.
:::

## Analytical Framework — Profitability and volatility are decomposed

```{r}
#| fig-width: 12.5
#| fig-height: 4.8

kpi_tbl <- tibble(
  Layer = factor(c("Profitability", "Frequency", "Severity", "Volatility"),
                 levels = c("Profitability", "Frequency", "Severity", "Volatility")),
  KPI = c("Loss ratio\nincurred ÷ premium",
          "Claim frequency\nclaims ÷ exposure",
          "Claim severity\nincurred ÷ claims",
          "Tail volatility\nskew and extreme loss"),
  Interpretation = c("Is pricing adequate?",
                     "How often do claims happen?",
                     "How costly is each claim?",
                     "How unstable is the segment?")
)

kpi_tbl |>
  ggplot(aes(x = Layer, y = 1, fill = Layer)) +
  geom_col(width = 0.72, colour = my_dark, linewidth = 0.35) +
  geom_text(aes(label = KPI), y = 0.68, size = 4.6, colour = "white", fontface = "bold", lineheight = 0.95) +
  geom_text(aes(label = Interpretation), y = 0.30, size = 3.7, colour = "white", lineheight = 0.95) +
  scale_fill_manual(values = c("#1F4E79", "#2F6DAE", "#6B9FD3", "#A9C7E8")) +
  coord_cartesian(ylim = c(0, 1.08)) +
  theme_void(base_size = 18) +
  theme(legend.position = "none") +
  labs(title = "Framework used to avoid misleading one-number profitability conclusions")
```

::: obs
-   **Loss ratio** is the primary profitability KPI.
-   **Frequency and severity** explain why the loss ratio changes.
-   **Tail volatility** identifies segments where averages are unstable because of extreme claims.
:::

::: notes
This slide fixes the interpretation logic. A high loss ratio alone is not enough. It could come from too many claims, expensive claims, inadequate premium, or a small number of extreme losses. Therefore the analysis separates profitability, frequency, severity, and volatility. This is also why several chart types are needed rather than a single bar chart.
:::

## Observation 2 — Portfolio growth did not translate into stable profitability

```{r}
#| fig-width: 13
#| fig-height: 4.8

p_money <- annual_summary |>
  pivot_longer(c(premium, incurred), names_to = "measure", values_to = "value") |>
  mutate(measure = recode(measure, premium = "Premium", incurred = "Incurred loss")) |>
  ggplot(aes(x = factor(year), y = value, fill = measure)) +
  geom_col(position = "dodge", colour = my_dark, linewidth = 0.3) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_fill_manual(values = c("Premium" = my_light_blue, "Incurred loss" = my_blue)) +
  labs(title = "Premium vs incurred", x = NULL, y = "Amount")

p_lr_trend <- annual_summary |>
  ggplot(aes(x = year, y = portfolio_loss_ratio)) +
  geom_line(linewidth = 1.2, colour = my_blue) +
  geom_point(size = 3.5, colour = my_blue) +
  geom_text(aes(label = percent(portfolio_loss_ratio, accuracy = 0.1)), vjust = -0.8, size = 4.2) +
  scale_x_continuous(breaks = annual_summary$year) +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0.06, 0.20))) +
  labs(title = "Loss ratio trend", x = NULL, y = "Loss ratio")

p_rate <- annual_summary |>
  ggplot(aes(x = year, y = premium_rate)) +
  geom_line(linewidth = 1.2, colour = my_blue) +
  geom_point(size = 3.5, colour = my_blue) +
  geom_text(aes(label = dollar(premium_rate, prefix = "€", accuracy = 1)), vjust = -0.8, size = 4.2) +
  scale_x_continuous(breaks = annual_summary$year) +
  scale_y_continuous(labels = dollar_format(prefix = "€"), expand = expansion(mult = c(0.06, 0.20))) +
  labs(title = "Premium per exposure year", x = NULL, y = "Premium rate")

p_money + p_lr_trend + p_rate
```

::: obs
-   Premium grew by **`r percent(annual_change$premium_growth, accuracy = 0.1)`**, while incurred losses grew by **`r percent(annual_change$incurred_growth, accuracy = 0.1)`**.
-   Portfolio loss ratio rose by **`r number(annual_change$loss_ratio_change_pp, accuracy = 0.1)` percentage points**.
-   **Implication: growth requires segment-level pricing control; volume alone did not stabilise profitability.**
:::

::: notes
The portfolio expanded, but the profitability trend is less favourable. The key comparison is not only premium against incurred amount, but also the loss ratio and premium rate. This shows whether the insurer is being paid adequately for the exposure it takes on. The visual pattern suggests that growth should be monitored together with underwriting quality.
:::

## Observation 3 — Claims are rare, but losses are tail-driven

```{r}
#| fig-width: 13
#| fig-height: 4.8

p_claim_count <- claim_count_tbl |>
  ggplot(aes(x = claim_count_label, y = pct)) +
  geom_col(fill = my_blue, colour = my_dark, linewidth = 0.35) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), vjust = -0.35, size = 4.2) +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Claim count concentration", x = "Claims per policy-year", y = "Share")

p_incurred_positive <- portfolio |>
  filter(total_incurred > 0) |>
  ggplot(aes(x = log1p(total_incurred))) +
  geom_histogram(bins = 45, fill = my_blue, colour = "white") +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(title = "Positive incurred loss", subtitle = "Zero-loss rows removed; log scale used", x = "log(1 + incurred)", y = "Count")

p_tail <- tail_tbl |>
  filter(policy_share <= 0.20) |>
  ggplot(aes(x = policy_share, y = cum_claim_share)) +
  geom_line(linewidth = 1.25, colour = my_blue) +
  geom_vline(xintercept = c(0.01, 0.05), linetype = "dashed", colour = "grey45") +
  geom_point(data = tibble(
    policy_share = c(0.01, 0.05),
    cum_claim_share = c(top1_loss_share, top5_loss_share)
  ), aes(x = policy_share, y = cum_claim_share), size = 3, colour = my_dark_blue) +
  geom_text(data = tibble(
    policy_share = c(0.01, 0.05),
    cum_claim_share = c(top1_loss_share, top5_loss_share),
    label = c(paste0("Top 1%: ", percent(top1_loss_share, accuracy = 1)),
              paste0("Top 5%: ", percent(top5_loss_share, accuracy = 1)))
  ), aes(label = label), vjust = -0.7, size = 3.7) +
  scale_x_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.20)) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(title = "Loss concentration curve", x = "Share of positive-loss policies", y = "Cumulative incurred loss")

p_claim_count + p_incurred_positive + p_tail
```

::: obs
-   **`r percent(overall$zero_claim_share, accuracy = 0.1)`** of policy-year records have zero claims.
-   Among positive-loss policies, the top **1%** contributes **`r percent(top1_loss_share, accuracy = 0.1)`** of incurred loss.
-   **Implication: frequency and severity must be modelled separately; raw mean loss is unstable.**
:::

::: notes
The previous version looked blank because zero-loss rows dominated the histograms. This revised slide separates claim count, positive incurred loss, and tail concentration. The chart choice is intentional: the bar chart shows claim rarity, the log histogram makes positive losses visible, and the concentration curve quantifies tail dependence.
:::

## Insight 1 — Product type is the strongest first-level profitability separator

```{r}
#| fig-width: 13
#| fig-height: 4.8

p_policy_lr <- policy_type_summary |>
  mutate(policy_type = fct_reorder(policy_type, portfolio_loss_ratio)) |>
  ggplot(aes(x = policy_type, y = portfolio_loss_ratio)) +
  geom_col(fill = my_blue, colour = my_dark, linewidth = 0.35) +
  geom_text(aes(label = percent(portfolio_loss_ratio, accuracy = 0.1)), hjust = -0.05, size = 4.1) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "firebrick") +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.24))) +
  labs(title = "Loss ratio", x = NULL, y = NULL)

p_policy_freq <- policy_type_summary |>
  mutate(policy_type = fct_reorder(policy_type, claim_frequency)) |>
  ggplot(aes(x = policy_type, y = claim_frequency)) +
  geom_col(fill = my_light_blue, colour = my_dark, linewidth = 0.35) +
  geom_text(aes(label = number(claim_frequency, accuracy = 0.01)), hjust = -0.05, size = 4.1) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.24))) +
  labs(title = "Claim frequency", x = NULL, y = "Claims / exposure")

p_policy_sev <- policy_type_summary |>
  mutate(policy_type = fct_reorder(policy_type, avg_severity)) |>
  ggplot(aes(x = policy_type, y = avg_severity)) +
  geom_col(fill = my_mid_blue, colour = my_dark, linewidth = 0.35) +
  geom_text(aes(label = dollar(avg_severity, prefix = "€", accuracy = 1)), hjust = -0.05, size = 4.1) +
  coord_flip() +
  scale_y_continuous(labels = dollar_format(prefix = "€"), expand = expansion(mult = c(0, 0.30))) +
  labs(title = "Average severity", x = NULL, y = NULL)

p_policy_lr + p_policy_freq + p_policy_sev
```

::: obs
-   Worst product: **`r as.character(worst_policy$policy_type)`**, loss ratio **`r percent(worst_policy$portfolio_loss_ratio, accuracy = 0.1)`**.
-   Best product: **`r as.character(best_policy$policy_type)`**, loss ratio **`r percent(best_policy$portfolio_loss_ratio, accuracy = 0.1)`**.
-   **Implication: product review should decompose loss ratio into claim frequency and severity before repricing.**
:::

::: notes
Product type is a natural first segmentation because it reflects coverage structure. Three charts are used rather than one because loss ratio alone does not explain the mechanism. The frequency and severity panels show whether profitability pressure comes from more claims or more expensive claims. This directly supports underwriting and pricing decisions.
:::

## Insight 2 — Bonus score carries risk information not fully neutralised by premium

```{r}
#| fig-width: 13
#| fig-height: 4.8

bonus_long <- bonus_summary |>
  select(bonus_score, portfolio_loss_ratio, claim_frequency, premium_rate) |>
  pivot_longer(
    c(portfolio_loss_ratio, claim_frequency, premium_rate),
    names_to = "metric", values_to = "value"
  ) |>
  mutate(
    metric = recode(metric,
      portfolio_loss_ratio = "Loss ratio",
      claim_frequency = "Claim frequency",
      premium_rate = "Premium rate"
    )
  )

p_bonus_lr <- bonus_summary |>
  ggplot(aes(x = bonus_score, y = portfolio_loss_ratio, group = 1)) +
  geom_line(linewidth = 1.1, colour = my_blue) +
  geom_point(size = 3, colour = my_blue) +
  geom_text(aes(label = percent(portfolio_loss_ratio, accuracy = 1)), vjust = -0.8, size = 3.8) +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0.05, 0.18))) +
  labs(title = "Loss ratio by bonus score", x = "Bonus score", y = NULL)

p_bonus_freq <- bonus_summary |>
  ggplot(aes(x = bonus_score, y = claim_frequency, group = 1)) +
  geom_line(linewidth = 1.1, colour = my_blue) +
  geom_point(size = 3, colour = my_blue) +
  geom_text(aes(label = number(claim_frequency, accuracy = 0.01)), vjust = -0.8, size = 3.8) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.18))) +
  labs(title = "Claim frequency by bonus score", x = "Bonus score", y = "Claims / exposure")

p_bonus_rate <- bonus_summary |>
  ggplot(aes(x = bonus_score, y = premium_rate, group = 1)) +
  geom_line(linewidth = 1.1, colour = my_blue) +
  geom_point(size = 3, colour = my_blue) +
  geom_text(aes(label = dollar(premium_rate, prefix = "€", accuracy = 1)), vjust = -0.8, size = 3.8) +
  scale_y_continuous(labels = dollar_format(prefix = "€"), expand = expansion(mult = c(0.05, 0.18))) +
  labs(title = "Premium rate by bonus score", x = "Bonus score", y = NULL)

p_bonus_lr + p_bonus_freq + p_bonus_rate
```

::: obs
-   Highest-loss bonus group: **`r as.character(highest_bonus$bonus_score)`**, loss ratio **`r percent(highest_bonus$portfolio_loss_ratio, accuracy = 0.1)`**.
-   Lowest-loss bonus group: **`r as.character(lowest_bonus$bonus_score)`**, loss ratio **`r percent(lowest_bonus$portfolio_loss_ratio, accuracy = 0.1)`**.
-   **Implication: bonus score should remain in monitoring because premium adjustment does not fully flatten risk differences.**
:::

::: notes
A line chart is appropriate because bonus score is ordered. It reveals whether risk moves monotonically or irregularly across score levels. The slide compares loss ratio, claim frequency, and premium rate to check whether higher-risk groups are being compensated by pricing. This is more informative than a single distribution plot.
:::

## Insight 3 — Payment frequency reveals behavioural risk differences

```{r}
#| fig-width: 13
#| fig-height: 4.8

p_pay_lr <- payment_summary |>
  mutate(payment_frequency = fct_reorder(payment_frequency, portfolio_loss_ratio)) |>
  ggplot(aes(x = payment_frequency, y = portfolio_loss_ratio)) +
  geom_col(fill = my_blue, colour = my_dark, linewidth = 0.35) +
  geom_text(aes(label = percent(portfolio_loss_ratio, accuracy = 0.1)), hjust = -0.05, size = 4.2) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.24))) +
  labs(title = "Loss ratio", x = NULL, y = NULL)

p_pay_freq <- payment_summary |>
  mutate(payment_frequency = fct_reorder(payment_frequency, claim_frequency)) |>
  ggplot(aes(x = payment_frequency, y = claim_frequency)) +
  geom_col(fill = my_light_blue, colour = my_dark, linewidth = 0.35) +
  geom_text(aes(label = number(claim_frequency, accuracy = 0.01)), hjust = -0.05, size = 4.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.24))) +
  labs(title = "Claim frequency", x = NULL, y = "Claims / exposure")

p_pay_rate <- payment_summary |>
  mutate(payment_frequency = fct_reorder(payment_frequency, premium_rate)) |>
  ggplot(aes(x = payment_frequency, y = premium_rate)) +
  geom_col(fill = my_mid_blue, colour = my_dark, linewidth = 0.35) +
  geom_text(aes(label = dollar(premium_rate, prefix = "€", accuracy = 1)), hjust = -0.05, size = 4.2) +
  coord_flip() +
  scale_y_continuous(labels = dollar_format(prefix = "€"), expand = expansion(mult = c(0, 0.28))) +
  labs(title = "Premium rate", x = NULL, y = NULL)

p_pay_lr + p_pay_freq + p_pay_rate
```

::: obs
-   Highest claim-frequency payment group: **`r as.character(highest_freq_payment$payment_frequency)`**, **`r number(highest_freq_payment$claim_frequency, accuracy = 0.01)` claims per exposure year**.
-   Worst loss-ratio payment group: **`r as.character(worst_payment$payment_frequency)`**, loss ratio **`r percent(worst_payment$portfolio_loss_ratio, accuracy = 0.1)`**.
-   **Implication: payment frequency is a behavioural risk signal, not only an administrative variable.**
:::

::: notes
Payment frequency is not a classic vehicle-risk variable, but it can reveal behaviour and selection effects. The bar chart is suitable here because categories are nominal and the purpose is ranking. The interpretation is careful: this shows association, not causality. It is a useful dashboard variable for underwriting and retention monitoring.
:::

## Insight 4 — Cancelled policies are materially adverse

```{r}
#| fig-width: 13
#| fig-height: 4.8

status_summary_clean <- status_summary |>
  mutate(policy_status = fct_recode(policy_status, Active = "A", Cancelled = "C"))

p_status_lr <- status_summary_clean |>
  ggplot(aes(x = policy_status, y = portfolio_loss_ratio)) +
  geom_col(fill = my_blue, colour = my_dark, linewidth = 0.35, width = 0.65) +
  geom_text(aes(label = percent(portfolio_loss_ratio, accuracy = 0.1)), vjust = -0.35, size = 5) +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.17))) +
  labs(title = "Loss ratio", x = "Policy status", y = NULL)

p_status_freq <- status_summary_clean |>
  ggplot(aes(x = policy_status, y = claim_frequency)) +
  geom_col(fill = my_light_blue, colour = my_dark, linewidth = 0.35, width = 0.65) +
  geom_text(aes(label = number(claim_frequency, accuracy = 0.01)), vjust = -0.35, size = 5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.17))) +
  labs(title = "Claim frequency", x = "Policy status", y = "Claims / exposure")

p_status_sev <- status_summary_clean |>
  ggplot(aes(x = policy_status, y = avg_severity)) +
  geom_col(fill = my_mid_blue, colour = my_dark, linewidth = 0.35, width = 0.65) +
  geom_text(aes(label = dollar(avg_severity, prefix = "€", accuracy = 1)), vjust = -0.35, size = 5) +
  scale_y_continuous(labels = dollar_format(prefix = "€"), expand = expansion(mult = c(0, 0.20))) +
  labs(title = "Average severity", x = "Policy status", y = NULL)

p_status_lr + p_status_freq + p_status_sev
```

::: obs
-   Cancelled policies show loss ratio of **`r percent(cancelled$portfolio_loss_ratio, accuracy = 0.1)`**, versus **`r percent(active$portfolio_loss_ratio, accuracy = 0.1)`** for active policies.
-   **Implication: cancellation is an adverse-experience monitoring signal, but should not be interpreted as causal without event sequencing.**
:::

::: notes
The chart compares status groups across profitability, frequency, and severity. This slide is strong for decision making but needs cautious wording. A cancelled policy may have cancelled after a claim; therefore, the status variable may partly be an outcome. The correct business action is to monitor cancellations as adverse-experience signals and investigate timing.
:::

## Insight 5 — Geography and usage context create uneven profitability pockets

```{r}
#| label: insight-5-geography-usage
#| echo: false
#| fig-width: 13.8
#| fig-height: 5.7
#| out-width: "100%"

area_display <- area_summary |>
  mutate(circulation_area = fct_reorder(circulation_area, portfolio_loss_ratio))

p_area_lr <- area_display |>
  ggplot(aes(
    x = circulation_area,
    y = portfolio_loss_ratio
  )) +
  geom_col(fill = my_blue, color = my_dark, linewidth = 0.3, width = 0.72) +
  geom_text(
    aes(label = percent(portfolio_loss_ratio, accuracy = 0.1)),
    hjust = -0.08,
    size = 3.3
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.18))
  ) +
  labs(
    title = "Loss ratio\nby circulation area",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(
      size = 11,
      face = "bold",
      hjust = 0.5,
      lineheight = 0.95,
      margin = margin(b = 8)
    ),
    axis.text = element_text(size = 9),
    panel.grid.major.y = element_blank(),
    plot.margin = margin(6, 20, 6, 6)
  )

p_area_freq <- area_display |>
  ggplot(aes(
    x = circulation_area,
    y = claim_frequency
  )) +
  geom_col(fill = "#D7E6F5", color = my_dark, linewidth = 0.3, width = 0.72) +
  geom_text(
    aes(label = number(claim_frequency, accuracy = 0.01)),
    hjust = -0.08,
    size = 3.3
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.18))
  ) +
  labs(
    title = "Claim frequency\nby circulation area",
    x = NULL,
    y = "Claims / exposure"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(
      size = 11,
      face = "bold",
      hjust = 0.5,
      lineheight = 0.95,
      margin = margin(b = 8)
    ),
    axis.text = element_text(size = 9),
    axis.title.x = element_text(size = 9, margin = margin(t = 6)),
    panel.grid.major.y = element_blank(),
    plot.margin = margin(6, 20, 6, 6)
  )

p_fuel <- fuel_summary |>
  mutate(fuel_type = fct_reorder(fuel_type, portfolio_loss_ratio)) |>
  ggplot(aes(
    x = fuel_type,
    y = portfolio_loss_ratio
  )) +
  geom_col(fill = "#6FA8DC", color = my_dark, linewidth = 0.3, width = 0.72) +
  geom_text(
    aes(label = percent(portfolio_loss_ratio, accuracy = 0.1)),
    hjust = -0.08,
    size = 3.3
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.18))
  ) +
  labs(
    title = "Loss ratio\nby fuel type",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(
      size = 11,
      face = "bold",
      hjust = 0.5,
      lineheight = 0.95,
      margin = margin(b = 8)
    ),
    axis.text = element_text(size = 9),
    panel.grid.major.y = element_blank(),
    plot.margin = margin(6, 20, 6, 6)
  )

(p_area_lr | p_area_freq | p_fuel) +
  plot_layout(widths = c(1, 1, 1))
```

::: obs
-   Circulation area and fuel type show material differences in loss ratio and frequency.
-   These variables are useful because they reflect **contextual risk exposure**, not only driver or product characteristics.
-   **Implication: rating review should check whether contextual variables are adequately priced.**
:::

::: notes
This slide broadens the analysis beyond product and customer variables. Circulation area and fuel type are not causal proofs, but they are plausible risk context variables. The chart design remains simple because the management question is categorical comparison. It identifies which contextual segments deserve deeper pricing review.
:::

## Insight 6 — Driver age and vehicle age interact, not just act independently

```{r}
#| fig-width: 11.5
#| fig-height: 5.1

age_heatmap <- portfolio |>
  filter(!is.na(driver_age_band), !is.na(vehicle_age_band)) |>
  group_by(driver_age_band, vehicle_age_band) |>
  summarise(
    policies = n(),
    premium = sum(total_premium, na.rm = TRUE),
    incurred = sum(total_incurred, na.rm = TRUE),
    claims = sum(total_claims, na.rm = TRUE),
    exposure = sum(total_exposure, na.rm = TRUE),
    portfolio_loss_ratio = incurred / premium,
    claim_frequency = claims / exposure,
    .groups = "drop"
  ) |>
  filter(policies >= 100)

age_heatmap |>
  ggplot(aes(x = vehicle_age_band, y = driver_age_band, fill = portfolio_loss_ratio)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = percent(portfolio_loss_ratio, accuracy = 1)), size = 4.2, colour = my_dark) +
  scale_fill_gradient(low = "#D8E6F3", high = "#1F4E79", guide = "none") +
  labs(
    title = "Heatmap: loss ratio by driver age × vehicle age",
    subtitle = "Cells with fewer than 100 policy-years excluded; colour intensity shows loss ratio",
    x = "Vehicle age band", y = "Driver age band"
  )
```

::: obs
-   Heatmap evidence shows that profitability varies by **combination** of driver age and vehicle age.
-   The chart is used because the analytical question is interactional; a one-variable bar chart would hide joint risk pockets.
-   **Implication: interaction-based monitoring is more useful than checking one rating factor at a time.**
:::

::: notes
The legend bug is removed because the exact values are printed inside each tile. A heatmap is the correct visual method because it shows interaction between two categorical bands. This satisfies the bivariate visual analytics requirement and gives a more decision-useful insight than two separate bar charts.
:::

## Method Add-on 1 — Funnel plot avoids overreacting to small segments

```{r}
#| fig-width: 12
#| fig-height: 5.1

brand_funnel <- portfolio |>
  group_by(vehicle_brand) |>
  summarise(
    exposure = sum(total_exposure, na.rm = TRUE),
    claims = sum(total_claims, na.rm = TRUE),
    claim_frequency = claims / exposure,
    .groups = "drop"
  ) |>
  filter(exposure >= quantile(exposure, 0.50, na.rm = TRUE)) |>
  mutate(vehicle_brand = fct_lump_n(vehicle_brand, n = 25, w = exposure)) |>
  group_by(vehicle_brand) |>
  summarise(
    exposure = sum(exposure),
    claims = sum(claims),
    claim_frequency = claims / exposure,
    .groups = "drop"
  ) |>
  filter(vehicle_brand != "Other")

overall_rate <- overall$claims / overall$exposure

funnel_curve <- tibble(
  exposure = seq(min(brand_funnel$exposure), max(brand_funnel$exposure), length.out = 200),
  expected = overall_rate,
  upper_95 = overall_rate + 1.96 * sqrt(overall_rate / exposure),
  lower_95 = pmax(overall_rate - 1.96 * sqrt(overall_rate / exposure), 0)
)

flagged_brand <- brand_funnel |>
  mutate(distance = abs(claim_frequency - overall_rate)) |>
  slice_max(distance, n = 6, with_ties = FALSE)

ggplot() +
  geom_ribbon(
    data = funnel_curve,
    aes(x = exposure, ymin = lower_95, ymax = upper_95),
    fill = my_light_blue, alpha = 0.8
  ) +
  geom_line(data = funnel_curve, aes(x = exposure, y = expected), colour = my_dark_blue, linewidth = 1.1) +
  geom_point(data = brand_funnel, aes(x = exposure, y = claim_frequency), colour = my_blue, size = 3, alpha = 0.85) +
  geom_text(
    data = flagged_brand,
    aes(x = exposure, y = claim_frequency, label = vehicle_brand),
    size = 3.3, vjust = -0.7, check_overlap = TRUE
  ) +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) +
  labs(
    title = "Funnel plot: claim frequency by vehicle brand",
    subtitle = "Blue band is an approximate 95% expected range after considering exposure size",
    x = "Exposure years", y = "Claims per exposure year"
  )
```

::: obs
-   Funnel plot is chosen because segment rankings can be misleading when exposure sizes differ.
-   Points outside or near the funnel boundary deserve investigation before being treated as true high-risk brands.
-   **Implication: fair comparison requires volume adjustment, not just sorting by loss ratio.**
:::

::: notes
This is an advanced visual analytics method aligned with the R4VA funnel plot topic. It directly addresses a common insurance analytics problem: small segments can look extreme simply because they have little exposure. The funnel plot makes the comparison fairer by showing expected variation under different exposure sizes.
:::

## Method Add-on 2 — Distribution shape confirms why simple averages are unstable

```{r}
#| fig-width: 13.2
#| fig-height: 5.1

# For this slide, zero-claim policies are excluded deliberately.
# Reason: the earlier slides already established that zero-claim records dominate the portfolio.
# This view focuses on adverse-experience policies so that distribution shape and segment tails are visible.
positive_lr <- portfolio |>
  filter(!is.na(loss_ratio), is.finite(loss_ratio), loss_ratio > 0) |>
  mutate(
    loss_ratio_display = pmin(loss_ratio, quantile(loss_ratio, 0.95, na.rm = TRUE)),
    policy_type = fct_reorder(policy_type, loss_ratio_display, .fun = median),
    payment_frequency = fct_lump_n(payment_frequency, n = 4)
  )

set.seed(608)
raincloud_dots <- positive_lr |>
  slice_sample(n = min(3000, nrow(positive_lr)))

p_rain <- ggplot(positive_lr, aes(x = policy_type, y = loss_ratio_display)) +
  stat_halfeye(
    adjust = 0.65,
    justification = -0.15,
    .width = 0,
    point_colour = NA,
    fill = my_light_blue
  ) +
  geom_boxplot(width = 0.13, outlier.shape = NA, fill = "white", colour = my_dark, linewidth = 0.35) +
  stat_dots(
    data = raincloud_dots,
    side = "left",
    justification = 1.08,
    binwidth = 0.035,
    dotsize = 0.22,
    colour = my_blue,
    alpha = 0.45
  ) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1), breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Raincloud: positive loss-ratio by product",
    subtitle = "Zero-claim records excluded; loss ratio capped at 95th percentile for readability",
    x = NULL, y = "Positive loss ratio"
  ) +
  theme(plot.title = element_text(size = 18), plot.subtitle = element_text(size = 11))

p_ridge <- positive_lr |>
  ggplot(aes(x = loss_ratio_display, y = payment_frequency, fill = payment_frequency)) +
  ggridges::geom_density_ridges(
    scale = 1.35,
    rel_min_height = 0.02,
    colour = "white",
    alpha = 0.90,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = rep(my_blue, n_distinct(positive_lr$payment_frequency))) +
  scale_x_continuous(labels = percent_format(accuracy = 1), breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Ridgeline: positive loss-ratio by payment frequency",
    subtitle = "Designed to compare distribution shape, not only group averages",
    x = "Positive loss ratio", y = NULL
  ) +
  theme(plot.title = element_text(size = 18), plot.subtitle = element_text(size = 11))

p_rain + p_ridge + plot_layout(widths = c(1.08, 0.92))
```

::: obs
-   The earlier zero-heavy chart explains claim occurrence; this slide isolates **positive-loss policies** to reveal severity and tail structure.
-   Raincloud is used because it combines density, median/IQR, and observation concentration in one view.
-   Ridgeline is used because payment frequency has multiple groups and the question is whether the **whole distribution shape** differs, not only the mean.
-   **Implication: advanced distribution plots are justified because profitability is skewed and tail-driven.**
:::

::: notes
The previous slide using the full loss-ratio distribution looked almost empty because the 99th percentile cap still allowed a very long x-axis and the portfolio is dominated by zero-claim records. This revised slide is more disciplined: zero-claim records are excluded on purpose, and the x-axis is capped at the 95th percentile of positive loss ratios. Raincloud is used for product type because it shows median, spread, and individual concentration. Ridgeline is used for payment frequency because it compares full distribution shape across several groups.
:::

## Statistical Validation — Visual differences are not only visual noise

```{r}
#| fig-width: 12
#| fig-height: 5

kruskal_variables <- c(
  "policy_type", "bonus_score", "payment_frequency", "policy_status",
  "fuel_type", "municipality_type", "circulation_area", "business_type"
)

kruskal_results <- map_dfr(
  kruskal_variables,
  \(v) {
    formula <- as.formula(paste("loss_ratio_w99 ~", v))
    broom::tidy(kruskal.test(formula, data = portfolio))
  },
  .id = "variable_index"
) |>
  mutate(
    variable = kruskal_variables[as.integer(variable_index)],
    neg_log10_p = pmin(-log10(p.value), 50),
    variable = str_replace_all(variable, "_", " ")
  )

kruskal_results |>
  ggplot(aes(x = fct_reorder(variable, neg_log10_p), y = neg_log10_p)) +
  geom_col(fill = my_blue, colour = my_dark, linewidth = 0.35) +
  geom_text(aes(label = number(neg_log10_p, accuracy = 0.1)), hjust = -0.08, size = 4) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.16))) +
  labs(
    title = "Kruskal-Wallis support for segment loss-ratio differences",
    subtitle = "Capped at -log10(p) = 50 for readability; higher values mean stronger evidence",
    x = NULL, y = "-log10(p-value)"
  )
```

::: obs
-   A non-parametric test is used because loss ratio is heavily skewed and zero-inflated.
-   The test supports segment differences, but does **not** prove causality.
-   **Implication: the visuals are strong enough to guide deeper actuarial modelling, not replace it.**
:::

::: notes
The statistical validation checks whether the visual differences are likely to be random noise. Kruskal-Wallis is more defensible than ANOVA because policy-level loss ratio is non-normal. The p-values are capped for readability. The conclusion remains disciplined: statistical evidence supports segmentation, but causality and final pricing require a formal actuarial model.
:::

## Final Recommendations

```{r}
#| fig-width: 13.4
#| fig-height: 5.2

rec_tbl <- tibble(
  Priority = c("1. Product review", "2. Pricing adequacy", "3. Cancellation monitoring", "4. Fair segment comparison"),
  `Management action` = c(
    "Decompose product performance by claim frequency and severity before repricing.",
    "Track premium rate against incurred-loss trend and claim frequency each year.",
    "Investigate whether cancellations occur after adverse claim experience.",
    "Use exposure-adjusted funnel checks before flagging brands or small cells."
  ),
  `Evidence from analysis` = c(
    glue("{as.character(worst_policy$policy_type)} has the highest product loss ratio: {percent(worst_policy$portfolio_loss_ratio, accuracy = 0.1)}"),
    glue("Loss ratio rose by {number(annual_change$loss_ratio_change_pp, accuracy = 0.1)} pp while premium rate changed {percent(annual_change$premium_rate_change, accuracy = 0.1)}"),
    glue("Cancelled policies: {percent(cancelled$portfolio_loss_ratio, accuracy = 0.1)} loss ratio vs active {percent(active$portfolio_loss_ratio, accuracy = 0.1)}"),
    "Funnel plot separates true adverse segments from small-exposure volatility."
  )
) |>
  mutate(Priority = factor(Priority, levels = rev(Priority))) |>
  pivot_longer(
    cols = c(`Management action`, `Evidence from analysis`),
    names_to = "Column",
    values_to = "Text"
  ) |>
  mutate(
    Column = factor(Column, levels = c("Management action", "Evidence from analysis")),
    Text = stringr::str_wrap(Text, width = 44)
  )

ggplot(rec_tbl, aes(x = Column, y = Priority)) +
  geom_tile(fill = my_blue, colour = my_dark, linewidth = 0.45, width = 0.96, height = 0.72) +
  geom_text(aes(label = Text), colour = "white", size = 4.15, lineheight = 0.92, fontface = "bold") +
  scale_x_discrete(position = "top") +
  labs(title = "Management implications from the visual analytics", x = NULL, y = NULL) +
  theme_minimal(base_size = 17) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold", size = 15, colour = my_dark),
    axis.text.y = element_text(face = "bold", size = 13, colour = my_dark),
    plot.title = element_text(face = "bold", size = 24, hjust = 0.5, margin = margin(b = 18)),
    plot.margin = margin(10, 28, 10, 28)
  )
```

::: obs
**Overall conclusion:** profitability pressure is not random. It is concentrated in identifiable product, payment, cancellation, and exposure-adjusted segment patterns. The next step is formal actuarial modelling, but the visual analytics already gives management a defensible shortlist for investigation.
:::

::: notes
This slide turns the visual findings into management action. The table deliberately separates the action from the evidence so the recommendation is not just descriptive. The key discipline is that the visual analytics does not claim to replace actuarial pricing. It narrows the modelling agenda: product type, payment frequency, cancellation status, and exposure-adjusted outliers should be investigated first.
:::
